---
title: "sdmTMB model description"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
bibliography: 'refs.bib'
csl: 'mee.csl'
vignette: >
  %\VignetteIndexEntry{sdmTMB model description}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: sentence
---

```{r setup, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Introduction

This vignette describes the statistical model underlying sdmTMB. sdmTMB fits spatial and spatiotemporal generalized linear mixed models (GLMMs) using Template Model Builder (TMB) [@kristensen2016], with spatial random fields approximated with the SPDE approach [@lindgren2011]. Further details are available in the sdmTMB paper [@sdmTMB], which should be referenced and cited when using sdmTMB in a publication.

# Other vignettes available

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

**If this vignette is being viewed on CRAN, note that many other vignettes describing how to use sdmTMB are available on the [documentation site](https://sdmTMB.github.io/sdmTMB/) under [Articles](https://sdmTMB.github.io/sdmTMB/articles/).**

</div>

\newcommand{\s}{\mathbf{s}}

<!-- Generic commands -->
\newcommand{\bs}[1]{\boldsymbol{#1}}   <!-- bold symbol (Greek) -->
\newcommand{\bv}[1]{\mathbf{#1}}       <!-- bold upright (Latin) -->
\newcommand{\RM}[1]{\mathrm{#1}}       <!-- roman/upright text -->

<!-- Specific commonly used symbols -->
\newcommand{\beps}{\boldsymbol{\epsilon}}
\newcommand{\bgam}{\boldsymbol{\gamma}}
\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\bomega}{\boldsymbol{\omega}}
\newcommand{\bdelta}{\boldsymbol{\delta}}
\newcommand{\bzeta}{\boldsymbol{\zeta}}
\newcommand{\bX}{\mathbf{X}}
\newcommand{\bQ}{\mathbf{Q}}
\newcommand{\bZ}{\mathbf{Z}}
\newcommand{\bH}{\mathbf{H}}
\newcommand{\Qg}{Q^{\mathrm{gg}}}
\newcommand{\bSigma}{\boldsymbol{\Sigma}}
\newcommand{\bA}{\mathbf{A}}

# Notation conventions

This vignette uses the following notation conventions, which generally follow the guidance in @edwards2019:

-   Greek symbols for parameters and random effects, except $\mathbf{b}$ for random intercepts and slopes to match lme4 and glmmTMB,

-   the Latin/Roman alphabet for data except for cases where another symbol is used by convention,

-   bold symbols for vectors or matrices (e.g., $\bomega$ is a vector and $\omega_{\s}$ is the value of $\bomega$ at a point in space $\s$), with Latin/Roman symbols in non-italics (e.g., $\mathbf{X}$, $\mathbf{Q}$),

-   $\phi$ for all distribution dispersion parameters for consistency with the code,

-   $\mathbb{E}[y]$ to define the expected value (mean) of variable $y$,

-   $\mathrm{Var}[y]$ to define the variance of the variable $y$,

-   $(\s, t)$ subscripts as shorthand for observation $i$, which was sampled at location $\s_i$ and time $t_i$ (several observations can share the same location and time),

-   a $^*$ superscript represents values interpolated to observation or prediction locations as opposed to values at mesh vertices (e.g., $\boldsymbol{\omega}$ vs. $\boldsymbol{\omega}^*$), and

-   where possible, notation has been chosen to match VAST [@thorson2019] to maintain consistency (e.g., $\boldsymbol{\omega}$ for spatial fields and $\boldsymbol{\epsilon}_t$ for spatiotemporal fields).

Tables of indices and symbols are summarized in [notation reference tables](#notation-reference-tables) at the end of this vignette.

# sdmTMB model structure

The complete sdmTMB model can be written as

$$
\begin{aligned}
y_{\s,t} &\sim \mathcal{D} \left( \mu_{\s,t}, \phi \right),\\
\mu_{\s,t} &= f^{-1} \left( \eta_{\s,t} \right),\\
\eta_{\s,t} &=
\bX^{\RM{main}}_{\s,t} \bbeta +
O_{\s,t} +
\sum_{k} \mathbf{z}_{k,\s,t}^\top \mathbf{b}_{k,g} +
\bX^{\RM{tvc}}_{\s,t} \bgam_{t} +
\bX^{\RM{svc}}_{\s,t} \bzeta_{\s} +
\omega_{\s} +
\epsilon_{\s,t},
\end{aligned}
$$

where

-   $y_{\s,t}$ represents the response data at point $\s$ and time $t$;
-   $\mathcal{D}$ represents the observation distribution ([family](#observation-model-families)) with mean $\mu_{\s,t}$, dispersion parameter $\phi$, and, for some families, additional parameters (e.g., the Tweedie power parameter);
-   $\mu$ represents the conditional mean on the response/data scale (i.e., $\mathbb{E}[y_{\s,t}] = \mu_{\s,t}$ conditional on the random effects);
-   $f$ represents a link function (e.g., log or logit) and $f^{-1}$ represents its inverse;
-   $\eta$ represents the linear predictor (i.e., in link space, before applying $f^{-1}$);
-   $\bX^{\RM{main}}$, $\bX^{\RM{tvc}}$, and $\bX^{\RM{svc}}$ represent design matrices (the superscript identifiers 'main' = main effects, 'tvc' = time-varying coefficients, and 'svc' = spatially varying coefficients);
-   $\bbeta$ represents a vector of fixed-effect coefficients;
-   $O_{\s,t}$ represents an offset: a covariate with a coefficient fixed at one that enters on the linear predictor scale (e.g., log effort with a log link);
-   $\mathbf{b}_{k,g}$ represents the vector of random intercepts and/or slopes for level $g$ of grouping factor $k$ (with $g$ being the level of factor $k$ for this observation) and $\mathbf{z}_{k,\s,t}$ the corresponding covariate values (a row of the random-effect design matrix $\bZ$);
-   $\bgam_{t}$ represents a vector of time-varying coefficients, where each coefficient $\gamma_{p,t}$ can follow a random walk or AR(1) process;
-   $\bzeta_{\s}$ represents the values of the spatially varying coefficients at location $\s$, where each coefficient field $\bzeta_{l} \sim \RM{MVNormal}(\bs{0},\bSigma_{\zeta,l})$ is a Gaussian random field;
-   $\omega_{\s}$ represents a spatial random field, $\bomega \sim \RM{MVNormal}(\bs{0},\bSigma_\omega)$, with $\bSigma_\omega$ approximating a Matérn covariance via the SPDE approach (described [below](#gaussian-random-fields)); and
-   $\epsilon_{\s,t}$ represents a spatiotemporal random field, $\beps_t \sim \RM{MVNormal}(\bs{0},\bSigma_{\epsilon})$, with Matérn spatial covariance and IID, AR(1), or random walk dynamics through time.

Penalized smoothers are included in $\bX^{\RM{main}}_{\s,t} \bbeta$ for brevity (their penalized coefficients are random effects; see [Smoothers](#smoothers)), and [nonlocal (distributed lag) terms](#nonlocal-formula-covariate-diffusion) are omitted from the overview.
Delta (hurdle) families have two linear predictors, each of which can contain the above components (see [Delta models](#delta-models)).

A single sdmTMB model will rarely, if ever, contain all of the above components.
Next, we will split the model to describe the various parts in more detail using '$\ldots$' to represent the other optional components.

## Main effects

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \bX^{\RM{main}}_{\s,t} \bbeta + \ldots \right)
\end{aligned}
$$

Within `sdmTMB()`, $\bX^{\RM{main}}_{\s,t} \bbeta$ is defined by the `formula` argument and represents the main-effect model matrix and a corresponding vector of coefficients.
This main effect formula can contain optional penalized smoothers or non-linear functions as defined below.

### Smoothers

Smoothers in sdmTMB are implemented with the same formula syntax familiar to mgcv [@wood2017a] users fitting GAMs (generalized additive models).
Smooths are implemented in the formula using `+ s(x)`, which implements a smooth from `mgcv::s()`.
Within these smooths, the same syntax commonly used in `mgcv::s()` can be applied, e.g. 2-dimensional smooths may be constructed with `+ s(x, y)`; smooths can be specific to various factor levels, `+ s(x, by = group)`; smooths can vary according to a continuous variable, `+ s(x, by = x2)`; the basis function dimensions may be specified, e.g. `+ s(x, k = 4)` (see `?mgcv::choose.k`); and various types of splines may be constructed such as cyclic splines to model seasonality, e.g. `+ s(month, bs = "cc", k = 12)`.

While mgcv can fit unpenalized (e.g., B-splines) or penalized splines (P-splines), sdmTMB only implements penalized splines.
The penalized splines are constructed in sdmTMB using the function `mgcv::smooth2random()`, which transforms splines into random effects (and associated design matrices) that are estimable in a mixed-effects modelling framework.
This is the same approach as is implemented in the R packages gamm4 [@wood2020] and brms [@burkner2017].

### Linear break-point threshold models

The linear break-point or "hockey stick" model can be used to describe threshold or asymptotic responses.
This function consists of two pieces, so that for $x < c$, $s(x) = x \cdot \lambda$, and for $x \ge c$, $s(x) = c \cdot \lambda$.
Here, $\lambda$ represents the slope of the function up to the threshold (break point) $c$, and the product $c \cdot \lambda$ represents the value at the asymptote.
No constraints are placed on $\lambda$ or $c$.

These models can be fit by including `+ breakpt(x)` in the model formula, where `x` is a covariate.
The formula can contain a single break-point covariate.

### Logistic threshold models

Models with logistic threshold relationships between a predictor and the response can be fit with the form

$$
s(x)=\tau + \psi\ { \left[ 1+{ e }^{ -\ln \left(19\right) \cdot \left( x-s50 \right)
     / \left(s95 - s50 \right) } \right] }^{-1},
$$

where $s(x)$ denotes the threshold function, $\psi$ is a scaling parameter (controlling the height and sign of the response curve; unconstrained), $\tau$ is an intercept, $s50$ is the value of $x$ at which the function has reached 50% of its total change ($s(x) = \tau + 0.5\psi$), and $s95$ is the value of $x$ at which it has reached 95% of its total change ($s(x) = \tau + 0.95\psi$).
The parameter $s50$ is unconstrained but $s95$ is constrained to be larger than $s50$.

These models can be fit by including `+ logistic(x)` in the model formula, where `x` is a covariate.
The formula can contain a single logistic covariate.

## Spatial random fields

Spatial random fields, $\omega_{\s}$, are included if `spatial = 'on'` (or `TRUE`) and omitted if `spatial = 'off'` (or `FALSE`).

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \ldots + \omega_{\s} + \ldots \right),\\
\bomega &\sim \RM{MVNormal} \left( \bs{0}, \bSigma_\omega \right),
\end{aligned}
$$

where $\bomega$ is the vector of field values at the mesh vertices.
The marginal standard deviation of $\bomega$, $\sigma_\omega$, is indicated by `Spatial SD` in the printed model output or as `sigma_O` in the output of `sdmTMB::tidy(fit, "ran_pars")`.
The 'O' is for 'omega' ($\omega$).

Internally, the random field is a Gaussian Markov random field (GMRF) defined by a sparse precision (inverse covariance) matrix $\bQ_\omega$:

$$
\bomega \sim \RM{MVNormal}\left(\bs{0}, \bQ^{-1}_\omega\right).
$$

$\bQ_\omega$ depends on the mesh and on the parameters $\kappa_\omega$ and $\tau_\omega$, which determine the spatial range and $\sigma_\omega$ (see [Matérn parameterization](#matérn-parameterization)).

## Spatiotemporal random fields

Spatiotemporal random fields are included by default if there are multiple time elements (`time` argument is not `NULL`) and can be set to IID (independent and identically distributed, `'iid'`; default), AR(1) (`'ar1'`), random walk (`'rw'`), or off (`'off'`) via the `spatiotemporal` argument.
These text values are case insensitive.

Spatiotemporal random fields are represented by $\beps_t$ within sdmTMB.
This has been chosen to match the representation in VAST [@thorson2019].
The marginal standard deviation of $\beps_t$ is indicated by `Spatiotemporal SD` in the printed model output or as `sigma_E` in the output of `sdmTMB::tidy(fit, "ran_pars")`.
The 'E' is for 'epsilon' ($\epsilon$).

### IID spatiotemporal random fields

IID spatiotemporal random fields (`spatiotemporal = 'iid'`) can be represented as

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \ldots + \epsilon_{\s,t} + \ldots \right),\\
\beps_{t} &\sim \RM{MVNormal} \left( \bs{0}, \bSigma_{\epsilon} \right).
\end{aligned}
$$

where $\epsilon_{\s,t}$ represents random field deviations at point $\s$ and time $t$.
The random fields are assumed independent across time steps.

Similarly to the spatial random fields, these spatiotemporal random fields (including all versions described below) are parameterized internally with a sparse precision matrix $\bQ_\epsilon$ (with parameters $\kappa_\epsilon$ and $\tau_\epsilon$)

$$
\beps_{t} \sim \RM{MVNormal}\left(\bs{0}, \bQ^{-1}_\epsilon\right).
$$

### AR(1) spatiotemporal random fields

First-order autoregressive, AR(1), spatiotemporal random fields (`spatiotemporal = 'ar1'`) add a parameter defining the correlation between random field deviations from one time step to the next.
They are defined as

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \ldots + \epsilon_{\s,t} + \ldots \right),\\
\beps_{t=1} &\sim \RM{MVNormal} (\bs{0}, \bSigma_{\epsilon}),\\
\beps_{t>1} &= \rho \beps_{t-1} + \sqrt{1 - \rho^2} \bdelta_{t},  \:
\bdelta_{t} \sim \RM{MVNormal} \left(\bs{0}, \bSigma_{\epsilon} \right),
\end{aligned}
$$
where $\rho$ is the correlation between subsequent spatiotemporal random fields and $\bdelta_t$ are IID spatial innovations.
The $\sqrt{1 - \rho^2}$ factor scales the innovations so that every time step has the same (stationary) marginal variance, $\sigma_\epsilon^2$.
The correlation $\rho$ allows for mean-reverting spatiotemporal fields, and is constrained to be $-1 < \rho < 1$.
Internally, the parameter is estimated as `ar1_phi`, which is unconstrained.
The parameter `ar1_phi` is transformed to $\rho$ with $\rho = 2 \left( \mathrm{logit}^{-1}(\texttt{ar1\_phi}) \right) - 1$, mapping `ar1_phi` to the range $(-1, 1)$.

### Random walk spatiotemporal random fields (RW)

Random walk spatiotemporal random fields (`spatiotemporal = 'rw'`) represent a model where the difference in spatiotemporal deviations from one time step to the next are IID.
They are defined as

$$
\begin{aligned}
\mu_{\s,t} &= f^{-1} \left( \ldots + \epsilon_{\s,t} + \ldots \right),\\
\beps_{t=1} &\sim \RM{MVNormal} (\bs{0}, \bSigma_{\epsilon}),\\
\beps_{t>1} &= \beps_{t-1} +  \bdelta_{t},  \:
\bdelta_{t} \sim \RM{MVNormal} \left(\bs{0}, \bSigma_{\epsilon} \right),
\end{aligned}
$$

where the distribution of the spatiotemporal field in the initial time step is the same as for the AR(1) model, but the absence of the $\rho$ parameter allows the spatiotemporal field to be non-stationary in time.
A random walk has no stationary variance: the marginal variance grows with each time step.
Therefore, $\sigma_\epsilon$ is the standard deviation of the innovations $\bdelta_t$ (and of the field in the first time step), not the marginal standard deviation of the field in later time steps.

## Time-varying regression parameters

Parameters can be modelled as time-varying according to a random walk or first-order autoregressive, AR(1), process. The time-series model is defined by `time_varying_type`. For all types:

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots +  \bX^{\RM{tvc}}_{\s,t} \bgam_{t} + \ldots \right),
\end{aligned}
$$
where $\bgam_{t}$ is an optional vector of time-varying regression parameters and $\bX^{\RM{tvc}}_{\s,t}$ is the corresponding model matrix with covariate values.
This is defined via the `time_varying` argument, assuming that the `time` argument is also supplied a column name.
`time_varying` takes a *one-sided* formula.
` ~ 1` implies a time-varying intercept.

The following equations describe the time-series process for a single time-varying coefficient $\gamma_{p,t}$, where $p$ indexes the time-varying coefficient.
When multiple time-varying coefficients are specified, each follows its time-series process independently with its own variance $\sigma_{\gamma,p}^2$ and (for AR1) correlation parameter $\rho_{\gamma,p}$.

For `time_varying_type = 'rw'`, the first time step is estimated as a fixed effect (equivalent to an improper flat prior):

$$
\begin{aligned}
  \gamma_{p,t>1} &\sim \RM{Normal} \left(\gamma_{p,t-1}, \sigma_{\gamma,p}^2 \right).
\end{aligned}
$$

Because the initial value is estimated as part of the time-varying formula, the same variable should not appear in the fixed-effects formula.
The formula `time_varying = ~ 1` implicitly represents a time-varying intercept (assuming the `time` argument has been supplied) and, in this case, the intercept should be omitted from the main-effects formula (e.g., `formula = y ~ 0 + ...` or `formula = y ~ -1 + ...`).

For `time_varying_type = 'rw0'`, the first time step is estimated from a mean-zero prior:

$$
\begin{aligned}
  \gamma_{p,t=1} &\sim \RM{Normal} \left(0, \sigma_{\gamma,p}^2 \right),\\
  \gamma_{p,t>1} &\sim \RM{Normal} \left(\gamma_{p,t-1}, \sigma_{\gamma,p}^2 \right).
\end{aligned}
$$
In this case, the time-varying variable (including the intercept) *should* be included in the main effects.

For `time_varying_type = 'ar1'`:

$$
\begin{aligned}
  \gamma_{p,t=1} &\sim \RM{Normal} \left(0, \sigma_{\gamma,p}^2 \right),\\
  \gamma_{p,t>1} &\sim \RM{Normal} \left(\rho_{\gamma,p}\gamma_{p,t-1}, (1 - \rho_{\gamma,p}^2) \sigma_{\gamma,p}^2 \right),
\end{aligned}
$$
where $\rho_{\gamma,p}$ is the correlation between subsequent time steps.
The first time step is given a mean-zero prior.

## Spatially varying coefficients (SVC)

Spatially varying coefficient models are defined as

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + \bX^{\RM{svc}}_{\s, t} \bzeta_{\s} + \ldots \right),\\
  \bzeta_{l} &\sim \RM{MVNormal} \left( \bs{0}, \bSigma_{\zeta,l} \right),
\end{aligned}
$$

where $\zeta_{l,\s}$ represents the value of the $l$-th spatially varying coefficient field $\bzeta_l$ at location $\s$.
When multiple spatially varying coefficients are specified, each is estimated as an independent spatial random field with its own marginal variance $\sigma_{\zeta,l}^2$.
$\bX^{\RM{svc}}_{\s, t}$ represents the corresponding design matrix for the spatially varying coefficient terms, as defined by a one-sided formula supplied to `spatial_varying`.
For example `spatial_varying = ~ 0 + x`, where `0` omits the intercept.
Because $\zeta_{l,\s}$ is a mean-zero deviation, the main effect of `x` should usually also be included in `formula` so that the coefficient at location $\s$ is $\beta_x + \zeta_{l,\s}$.

The random fields are parameterized internally with a sparse precision matrix $\bQ_{\zeta,l}$:

$$
\bzeta_{l} \sim \RM{MVNormal}\left(\bs{0}, \bQ^{-1}_{\zeta,l}\right).
$$

Each $\bQ_{\zeta,l}$ has its own $\tau_{\zeta,l}$ (and therefore $\sigma_{\zeta,l}$) but shares $\kappa$ with the spatial random field $\bomega$.

## Random intercepts and slopes

Multilevel (hierarchical) intercepts and slopes follow the familiar lme4/glmmTMB formulation.
Each grouping factor $k$ can carry an intercept-only term or a vector of random effects (intercept plus slopes) with a full covariance matrix:

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + \sum_{k} \mathbf{z}_{k,\s,t}^\top \mathbf{b}_{k,g} + \ldots \right),\\
  \mathbf{b}_{k,g} &\sim \RM{MVNormal} \left( \bs{0}, \bSigma_{k} \right),
\end{aligned}
$$

where $g$ is the level of grouping factor $k$ for the observation at $(\s, t)$, $\mathbf{b}_{k,g}$ is the vector of random effects for that level, and $\mathbf{z}_{k,\s,t}$ contains the corresponding covariate values (1 for an intercept and the covariate value for each slope).
Random effects are independent across levels and grouping factors.
The scalar special case `(1 | g)` is a random intercept $b_{g} \sim \RM{Normal}(0, \sigma_{b}^2)$.

Use lme4-style syntax in `formula`, e.g. `(1 | g)` for random intercepts, `(1 + x | g)` for correlated intercepts and slopes, or `(1 | g1) + (1 + x | g2)` to give each grouping factor its own (co)variance.

Internally, standard deviations are estimated on the log scale, and correlations are represented by unconstrained parameters of a Cholesky factor of the correlation matrix (TMB's `UNSTRUCTURED_CORR`), which guarantees a positive-definite $\bSigma_{k}$ for each grouping factor.

## Offset terms

Offset terms can be included through the `offset` argument in `sdmTMB()`.
These are included in the linear predictor as

$$
\begin{aligned}
  \mu_{\s,t} &= f^{-1} \left( \ldots + O_{\s,t} + \ldots \right),
\end{aligned}
$$

where $O_{\s,t}$ is an offset term---a variable on the linear predictor scale with its coefficient fixed at one.
With a log link, the offset is usually a **log-transformed** measure of effort (e.g., `log(area_swept)`), so that $\mu_{\s,t}$ is proportional to effort.

When predicting, `predict.sdmTMB()` without `newdata` uses the offset from the fitted model.
With `newdata`, the offset is 0 unless an `offset` vector is supplied to `predict.sdmTMB()`.
Therefore, if the offset represents log effort, predictions on `newdata` are for one unit of effort (`log(1) = 0`) by default.

## Nonlocal formula (covariate diffusion)

Nonlocal (distributed lag) terms let the effect of a covariate be spread over nearby locations and/or previous time steps, rather than acting only where and when the covariate was measured.
This can be useful when ecological responses are delayed, transported, or accumulated through space and time.
They are added with the `nonlocal_formula` argument to `sdmTMB()` as a one-sided formula using `diffusion()` (space) and/or `time_lag()` (time) wrappers, e.g., `~ diffusion(x) + time_lag(x)`.
`time_lag()` terms require a `time` argument.
See (and cite) @lindmark2025 for the spatial diffusion model and @thorson2026 for the extension to time; those papers give the full formulation and guidance on interpretation.

Each covariate $x_r$ in `nonlocal_formula` is transformed into a smoothed covariate $\tilde{x}_r$, which enters the linear predictor with an estimated coefficient:

$$
\mu_{\s,t} = f^{-1} \left(\ldots + \sum_{r} \beta^{\RM{nl}}_{r}\,\tilde{x}_{r,\s,t} + \ldots \right).
$$

If the same covariate appears in both wrappers, as in `~ diffusion(x) + time_lag(x)`, both apply to one transformed covariate with one coefficient.
If the covariates differ, as in `~ diffusion(x1) + time_lag(x2)`, the model has two transformed covariates and two coefficients.

For `diffusion(x)`, the covariate at time $t$, $\mathbf{x}_t$ (evaluated at the mesh vertices), is smoothed in space by solving

$$
\left(\mathbf{C} + \kappa_{S}^{-2}\mathbf{G}\right)\tilde{\mathbf{x}}_{t} = \mathbf{C}\mathbf{x}_{t},
$$

where $\mathbf{C}$ and $\mathbf{G}$ are the SPDE mass and stiffness matrices (see [Matérn parameterization](#matérn-parameterization)) and $\kappa_S^{-1}$ controls the spatial scale of the diffusion.
The output reports this scale as `RMSDK` (the root-mean-squared displacement of the diffusion kernel, $2/\kappa_S$, in the units of the mesh coordinates).

For `time_lag(x)`, the covariate at each vertex is an exponentially weighted average of current and past values:

$$
\tilde{x}_{\s,t} = (1 - \rho_T)\,x_{\s,t} + \rho_T\,\tilde{x}_{\s,t-1}, \qquad \rho_T = \frac{\kappa_T}{1 + \kappa_T} \in [0, 1),
$$

with $\tilde{x}_{\s,0} = 0$.
Larger $\rho_T$ means a longer memory of past covariate values.

When both wrappers are applied to the same covariate, spatial smoothing and temporal propagation act on the same evolving field (this corresponds to `kappaST = 0` in @thorson2026), and the reported `MSDK` and `RMSDK` account for both.
Separate $\kappa_S$ and $\kappa_T$ parameters are estimated for each covariate.

## Areal autoregressive models (SAR/CAR)

In addition to continuous-space SPDE random fields, sdmTMB can fit areal spatial autoregressive models with `spatial_model = "sar"` or `spatial_model = "car"`.
These require an areal domain supplied to `mesh` with the help of `make_areal_domain()`.

For both areal options, the latent spatial and spatiotemporal effects are Gaussian Markov random fields with precision defined by the areal weight matrix $\mathbf{W}$:

$$
\bomega \sim \RM{MVNormal}\left(\bs{0}, \sigma_\omega^2 \bQ^{-1}\right), \:
\beps_t \sim \RM{MVNormal}\left(\bs{0}, \sigma_\epsilon^2 \bQ^{-1}\right).
$$

Unlike the SPDE case, $\sigma_\omega$ and $\sigma_\epsilon$ here are scale parameters, not marginal standard deviations: the marginal variance of each area is $\sigma^2 (\bQ^{-1})_{ii}$, which depends on $\mathbf{W}$ and the autocorrelation parameter and generally differs among areas.

For SAR (`spatial_model = "sar"`), the precision matrix is

$$
\bQ_{\mathrm{SAR}} = \left(\mathbf{I} - \rho_{\mathrm{SAR}}\mathbf{W}\right)^\top
\left(\mathbf{I} - \rho_{\mathrm{SAR}}\mathbf{W}\right),
$$

where $\rho_{\mathrm{SAR}} \in (-1, 1)$ is reported as `rho_sar`.
By default, SAR uses a row-normalized $\mathbf{W}$ (`sar_weight_style = "row"`), with an option to use raw weights.
With `sar_weight_style = "raw"`, SAR uses the unnormalized adjacency/weight matrix directly, so neighbour influence depends on the original edge weights (and, if unweighted, on the number of neighbours) rather than being scaled to sum to 1 within each row.

For CAR (`spatial_model = "car"`), the precision matrix is

$$
\bQ_{\mathrm{CAR}} = \mathbf{D} - \alpha_{\mathrm{CAR}}\mathbf{W},
$$

where $\mathbf{D}$ is diagonal with elements $D_{ii} = \sum_j W_{ij}$ (the number of neighbours for unweighted adjacency; set to 1 for areas with no neighbours) and $\alpha_{\mathrm{CAR}} \in (0, 1)$ is reported as `alpha_car`.
CAR requires a symmetric adjacency matrix.

# Observation model families

Here we describe the main observation families that are available in sdmTMB and comment on their parametrization, statistical properties, utility, and code representation in sdmTMB.
Families are grouped by outcome type to make it easier to locate an appropriate observation model.

## Bounded or binary outcomes (0–1, proportions)

### Binomial

$$
\operatorname{Binomial} \left(N, \mu \right)
$$ where $N$ is the size or number of trials, and $\mu$ is the probability of success for each trial.
If $N = 1$, the distribution becomes the Bernoulli distribution.
Internally, the distribution is parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaecb5a18095a320b42e2d20c4b120f5f5) in TMB, which is numerically stable when probabilities approach 0 or 1.
Following the structure of `stats::glm()`, lme4, and glmmTMB, a binomial family can be specified in one of 4 ways:

1.  the response may be a factor (and the model classifies the first level versus all others)
2.  the response may be binary (0/1)
3.  the response can be a matrix of form `cbind(success, failure)`, or
4.  the response may be the observed proportions, and the `weights` argument is used to specify the Binomial size ($N$) parameter (`probability ~ ..., weights = N`).

Code defined [within TMB](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaee11f805f02bc1febc6d7bf0487671be).

Example: `family = binomial(link = "logit")`

### Beta-binomial

$$
\operatorname{BetaBinomial} \left(N, \mu, \phi \right)
$$ where $N$ is the number of trials, $\mu$ is the mean success probability, and $\phi$ is a precision parameter that controls overdispersion relative to a Binomial distribution.
The implied Beta parameters are $\alpha = \mu \phi$ and $\beta = (1 - \mu)\phi$, and the variance is

$$
\mathrm{Var}[y] = N \mu (1 - \mu)\, \frac{\phi + N}{\phi + 1},
$$

which exceeds the Binomial variance unless $\phi \to \infty$.
Available links are logit and cloglog.
This family is useful for overdispersed counts of successes/failures (e.g., aggregated Bernoulli data, proportions with extra-binomial variation).

Code defined [within sdmTMB](https://github.com/sdmTMB/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L505-L512).

Example: `family = betabinomial(link = "logit")`

### Beta

$$
\operatorname{Beta} \left(\mu \phi, (1 - \mu) \phi \right)
$$ where $\mu$ is the mean and $\phi$ is a precision parameter.
This parametrization follows @ferrari2004 and the betareg R package [@cribari-neto2010].
The variance is $\mu (1 - \mu) / (\phi + 1)$.

Code defined [within TMB](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga5324c83759d5211c7c2fbbad37fa8e59).

Example: `family = Beta(link = "logit")`

### Ordered beta

The ordered beta distribution [@kubinecOrderedBeta2023] is for continuous proportions on the closed interval $[0, 1]$ that include exact zeros and/or ones.
A single linear predictor $\eta$ and two estimated cutpoints $c_1 < c_2$ define $\Pr(y = 0) = 1 - \mathrm{logit}^{-1}(\eta - c_1)$ and $\Pr(y = 1) = \mathrm{logit}^{-1}(\eta - c_2)$, and values in $(0, 1)$ follow $\operatorname{Beta}(\mu \phi, (1 - \mu)\phi)$ with $\mu = \mathrm{logit}^{-1}(\eta)$.
It is a parsimonious alternative to zero-one-inflated beta models because all three components share one linear predictor.

Example: `family = ordbeta(link = "logit")`

## Count data

### Poisson

$$
\operatorname{Poisson} \left( \mu \right)
$$ where $\mu$ represents the mean and $\mathrm{Var}[y] = \mu$.

Code defined [within TMB](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa1ed15503e1441a381102a8c4c9baaf1).

Example: `family = poisson(link = "log")`

### Censored Poisson

A Poisson distribution in which some observations are censored: the true count is only known to lie in the interval $[y, U]$, where $U$ may be infinite (right censoring).
The likelihood of a censored observation is $\Pr(y \le Y \le U)$ under $\operatorname{Poisson}(\mu)$.
Upper bounds are supplied with `sdmTMBcontrol(censored_upper = ...)`, where `NA` indicates no upper bound and a value equal to $y$ indicates an uncensored observation.
@watson2023 developed this approach to account for hook competition in longline surveys, where observed catch counts are lower bounds on what would have been caught in the absence of competition for baited hooks.

Example: `family = censored_poisson(link = "log")`

### Negative Binomial 2 (NB2)

$$
\operatorname{NB2} \left( \mu, \phi \right)
$$

where $\mu$ is the mean and $\phi$ is the dispersion parameter.
The variance scales quadratically with the mean $\mathrm{Var}[y] = \mu + \mu^2 / \phi$ [@hilbe2011].
The NB2 parametrization is more commonly seen in ecology than the NB1.
Internally, the distribution is parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.

Code defined [within TMB](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga76266c19046e04b651fce93aa0810351).

Example: `family = nbinom2(link = "log")`

### Negative Binomial 1 (NB1)

$$
\operatorname{NB1} \left( \mu, \phi \right)
$$

where $\mu$ is the mean and $\phi$ is the dispersion parameter.
The variance scales linearly with the mean $\mathrm{Var}[y] = \mu + \mu \phi = \mu (1 + \phi)$ [@hilbe2011].
Internally, the distribution is parameterized as the [robust version](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gaa23e3ede4669d941b0b54314ed42a75c) in TMB.

Code defined [within sdmTMB](https://github.com/sdmTMB/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/sdmTMB.cpp#L577-L582) based on NB2 and borrowed from glmmTMB.

Example: `family = nbinom1(link = "log")`

### Truncated negative binomial

Zero-truncated versions of the NB2 and NB1 distributions, for positive counts ($y \ge 1$).
Here, $\mu$ is the mean of the untruncated distribution; the mean of the truncated distribution is $\mu / \Pr(Y > 0)$.
These are mainly used as the positive component of delta models for counts (e.g., `delta_truncated_nbinom2()`).

Example: `family = truncated_nbinom2(link = "log")` or `family = truncated_nbinom1(link = "log")`

### Negative binomial 2 mixture

This is a 2 component mixture that extends the NB2 distribution,
following mixture-distribution approaches for aggregation (e.g., schooling) in survey data [@thorson2011].

$$
(1 - p) \cdot \operatorname{NB2} \left( \mu_1, \phi \right) + p \cdot \operatorname{NB2} \left( \mu_2, \phi \right)
$$

where $\mu_1$ is the mean of the first (smaller component) of the distribution, $\mu_2$ is the mean of the larger component, and $p$ controls the contribution of each component to the mixture.

Example: `family = nbinom2_mix(link = "log")`

## Positive continuous outcomes

### Gamma

$$
\operatorname{Gamma} \left( \phi, \frac{\mu}{\phi}  \right)
$$ where $\phi$ represents the Gamma shape and $\mu / \phi$ represents the scale.
The mean is $\mu$ and variance is $\mu^2 / \phi$.

Code defined [within TMB](https://kaskr.github.io/adcomp/group__R__style__distribution.html#gab0e2205710a698ad6a0ed39e0652c9a3).

Example: `family = Gamma(link = "log")`

### Lognormal

sdmTMB uses the "bias-corrected" lognormal distribution where $\phi$ represents the standard deviation in log-space:

$$
\operatorname{Lognormal} \left( \log \mu - \frac{\phi^2}{2}, \phi^2 \right).
$$ Because of the bias correction, $\mathbb{E}[y] = \mu$ and $\mathrm{Var}[\log y] = \phi^2$.

Code defined [within sdmTMB](https://github.com/sdmTMB/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L47-L54) based on the TMB `dnorm()` normal density.

Example: `family = lognormal(link = "log")`

### Generalized gamma

$$
\operatorname{GenGamma} \left( \mu, \phi, \Qg \right)
$$

sdmTMB implements the @prenticeLogGamma1974 parameterization introduced for spatiotemporal models and index standardization by @dunicGeneralizedGamma2025, with parameters mean $\mu$, scale $\phi$ (standard deviation on the log scale), and a shape parameter $\Qg$ (reported as `Generalized gamma Q`).

Here, $\mu$ is the mean on the data scale, $\phi$ is the log-scale standard deviation, and $\Qg$ controls the shape ($\Qg \to 0$ yields the lognormal; $\Qg = \phi$ yields the gamma).
See @dunicGeneralizedGamma2025 for the full PDF in the Prentice formulation as implemented.
Links available: identity, log, inverse.
This flexibility is useful for right-skewed positive responses with tails heavier or lighter than gamma/lognormal, and is often paired in a hurdle/mixture as `delta_gengamma()` for zero-inflated biomass or catch data (see @dunicGeneralizedGamma2025).

Code defined [within sdmTMB](https://github.com/sdmTMB/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L7-L43).

Example: `family = gengamma(link = "log")`; delta/hurdle: `family = delta_gengamma(link1 = "logit", link2 = "log")`.

### Tweedie

$$
\operatorname{Tweedie} \left(\mu, p, \phi \right), \: 1 < p < 2
$$

where $\mu$ is the mean, $p$ is the power parameter constrained between 1 and 2, and $\phi$ is the dispersion parameter.
The Tweedie distribution (a compound Poisson-gamma distribution for $1 < p < 2$) can be helpful for modelling data that are non-negative and continuous with exact zeros. The variance is $\phi \mu^p$.

Internally, $p = \mathrm{logit}^{-1} (\texttt{thetaf}) + 1$, which constrains it between 1 and 2, and `thetaf` is estimated as an unconstrained parameter.

The [source code](https://kaskr.github.io/adcomp/tweedie_8cpp_source.html) is implemented as in the [cplm](https://CRAN.R-project.org/package=cplm) package [@zhang2013a] and is based on @dunn2005.
The TMB version is defined [here](https://kaskr.github.io/adcomp/group__R__style__distribution.html#ga262f3c2d1cf36f322a62d902a608aae0).

Example: `family = tweedie(link = "log")`

### Gamma mixture

This is a 2 component mixture that extends the Gamma distribution, motivated by mixture-distribution treatments of aggregation in survey data [@thorson2011],

$$
(1 - p) \cdot \operatorname{Gamma} \left( \phi, \frac{\mu_{1}}{\phi}  \right) + p \cdot \operatorname{Gamma} \left( \phi, \frac{\mu_{2}}{\phi}  \right),
$$ where $\phi$ represents the Gamma shape, $\mu_{1} / \phi$ represents the scale for the first (smaller component) of the distribution, $\mu_{2} / \phi$ represents the scale for the second (larger component) of the distribution, and $p$ controls the contribution of each component to the mixture (also interpreted as the probability of larger events).

The mean is $(1-p) \cdot \mu_{1} + p \cdot \mu_{2}$.
The variance follows the usual mixture formula:
$(1-p)\left(\mu_1^2 / \phi + \mu_1^2\right) + p\left(\mu_2^2 / \phi + \mu_2^2\right) - \left[(1-p)\mu_1 + p\mu_2\right]^2$.

Here, and for the other mixture distributions, the probability of the larger mean can be obtained from `plogis(fit$model$par[["logit_p_extreme"]])` and the ratio of the larger mean to the smaller mean can be obtained from `1 + exp(fit$model$par[["log_ratio_mix"]])`.
The standard errors are available in the TMB sdreport: `fit$sd_report`.

If you wish to fix the probability of a large (i.e., extreme) mean, which can be hard to estimate, you can fix this value and pass this to the family:

```r
sdmTMB(...,
  family = gamma_mix(link = "log", p_extreme = 0.01)
)
```

See also `family = delta_gamma_mix()` for an extension incorporating this distribution with delta models.

### Lognormal mixture

This is a 2 component mixture that extends the lognormal distribution, again in the spirit of mixture approaches for aggregating/schooling data [@thorson2011],

$$
(1 - p) \cdot \operatorname{Lognormal} \left( \log \mu_{1} - \frac{\phi^2}{2}, \phi^2 \right) + p \cdot \operatorname{Lognormal} \left( \log \mu_{2} - \frac{\phi^2}{2}, \phi^2 \right).
$$

Because of the bias correction, $\mathbb{E}[y] = (1-p) \cdot \mu_{1} + p \cdot \mu_{2}$.
The log-scale variance of the mixture is not simply $\phi^2$; it can be obtained with the standard mixture-variance formula using component log-means $\log \mu_i - \phi^2/2$ and log-variance $\phi^2$.

As with the Gamma mixture, $p$ controls the contribution of each component to the mixture (also interpreted as the probability of larger events).

Example: `family = lognormal_mix(link = "log")`. See also `family = delta_lognormal_mix()` for an extension incorporating this distribution with delta models. Like with the gamma mixture, fixed probabilities of extreme events ($p$ in notation above) can be passed in, e.g.

```r
sdmTMB(...,
  family = delta_lognormal_mix(p_extreme = 0.01)
)
```

## Continuous real-valued outcomes

### Gaussian

$$
\operatorname{Normal} \left( \mu, \phi^2 \right)
$$ where $\mu$ is the mean and $\phi$ is the standard deviation.
The variance is $\phi^2$.

Example: `family = gaussian(link = "identity")`

Code defined [within TMB](https://kaskr.github.io/adcomp/dnorm_8hpp.html).

### Student-t

$$
\operatorname{Student-t} \left( \mu, \phi, \nu \right)
$$

where $\mu$ is the location (mean for $\nu > 1$), $\phi$ is a scale parameter (not the standard deviation; the variance is $\phi^2 \nu / (\nu - 2)$ for $\nu > 2$), and $\nu$, the degrees of freedom (`df`), is estimated by default and can optionally be fixed by the user.
Lower values of $\nu$ result in heavier tails compared to the Gaussian distribution.
Above approximately `df = 20`, the distribution becomes very similar to the Gaussian.
The Student-t distribution with a low degrees of freedom (e.g., $\nu \le 7$) can be helpful for modelling data that would otherwise be suitable for Gaussian but needs an approach that is robust to outliers [e.g., @anderson2017c].

Code defined [within sdmTMB](https://github.com/sdmTMB/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L37-L45) based on the `dt()` distribution in TMB.

Example: `family = student(link = "identity", df = 7)`

## Zero-inflated and hurdle structures

### Delta models

sdmTMB allows for several different kinds of delta (also known as hurdle) models. These families are implemented by specifying the family as a delta distribution. For example:

```r
sdmTMB(
  ...,
  family = delta_gamma()
)
```

The list of supported families is included in the documentation on [additional families](https://sdmTMB.github.io/sdmTMB/reference/families.html), [delta models](https://sdmTMB.github.io/sdmTMB/articles/delta-models.html), and [Poisson-link delta models](https://sdmTMB.github.io/sdmTMB/articles/poisson-link.html). By default, the `delta_*` families don't use the Poisson link, but this structure can be specified with `delta_gamma(type = "poisson-link")`, which follows the formulation in @thorson2018poisson.

In the "standard" delta model implementation, sdmTMB constructs two internal models (formally, two "linear predictors"), with the first model representing presence-absence and the second model representing the positive component (such as catch rates in fisheries applications).
The positive component can be continuous (e.g., `delta_gamma()`, `delta_lognormal()`, `delta_gengamma()`), a proportion in $(0, 1)$ (`delta_beta()`), or a positive count from a zero-truncated negative binomial (`delta_truncated_nbinom2()`, `delta_truncated_nbinom1()`).
The expected value is the product of the probability of a non-zero observation and the expected value of the positive component.
Default links associated with each family can be inspected with `delta_lognormal()` and equivalent functions. For the standard delta models, the default first linear predictor link is logit. For the Poisson-link type, the first linear predictor link is log.

The `formula`, `spatial`, `spatiotemporal`, and `share_range` arguments of `sdmTMB()` can be specified independently as a 2-element list. For example, the spatial random field might be estimated for only the first linear predictor with:

```r
sdmTMB(
  ...,
  family = delta_gamma(),
  spatial = list("on", "off")
)
```

Or we may want separate main-effects formulas. For example:

```r
sdmTMB(
  formula = list(
    y ~ depth + I(depth^2),
    y ~ 1),
  family = delta_gamma()
)
```

All other arguments are shared between the linear predictors.

Currently if delta models contain smoothers, both components must have the same main-effects formula.

# Gaussian random fields

## Matérn parameterization

The spatial and spatiotemporal random fields in sdmTMB are Gaussian random fields with (approximately) Matérn covariance.
The Matérn covariance between the field values at locations $\s_j$ and $\s_k$, separated by distance $d_{jk} = \lVert \s_j - \s_k \rVert$, is

$$
\mathrm{Cov}\left( \omega_{\s_j}, \omega_{\s_k} \right) = \frac{\sigma^2}{2^{\nu - 1}\Gamma(\nu)} (\kappa d_{jk})^\nu K_\nu \left( \kappa d_{jk} \right),
$$

where $\sigma^2$ is the marginal variance, $\nu$ controls the smoothness of the field, $\Gamma$ is the gamma function, $K_\nu$ is the modified Bessel function of the second kind, and $\kappa$ is a scaling parameter (larger $\kappa$ means correlation decays faster with distance).

Computing with a dense Matérn covariance matrix is slow for many locations.
Instead, sdmTMB uses the stochastic partial differential equation (SPDE) approach of @lindgren2011.
A Gaussian field with Matérn covariance is the solution to the SPDE $(\kappa^2 - \Delta)^{\alpha/2}(\tau\,\omega(\s)) = \mathcal{W}(\s)$, where $\Delta$ is the Laplacian, $\mathcal{W}$ is Gaussian white noise, and $\alpha = \nu + d/2$ for $d$ spatial dimensions.
sdmTMB uses $d = 2$ and $\alpha = 2$, so $\nu = 1$.
Approximating the solution on a triangulated mesh with piecewise-linear basis functions (the finite element method) gives field values at the mesh vertices with a sparse precision matrix

$$
\bQ = \tau^2 \left( \kappa^4 \mathbf{C} + 2 \kappa^2 \mathbf{G} + \mathbf{G} \mathbf{C}^{-1} \mathbf{G} \right),
$$

where $\mathbf{C}$ (a diagonal "mass" matrix) and $\mathbf{G}$ (a sparse "stiffness" matrix) depend only on the mesh.
The sparsity of $\bQ$ is what makes the computations fast.

The parameters $\kappa$ and $\tau$ are reported as the range and marginal standard deviation:

$$
\textrm{range} = \frac{\sqrt{8}}{\kappa}, \qquad
\sigma = \frac{1}{\sqrt{4 \pi}\, \tau \kappa}.
$$

The range is the distance at which the correlation between two points has dropped to approximately 0.13.
Note that for a fixed $\tau$, increasing $\kappa$ decreases $\sigma$.

In the case of a spatiotemporal model with both spatial and spatiotemporal fields, if `share_range = TRUE` in `sdmTMB()` (the default), then a single $\kappa$ and range are estimated with separate $\sigma_\omega$ and $\sigma_\epsilon$.
This often makes sense since data are often only weakly informative about $\kappa$.
If `share_range = FALSE`, then separate $\kappa_\omega$ and $\kappa_\epsilon$ are estimated.
Spatially varying coefficient fields always share $\kappa$ with the spatial random field.

## Projection $\bA$ matrix

In the SPDE approach, the random field at any location is defined by the piecewise-linear basis expansion $\omega(\s) = \sum_k \psi_k(\s)\, \omega_k$, where $\omega_k$ is the field value at mesh vertex $k$ and $\psi_k$ is a "tent" function that is 1 at vertex $k$ and decreases linearly to 0 at the neighbouring vertices.
Evaluating the field at observed or prediction locations is therefore a multiplication by a sparse projection matrix $\bA$ with elements $A_{ik} = \psi_k(\s_i)$ [@lindgren2015]:

$$
\bomega^* = \bA \bomega,
$$ where $\bomega^*$ represents the values of the spatial random fields at the observed locations or predicted data locations.
The matrix $\bA$ has a row for each data point or prediction point and a column for each mesh vertex.
Each row has at most three non-zero elements: the barycentric weights of the vertices of the triangle containing that location.
This interpolation is part of the model definition rather than a separate approximation step.
The same interpolation happens for any spatiotemporal random fields

$$
\beps_t^* = \bA \beps_t.
$$

## Anisotropy

TMB allows for anisotropy, where spatial covariance may depend on direction ([full details](https://kaskr.github.io/adcomp/namespaceR__inla.html)).
Anisotropy can be turned on or off with the logical `anisotropy` argument to `sdmTMB()`.
There are a number of ways to implement anisotropic covariance [@fuglstad2015a], and we adopt geometric anisotropy defined by a 2-parameter matrix $\bH$, which replaces Euclidean distance with $d_{jk} = \sqrt{(\s_j - \s_k)^\top \bH^{-1} (\s_j - \s_k)}$.
The elements of $\bH$ are defined by the parameters $h_1 > 0$ and $h_2$ so that $H_{1,1} = h_{1}$, $H_{1,2} = H_{2,1} = h_{2}$ and $H_{2,2} = (1 + h_{2}^2) / h_{1}$.
$\bH$ is symmetric and positive definite with determinant 1, so it stretches distances along one axis and shrinks them along the perpendicular axis without changing area.
The result is a different range in each direction, with the orientation of the major axis estimated.

Once a model is fitted with `sdmTMB()`, the anisotropy relationships may be plotted using the `plot_anisotropy()` function, which takes the fitted object as an argument.
If a barrier mesh is used, anisotropy is disabled.

## Incorporating physical barriers into the SPDE

In some cases the spatial domain of interest may be complex and bounded by some barrier such as by land or water (e.g., coastlines, islands, lakes).
SPDE models allow for physical barriers to be incorporated into the modelling [@bakka2019].
With `sdmTMB()` models, the mesh construction occurs in two steps: the user (1) constructs a mesh with a call to `sdmTMB::make_mesh()`, and (2) passes the mesh to `sdmTMBextra::add_barrier_mesh()`.
The barriers must be constructed as `sf` objects [@pebesma2018] with polygons defining the barriers.
See `?sdmTMBextra::add_barrier_mesh` for an example.

The barrier implementation requires the user to select a fraction value (`range_fraction` argument) that defines the fraction of the usual spatial range when crossing the barrier [@bakka2019].
For example, if the range was estimated at 10 km, `range_fraction = 0.2` would assign a 2 km range to the triangles marked as barrier.
That makes correlation decay much faster through the barrier than around it, which effectively discourages spatial smoothing from “jumping” across landmasses.
From experimentation, values around 0.1 or 0.2 seem to work well but values much lower than 0.1 can result in convergence issues.

[This website](https://haakonbakkagit.github.io/btopic128.html) by Francesco Serafini and Haakon Bakka provides an illustration with INLA.
The implementation within TMB was borrowed from code written by Olav Nikolai Breivik and Hans Skaug at the [TMB Case Studies](https://github.com/skaug/tmb-case-studies) GitHub site.

# Optimization

## Optimization details

The sdmTMB model is fit by maximum marginal likelihood.
The joint likelihood is written with [RTMB](https://CRAN.R-project.org/package=RTMB) (the default backend; a C++ TMB template can be used instead with `sdmTMBcontrol(backend = "tmb")`), and Template Model Builder (TMB) [@kristensen2016] uses automatic differentiation and the Laplace approximation to integrate over the random effects.
This yields the marginal log likelihood of the fixed effects and its gradient, and the negative marginal log likelihood is minimized with the non-linear optimization routine `stats::nlminb()` in R [@gay1990; @r2021].
For given fixed-effect values, the random effects are set to the values that maximize the joint log likelihood (their conditional modes), around which the Laplace approximation is taken [@kristensen2016].

Like AD Model Builder [@fournier2012ad], sdmTMB fits parameters in phases by default (`multiphase = TRUE` in `sdmTMB::sdmTMBcontrol()`).
Phased estimation is usually faster and more stable, but not always, because it requires building the model an extra time.
In sdmTMB, phased estimation proceeds by first estimating the fixed-effect parameters without the random fields to obtain starting values.
In the second phase, the random-effect parameters (and their variances) are also estimated.
Fixed-effect parameters are also estimated in the second phase and are initialized at their estimates from the first phase.

In some cases, a single call to `stats::nlminb()` may not result in convergence (e.g., the maximum gradient of the marginal likelihood with respect to fixed-effect parameters is not small enough yet), and the algorithm may need to be run multiple times.
In the `sdmTMB::sdmTMBcontrol()` function, we include an argument `nlminb_loops` that will restart the optimization at the previous best values.
The number of `nlminb_loops` should generally be small (e.g., 2 or 3), and defaults to 1.
After `stats::nlminb()`, sdmTMB takes Newton steps to further reduce the gradient: it computes the Hessian $\mathcal{H}$ of the negative marginal log likelihood numerically with `stats::optimHess()` and updates the fixed effects $\boldsymbol{\theta}$ as $\boldsymbol{\theta} - \mathcal{H}^{-1} \mathbf{g}$, where $\mathbf{g}$ is the gradient.
The number of Newton steps is set with `newton_loops` in `sdmTMB::sdmTMBcontrol()` (default 1).
A step is only accepted if it does not increase the negative marginal log likelihood.
If a model is already fit, the function `sdmTMB::run_extra_optimization()` can run additional optimization loops with either routine to further reduce the maximum gradient.

## Assessing convergence

The `sanity()` function runs a set of basic convergence checks on a fitted model (e.g., the Hessian is positive definite, gradients are small, standard errors are not `NA` or very large, and random field variances have not collapsed to zero) and is a good first step.
Much of the guidance around diagnostics and glmmTMB also applies to sdmTMB, e.g. [the glmmTMB vignette on troubleshooting](https://CRAN.R-project.org/package=glmmTMB).
Optimization with `stats::nlminb()` involves specifying the number of iterations and evaluations (`eval.max` and `iter.max`) and the tolerances (`abs.tol`, `rel.tol`, `x.tol`, `xf.tol`)---a greater number of iterations and smaller tolerance thresholds increase the chance that the optimal solution is found, but more evaluations translates into longer computation time.
Warnings of non-positive-definite Hessian matrices (accompanied by parameters with `NA`s for standard errors) often mean models are improperly specified given the data.
Standard errors can be observed in the output of `print.sdmTMB()` or by checking `fit$sd_report`.
The maximum gradient of the marginal likelihood with respect to fixed-effect parameters can be checked by inspecting `fit$gradients`.
Guidance varies, but gradients should be very close to zero (e.g., on the order of $10^{-3}$ or smaller after reasonable parameter scaling) before assuming the fitting routine is consistent with convergence.
If maximum gradients are already relatively small, they can sometimes be reduced further with additional optimization calls beginning at the previous best parameter vector as described above with `sdmTMB::run_extra_optimization()`.

# Notation reference tables

Tables of all major indices (Table 1) and symbols (Table 2) are provided here for quick reference.

```{r indices}
subscripts <- tibble::tribble(
  ~Symbol, ~Description,
  "$\\s$", "Index for space; a vector of x and y coordinates",
  "$t$", "Index for time",
  "$g$", "Level of a random-effect grouping factor",
  "$k$", "Random-effect grouping factor",
  "$p$", "Index for time-varying coefficient",
  "$l$", "Index for spatially varying coefficient (SVC)"
)
knitr::kable(subscripts, caption = "Table 1: Subscript notation", escape = FALSE, booktabs = TRUE)
```

```{r symbols, results='asis'}
symbols <- tibble::tribble(
   ~Symbol, ~Code, ~Description,
  "$y$", "`y_i`", "Observed response data",
  "$\\mu$", "`mu`", "Mean",
  "$\\eta$", "`eta_i`", "Linear predictor before applying the inverse link ($f^{-1}$)",
  "$\\phi$", "`phi`", "A dispersion parameter for a distribution (estimated as `ln_phi`)",
  "$f$", "`fit$family$link`", "Link function",
  "$f^{-1}$", "`fit$family$linkinv`", "Inverse link function",
  "$\\boldsymbol{\\beta}$", "`b_j`", "Fixed-effect coefficient vector",
  "$\\bX$", "`X_ij`", "A predictor model matrix",
  "$\\bZ$", "`Zt_list`", "Random effect design matrix; $\\mathbf{z}_{k,\\s,t}$ is the row for one observation and grouping factor $k$",
  "$O_{\\s, t}$", "`offset_i`", "An offset variable at point $\\s$ and time $t$",

  "$\\omega_{\\s}$", "`omega_s`", "Spatial random field at point $\\s$ (vertex)",
  "$\\omega_{\\s}^*$", "`omega_s_A`", "Spatial random field at point $\\s$ (interpolated)",
  "$\\zeta_{\\s}$", "`zeta_s`", "Spatially varying coefficient random field at point $\\s$ (vertex)",
  "$\\zeta_{\\s}^*$", "`zeta_s_A`", "Spatially varying coefficient random field at point $\\s$ (interpolated)",
  "$\\epsilon_{\\s, t}$", "`epsilon_st`", "Spatiotemporal random field at point $\\s$ and time $t$ (vertex)",
  "$\\epsilon_{\\s, t}^*$", "`epsilon_st_A_vec`", "Spatiotemporal random field at point $\\s$ and time $t$ (interpolated)",
  "$\\delta_{\\s,t}$", "-", "AR(1) or random walk spatiotemporal innovations (vertex)",
  "$\\gamma_{p,t}$", "`b_rw_t`", "Time-varying coefficients",
  "$\\mathbf{b}_{k,g}$", "`re_b_pars`", "Vector of random intercepts/slopes for level $g$ of grouping factor $k$",

  "$\\boldsymbol{\\Sigma}_\\omega$", "-", "Spatial random field covariance matrix",
  "$\\boldsymbol{\\Sigma}_\\zeta$", "-", "Spatially varying coefficient random field covariance matrix",
  "$\\boldsymbol{\\Sigma}_\\epsilon$", "-", "Spatiotemporal random field covariance matrix",
  "$\\bQ_\\omega$", "`Q`", "Spatial random field precision matrix",
  "$\\bQ_\\zeta$", "`Q`", "Spatially varying coefficient random field precision matrix (shared with $\\bQ_\\omega$ up to $\\tau$)",
  "$\\bQ_\\epsilon$", "`Q_st`", "Spatiotemporal random field precision matrix (`Q` if the range is shared)",
  "$\\bSigma_{k}$", "`re_cov_pars`", "Random effect covariance matrix for grouping factor $k$ (log SDs and correlation parameters)",

  "$\\sigma_\\omega$", "`sigma_O`", "Spatial random field marginal standard deviation",
  "$\\sigma_\\epsilon$", "`sigma_E`", "Spatiotemporal random field marginal standard deviation",
  "$\\sigma_\\zeta$", "`sigma_Z`", "Spatially varying coefficient random field marginal standard deviation",
  "$\\sigma_\\gamma$", "`sigma_V`", "Time-varying coefficient innovation standard deviation",
  "$\\tau$", "`ln_tau_O`, `ln_tau_E`, `ln_tau_Z`", "SPDE precision scaling parameters (log scale)",
  "$\\kappa_\\omega$", "`kappa[1, ]`", "Spatial Matérn scaling parameter (estimated as `ln_kappa`)",
  "$\\kappa_\\epsilon$", "`kappa[2, ]`", "Spatiotemporal Matérn scaling parameter",
  "$\\sqrt{8}/\\kappa$", "`range`", "Distance at which correlation drops to approximately 0.13",

  "$\\rho$", "`rho`", "Correlation between spatiotemporal random fields in subsequent time steps (estimated as `ar1_phi`)",
  "$\\rho_{\\gamma}$", "`rho_time`", "Correlation between time-varying coefficients in subsequent time steps",
  "$\\rho_{\\mathrm{SAR}}$", "`rho_sar`", "SAR spatial autocorrelation parameter",
  "$\\alpha_{\\mathrm{CAR}}$", "`alpha_car`", "CAR spatial autocorrelation parameter",
  "$\\lambda$, $c$", "`s_slope`, `s_cut`", "Breakpoint slope and break point",
  "$s50$, $s95$, $\\psi$", "`s50`, `s95`, `s_max`", "Logistic threshold parameters",
  "$\\kappa_S$, $\\kappa_T$", "`kappaS_nl`, `kappaT_nl`", "Nonlocal (distributed lag) spatial and temporal scale parameters",
  "$p$ (Tweedie)", "`tweedie_p`", "Tweedie power parameter (estimated as `thetaf`)",
  "$\\nu$", "`student_df`", "Student-t degrees of freedom",
  "$\\Qg$", "`gengamma_Q`", "Generalized gamma shape parameter ($Q \\rightarrow 0$ gives lognormal; $Q=\\phi$ gives gamma)",
  "$c_1$, $c_2$", "`psi`", "Ordered beta cutpoints",
  "$p$ (mixture)", "`p_extreme`", "Probability of the larger mixture component",
  "$\\bA$", "`A_station`", "Sparse projection matrix to interpolate from mesh vertices to the unique data or prediction locations",
  "$\\bH$", "`H`", "2-parameter matrix defining geometric anisotropy (estimated as `ln_H_input`)"
)
knitr::kable(symbols, caption = "Table 2: Symbol notation, names in the model output or RTMB model code, and descriptions.", escape = FALSE, booktabs = TRUE,
   linesep = c(
    rep('', 9), # data, mean, link, fixed effects
    '\\addlinespace',
    rep('', 8), # random effects and fields
    '\\addlinespace',
    rep('', 6), # covariance and precision
    '\\addlinespace',
    rep('', 7), # standard deviations and range
    '\\addlinespace',
    rep('', 99) # other parameters
  ))
```

# References
