---
title: "Areal models in sdmTMB: SAR, CAR, and SPDE comparison"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Areal models in sdmTMB: SAR, CAR, and SPDE comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**If the code in this vignette has not been evaluated, a rendered version is available on the [documentation site](https://sdmTMB.github.io/sdmTMB/index.html) under 'Articles'.**

```{r setup, include = FALSE, cache=FALSE}
dplyr_installed <- require("dplyr", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
sf_installed <- require("sf", quietly = TRUE)
pkgs <- dplyr_installed && ggplot_installed && sf_installed
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = EVAL,
  purl = EVAL
)
```

```{r packages, message=FALSE, warning=TRUE}
library(dplyr)
library(ggplot2)
library(sdmTMB)
library(sf)
```

This vignette demonstrates three ways to model spatial structure in areal data using the built-in Ohio lung cancer dataset:

- SAR (`spatial_model = "sar"`) on polygons
- CAR (`spatial_model = "car"`) on polygons
- SPDE (`spatial_model = "spde"`) on polygon centroids as a geostatistical point-based version

The areal response is counts (`cases`) with population offset (`log(pop)`), so fitted values are interpreted as expected counts or rates.

```{r data}
dat <- ohio_df
glimpse(dat)
```

## Build an areal domain from polygons

For SAR and CAR models, we pass an areal domain (adjacency + IDs) to `mesh` via `make_areal_domain()`:

```{r domain}
domain <- make_areal_domain(
  data = dat,
  spatial_domain = ohio_sf,
  space_column = "county",
  id_column = "county"
)
```

## Fit SAR and CAR areal models

Both models use the same fixed effects and random-effects structure. The difference is the spatial process (`sar` vs `car`).

```{r fits-areal}
fit_sar <- sdmTMB(
  cases ~ 0 + as.factor(year) + pct_male,
  data = dat,
  mesh = domain,
  spatial_model = "sar",
  time = "year",
  family = poisson(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  offset = log(dat$pop)
)

fit_car <- sdmTMB(
  cases ~ 0 + as.factor(year) + pct_male,
  data = dat,
  mesh = domain,
  spatial_model = "car",
  time = "year",
  family = poisson(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  offset = log(dat$pop)
)
```

```{r sanity-and-print}
sanity(fit_sar)
sanity(fit_car)
fit_sar
fit_car
```

You can inspect random-effect parameters and compare model fit summaries:

```{r model-summaries}
tidy(fit_sar, "ran_pars")
tidy(fit_car, "ran_pars")
AIC(fit_sar, fit_car)
```

## Predict fitted rates for SAR and CAR

We predict fitted counts at observed county-year combinations and convert to rates per 100,000.

```{r predict-areal}
pred_sar <- predict(fit_sar, newdata = dat, type = "response", offset = log(dat$pop))
pred_car <- predict(fit_car, newdata = dat, type = "response", offset = log(dat$pop))

pred_dat <- data.frame(
  county = dat$county,
  year = dat$year,
  pop = dat$pop,
  cases = dat$cases,
  rate_obs = 1e5 * dat$cases / dat$pop,
  rate_fit_sar = 1e5 * pred_sar$est / dat$pop,
  rate_fit_car = 1e5 * pred_car$est / dat$pop,
  omega_s_sar = pred_sar$omega_s,
  omega_s_car = pred_car$omega_s,
  epsilon_st_sar = pred_sar$epsilon_st,
  epsilon_st_car = pred_car$epsilon_st
)

pred_map <- left_join(ohio_sf, pred_dat, by = "county")
```

```{r map-rates-areal}
ggplot(pred_map) +
  geom_sf(aes(fill = rate_fit_sar), colour = "grey40") +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  labs(title = "SAR fitted mortality rate per 100,000")
```

```{r map-rates-car}
ggplot(pred_map) +
  geom_sf(aes(fill = rate_fit_car), colour = "grey40") +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  labs(title = "CAR fitted mortality rate per 100,000")
```

## SPDE approximation using county centroids

The SPDE approach is a continuous-space geostatistical model and does not directly use areal polygons. A common approximation is to collapse each polygon to a centroid and fit a point-based model. This may be reasonable when polygons are relatively small and similar in size and the latent field is smooth within polygons.

```{r spde-prep}
ohio_centroids <- suppressWarnings(st_centroid(ohio_sf))
ohio_xy <- st_coordinates(ohio_centroids)

ohio_spde_dat <- left_join(
  ohio_df,
  data.frame(
    county = ohio_sf$county,
    X = ohio_xy[, "X"] / 1000,
    Y = ohio_xy[, "Y"] / 1000
  ),
  by = "county"
)

ohio_spde_mesh <- make_mesh(ohio_spde_dat, c("X", "Y"), cutoff = 20)
plot(ohio_spde_mesh)
```

```{r spde-fit}
fit_spde <- sdmTMB(
  cases ~ 0 + as.factor(year) + pct_male,
  data = ohio_spde_dat,
  mesh = ohio_spde_mesh,
  spatial_model = "spde",
  family = poisson(link = "log"),
  time = "year",
  spatial = "on",
  spatiotemporal = "iid",
  offset = log(ohio_spde_dat$pop)
)

fit_spde

AIC(fit_spde, fit_car, fit_sar)
```

```{r spde-predict}
pred_spde <- predict(
  fit_spde,
  newdata = ohio_spde_dat,
  type = "response",
  offset = log(ohio_spde_dat$pop)
)

pred_spde_dat <- data.frame(
  county = ohio_spde_dat$county,
  year = ohio_spde_dat$year,
  rate_fit_spde = 1e5 * pred_spde$est / ohio_spde_dat$pop
)

pred_map_spde <- left_join(ohio_sf, pred_spde_dat, by = "county")
```

```{r map-rates-spde}
ggplot(pred_map_spde) +
  geom_sf(aes(fill = rate_fit_spde), colour = "grey40") +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  labs(title = "SPDE (centroid approximation) fitted mortality rate per 100,000")
```

## Compare fitted rates across models

```{r compare-rates}
combined <- pred_map_spde
combined$rate_fit_sar <- pred_map$rate_fit_sar
combined$rate_fit_car <- pred_map$rate_fit_car

ggplot(combined, aes(rate_fit_sar, rate_fit_spde)) +
  geom_point(alpha = 0.6) +
  coord_fixed() +
  geom_abline(intercept = 0, slope = 1) +
  labs(title = "SAR vs SPDE fitted rates")
```

```{r compare-rates-car-spde}
ggplot(combined, aes(rate_fit_car, rate_fit_spde)) +
  geom_point(alpha = 0.6) +
  coord_fixed() +
  geom_abline(intercept = 0, slope = 1) +
  labs(title = "CAR vs SPDE fitted rates")
```

```{r compare-rates-sar-car}
ggplot(combined, aes(rate_fit_sar, rate_fit_car)) +
  geom_point(alpha = 0.6) +
  coord_fixed() +
  geom_abline(intercept = 0, slope = 1) +
  labs(title = "SAR vs CAR fitted rates")
```

For areal outcomes, SAR/CAR are usually the most straightforward models to fit because they encode adjacency among polygons explicitly. The SPDE method, on the other hand, assumes an underlying continuous spatial process. In this case, we find similar support for all 3 models with a slightly higher log likelihood for the geostatistical SPDE model.

