---
title: "Grid areal models in sdmTMB: CAR, SAR, and SPDE comparison"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grid areal models in sdmTMB: CAR, SAR, and SPDE comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**If the code in this vignette has not been evaluated, a rendered version is available on the [documentation site](https://sdmTMB.github.io/sdmTMB/index.html) under 'Articles'.**

```{r setup, include = FALSE, cache=FALSE}
dplyr_installed <- require("dplyr", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
sf_installed <- require("sf", quietly = TRUE)
pkgs <- dplyr_installed && ggplot_installed && sf_installed
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = EVAL,
  purl = EVAL
)
```

```{r packages, message=FALSE, warning=TRUE}
library(dplyr)
library(ggplot2)
library(sdmTMB)
library(sf)
```

This vignette converts point-based dogfish observations into an areal grid and compares:

- CAR model on the areal grid
- SAR model on the areal grid
- SPDE model on point data, aggregated back to the same grid for comparison

The practical goal is to show how model choice changes spatial borrowing across neighbouring cells and whether conclusions are robust to treating space as areal (CAR/SAR) versus continuous (SPDE).

## Build an areal grid and aggregate observations

The original `dogfish` data are tow-level points. To fit CAR/SAR, we first define polygonal areal units (grid cells), assign each tow to a cell, and then aggregate to one observation per `cell_id` and `year`.

```{r grid-build}
dogfish_points <- st_as_sf(dogfish, coords = c("X", "Y"), crs = NA)

dogfish_boundary <- st_union(dogfish_points) |>
  st_convex_hull() |>
  st_as_sf()

dogfish_grid_obj <- make_areal_grid(
  dogfish,
  xy_cols = c("X", "Y"),
  spatial_domain = dogfish_boundary,
  n = c(25L, 20L),
  space_column = "cell_id"
)

plot(dogfish_grid_obj$grid[[1]])
```

Now we'll create a cell-aggregated version of our data. We'll sum the catch weight and area swept by cell and year and take the mean depth.

```{r aggregate}
raw_points <- dogfish_grid_obj$data

cell_data <- raw_points |>
  group_by(cell_id, year) |>
  summarise(
    catch_weight = sum(catch_weight, na.rm = TRUE),
    area_swept = sum(area_swept, na.rm = TRUE),
    depth = mean(depth, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(log_depth = log(depth))
```

## Fit CAR and SAR areal models

We fit the same linear predictor and distribution in both models so the only intended difference is the spatial structure. That makes subsequent CAR-vs-SAR differences easier to attribute to the neighbourhood process rather than to formula changes.

```{r fit-areal}
fit_car <- sdmTMB(
  catch_weight ~ log_depth,
  data = cell_data,
  mesh = dogfish_grid_obj$domain,
  spatial_model = "car",
  time = "year",
  family = tweedie(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  offset = log(cell_data$area_swept)
)

fit_sar <- sdmTMB(
  catch_weight ~ log_depth,
  data = cell_data,
  mesh = dogfish_grid_obj$domain,
  spatial_model = "sar",
  time = "year",
  family = tweedie(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  offset = log(cell_data$area_swept)
)
```

```{r checks-areal}
fit_car
fit_sar
AIC(fit_car, fit_sar)
```

At this stage, check that both models converge and that fixed-effect magnitudes are broadly comparable. Large differences can indicate sensitivity to the areal dependence specification.

## Predict CAR/SAR on observed cells only

Here we predict on the observed `cell_id`-`year` rows (`cell_data`) rather than expanding to all cell-year combinations. This keeps predictions aligned to where data support exists.

```{r pred-areal}
pred_car <- predict(
  fit_car,
  newdata = cell_data,
  type = "response",
  offset = rep(0, nrow(cell_data))
)

pred_sar <- predict(
  fit_sar,
  newdata = cell_data,
  type = "response",
  offset = rep(0, nrow(cell_data))
)

pred_grid <- left_join(
  dogfish_grid_obj$grid,
  cell_data |>
    mutate(
      est_car = pred_car$est,
      est_sar = pred_sar$est,
      omega_s_car = pred_car$omega_s,
      omega_s_sar = pred_sar$omega_s,
      epsilon_st_car = pred_car$epsilon_st,
      epsilon_st_sar = pred_sar$epsilon_st
    )
) |> 
  filter(!is.na(year))
```

```{r map-car}
ggplot(pred_grid) +
  geom_sf(aes(fill = est_car), colour = "grey40") +
  facet_wrap(~year) +
  scale_fill_viridis_c(trans = "log10") +
  labs(fill = "Predicted catch", title = "CAR fitted catch (per unit effort)")
```

```{r map-sar}
ggplot(pred_grid) +
  geom_sf(aes(fill = est_sar), colour = "grey40") +
  facet_wrap(~year) +
  scale_fill_viridis_c(trans = "log10") +
  labs(fill = "Predicted catch", title = "SAR fitted catch (per unit effort)")
```

```{r map-car-omega}
ggplot(pred_grid) +
  geom_sf(aes(fill = omega_s_car), colour = "grey40") +
  scale_fill_gradient2() +
  labs(fill = "Omega", title = "CAR spatial random effect (omega_s)")
```

```{r map-car-epsilon}
ggplot(pred_grid) +
  geom_sf(aes(fill = epsilon_st_car), colour = "grey40") +
  facet_wrap(~year) +
  scale_fill_gradient2() +
  labs(fill = "Epsilon", title = "CAR spatiotemporal random effect (epsilon_st)")
```

## Fit an SPDE model on the original point data

Next we fit a continuous-space SPDE model to the original tow locations. This treats observations as point-referenced rather than area-averaged and therefore answers a slightly different spatial question.

```{r spde-fit}
spde_mesh <- make_mesh(dogfish, c("X", "Y"), cutoff = 10)
plot(spde_mesh)

fit_spde <- sdmTMB(
  catch_weight ~ log(depth),
  data = dogfish,
  mesh = spde_mesh,
  spatial_model = "spde",
  time = "year",
  family = tweedie(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  offset = log(dogfish$area_swept)
)

fit_spde
```

## Fit an SPDE model on the gridded data

We can also fit SPDE directly to the gridded response (`cell_data`) by attaching cell-center coordinates. This lets us compare point-data and gridded-data SPDE fits on the same prediction support.

```{r spde-grid-fit}
grid_centers <- dogfish_grid_obj$grid |>
  st_drop_geometry() |>
  select(cell_id, X, Y)

cell_data_xy <- cell_data |>
  left_join(grid_centers, by = "cell_id")

spde_grid_mesh <- make_mesh(cell_data_xy, c("X", "Y"), cutoff = 10)
plot(spde_grid_mesh)

fit_spde_grid <- sdmTMB(
  catch_weight ~ log_depth,
  data = cell_data_xy,
  mesh = spde_grid_mesh,
  spatial_model = "spde",
  time = "year",
  family = tweedie(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  offset = log(cell_data_xy$area_swept)
)

fit_spde_grid
```

## Project SPDE predictions onto observed grid-cell locations

To align supports directly, we predict both SPDE variants at grid-cell center locations for the same observed `cell_id`-`year` rows used by the CAR/SAR fits.

```{r spde-to-grid}
spde_newdata <- cell_data_xy |>
  select(cell_id, year, depth) |>
  left_join(grid_centers, by = "cell_id")

pred_spde_grid <- predict(
  fit_spde,
  newdata = spde_newdata,
  type = "response",
  offset = rep(0, nrow(spde_newdata))
)

pred_spde_gridfit <- predict(
  fit_spde_grid,
  newdata = cell_data_xy,
  type = "response",
  offset = rep(0, nrow(cell_data_xy))
)

pred_grid_compare <- left_join(
  pred_grid,
  spde_newdata |>
    mutate(est_spde = pred_spde_grid$est) |>
    select(cell_id, year, est_spde) |>
    left_join(
      cell_data_xy |>
        mutate(est_spde_gridfit = pred_spde_gridfit$est) |>
        select(cell_id, year, est_spde_gridfit),
      by = c("cell_id", "year")
    ),
  by = c("cell_id", "year")
)
```

```{r map-spde-grid}
ggplot(pred_grid_compare) +
  geom_sf(aes(fill = est_spde), colour = "grey40") +
  facet_wrap(~year) +
  scale_fill_viridis_c(trans = "log10", na.value = "grey90") +
  labs(fill = "Predicted catch", title = "SPDE fitted catch from point-data fit, predicted at grid centers")
```

```{r map-spde-gridfit}
ggplot(pred_grid_compare) +
  geom_sf(aes(fill = est_spde_gridfit), colour = "grey40") +
  facet_wrap(~year) +
  scale_fill_viridis_c(trans = "log10", na.value = "grey90") +
  labs(fill = "Predicted catch", title = "SPDE fitted catch from gridded-data fit")
```

## Compare fitted values

These scatter plots are a quick agreement diagnostic:

```{r compare-models}
ggplot(st_drop_geometry(pred_grid_compare), aes(est_car, est_sar)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1) +
  coord_fixed() +
  labs(title = "CAR vs SAR fitted values")
```

```{r compare-car-spde}
ggplot(st_drop_geometry(pred_grid_compare), aes(est_car, est_spde)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1) +
  coord_fixed() +
  labs(title = "CAR vs SPDE fitted values")
```

```{r compare-car-spde-gridded}
ggplot(st_drop_geometry(pred_grid_compare), aes(est_car, est_spde_gridfit)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1) +
  coord_fixed() +
  labs(title = "SPDE point-data fit vs SPDE gridded-data fit")
```

For gridded areal responses, CAR/SAR define spatial dependence directly on cell adjacency and are usually the primary formulations.

For continuous data, a geostatistical SPDE model is a more natural approach. However, sometimes with very large datasets, there may be an advantage to aggregating to a grid and using a CAR or SAR model.
