---
title: "Grid areal models in sdmTMB: CAR, SAR, and SPDE comparison"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grid areal models in sdmTMB: CAR, SAR, and SPDE comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**If the code in this vignette has not been evaluated, a rendered version is available on the [documentation site](https://sdmTMB.github.io/sdmTMB/index.html) under 'Articles'.**

```{r setup, include = FALSE, cache=FALSE}
dplyr_installed <- require("dplyr", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
sf_installed <- require("sf", quietly = TRUE)
pkgs <- dplyr_installed && ggplot_installed && sf_installed
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = EVAL,
  purl = EVAL
)
```

```{r packages, message=FALSE, warning=TRUE}
library(dplyr)
library(ggplot2)
library(sdmTMB)
library(sf)
```

This vignette converts point-based dogfish observations into an areal grid and compares:

- CAR model on the areal grid
- SAR model on the areal grid
- SPDE model on point data, summarized back to the same grid for comparison

The practical goal is to show how model choice changes spatial borrowing across neighbouring cells and whether conclusions are robust to treating space as areal (CAR/SAR) versus continuous (SPDE).

## Build an areal grid and assign observations

The original `dogfish` data are tow-level points. To fit CAR/SAR, we first define polygonal areal units (grid cells) and assign each tow to a cell. We keep the tow-level observations for model fitting so that the model sees the original catch, effort, and tow-specific covariates.

```{r grid-build}
dogfish_points <- st_as_sf(dogfish, coords = c("X", "Y"), crs = NA)

dogfish_boundary <- st_union(dogfish_points) |>
  st_convex_hull() |>
  st_as_sf()

dogfish_grid_obj <- make_areal_grid(
  dogfish,
  xy_cols = c("X", "Y"),
  spatial_domain = dogfish_boundary,
  n = c(25L, 20L),
  space_column = "cell_id"
)

plot(dogfish_grid_obj$grid[[1]])
```

For very large datasets, one could aggregate observations to one row per `cell_id` and `year`, for example by summing catch and effort and averaging covariates. That can reduce computation, but it changes the observation support and loses within-cell tow-level variation in covariates, catches, zeros, and effort.

```{r areal-data}
areal_data <- dogfish_grid_obj$data |>
  mutate(log_depth = log(depth))
```

## Fit CAR and SAR areal models

We fit the same linear predictor and distribution in both models so the only intended difference is the spatial structure. That makes subsequent CAR-vs-SAR differences easier to attribute to the neighbourhood process rather than to formula changes.

```{r fit-areal}
fit_car <- sdmTMB(
  catch_weight ~ log_depth,
  data = areal_data,
  mesh = dogfish_grid_obj$domain,
  spatial_model = "car",
  time = "year",
  family = tweedie(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  offset = log(areal_data$area_swept)
)

fit_sar <- sdmTMB(
  catch_weight ~ log_depth,
  data = areal_data,
  mesh = dogfish_grid_obj$domain,
  spatial_model = "sar",
  time = "year",
  family = tweedie(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  offset = log(areal_data$area_swept)
)
```

```{r checks-areal}
fit_car
fit_sar
AIC(fit_car, fit_sar)
```

At this stage, check that both models converge and that fixed-effect magnitudes are broadly comparable. Large differences can indicate sensitivity to the areal dependence specification.

## Predict CAR/SAR on observed tows

Here we predict on the observed tow rows (`areal_data`) rather than expanding to all cell-year combinations. This keeps predictions aligned to where data support exists. We then summarize fitted values to `cell_id`-`year` rows for plotting only.

```{r pred-areal}
pred_car <- predict(
  fit_car,
  newdata = areal_data,
  type = "response",
  offset = rep(0, nrow(areal_data))
)

pred_sar <- predict(
  fit_sar,
  newdata = areal_data,
  type = "response",
  offset = rep(0, nrow(areal_data))
)

pred_areal <- areal_data |>
  mutate(
    est_car = pred_car$est,
    est_sar = pred_sar$est,
    omega_s_car = pred_car$omega_s,
    omega_s_sar = pred_sar$omega_s,
    epsilon_st_car = pred_car$epsilon_st,
    epsilon_st_sar = pred_sar$epsilon_st
  )

pred_cell_year <- pred_areal |>
  group_by(cell_id, year) |>
  summarise(
    est_car = mean(est_car),
    est_sar = mean(est_sar),
    omega_s_car = first(omega_s_car),
    omega_s_sar = first(omega_s_sar),
    epsilon_st_car = first(epsilon_st_car),
    epsilon_st_sar = first(epsilon_st_sar),
    .groups = "drop"
  )

pred_grid <- left_join(
  dogfish_grid_obj$grid,
  pred_cell_year,
  by = "cell_id"
) |> 
  filter(!is.na(year))
```

```{r map-car}
ggplot(pred_grid) +
  geom_sf(aes(fill = est_car), colour = "grey40") +
  facet_wrap(~year) +
  scale_fill_viridis_c(trans = "log10") +
  labs(fill = "Predicted catch", title = "Mean CAR fitted catch (per unit effort)")
```

```{r map-sar}
ggplot(pred_grid) +
  geom_sf(aes(fill = est_sar), colour = "grey40") +
  facet_wrap(~year) +
  scale_fill_viridis_c(trans = "log10") +
  labs(fill = "Predicted catch", title = "Mean SAR fitted catch (per unit effort)")
```

```{r map-car-omega}
ggplot(pred_grid) +
  geom_sf(aes(fill = omega_s_car), colour = "grey40") +
  scale_fill_gradient2() +
  labs(fill = "Omega", title = "CAR spatial random effect (omega_s)")
```

```{r map-car-epsilon}
ggplot(pred_grid) +
  geom_sf(aes(fill = epsilon_st_car), colour = "grey40") +
  facet_wrap(~year) +
  scale_fill_gradient2() +
  labs(fill = "Epsilon", title = "CAR spatiotemporal random effect (epsilon_st)")
```

## Fit an SPDE model on the original point data

Next we fit a continuous-space geostatistical SPDE model to the original tow locations. 

```{r spde-fit}
spde_mesh <- make_mesh(dogfish, c("X", "Y"), cutoff = 10)
plot(spde_mesh)

fit_spde <- sdmTMB(
  catch_weight ~ log(depth),
  data = dogfish,
  mesh = spde_mesh,
  spatial_model = "spde",
  time = "year",
  family = tweedie(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  offset = log(dogfish$area_swept)
)

fit_spde

AIC(fit_spde, fit_car)
```

## Project SPDE predictions onto observed grid-cell locations

To align supports directly, we predict both SPDE variants at grid-cell center locations for the same observed tow rows used by the CAR/SAR fits. We then summarize fitted values to `cell_id`-`year` rows for maps and model-comparison plots.

```{r spde-to-grid}
spde_newdata <- areal_data_xy |>
  mutate(X = X_cell, Y = Y_cell)

pred_spde_grid <- predict(
  fit_spde,
  newdata = spde_newdata,
  type = "response",
  offset = rep(0, nrow(spde_newdata))
)

pred_spde_gridfit <- predict(
  fit_spde_grid,
  newdata = areal_data_xy,
  type = "response",
  offset = rep(0, nrow(areal_data_xy))
)

pred_spde_cell_year <- spde_newdata |>
  mutate(
    est_spde = pred_spde_grid$est,
    est_spde_gridfit = pred_spde_gridfit$est
  ) |>
  group_by(cell_id, year) |>
  summarise(
    est_spde = mean(est_spde),
    est_spde_gridfit = mean(est_spde_gridfit),
    .groups = "drop"
  )

pred_grid_compare <- left_join(
  pred_grid,
  pred_spde_cell_year,
  by = c("cell_id", "year")
)
```

```{r map-spde-grid}
ggplot(pred_grid_compare) +
  geom_sf(aes(fill = est_spde), colour = "grey40") +
  facet_wrap(~year) +
  scale_fill_viridis_c(trans = "log10", na.value = "grey90") +
  labs(fill = "Predicted catch", title = "SPDE fitted catch from point-data fit, predicted at grid centers")
```

## Compare fitted values

These scatter plots are a quick agreement diagnostic:

```{r compare-models}
ggplot(st_drop_geometry(pred_grid_compare), aes(est_car, est_sar)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1) +
  coord_fixed() +
  labs(title = "CAR vs SAR fitted values")
```

```{r compare-car-spde}
ggplot(st_drop_geometry(pred_grid_compare), aes(est_car, est_spde)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1) +
  coord_fixed() +
  labs(title = "CAR vs SPDE fitted values")
```

For gridded areal responses, CAR/SAR define spatial dependence directly on cell adjacency and are usually the primary formulations.

For continuous data, a geostatistical SPDE model is a more natural approach. However, sometimes with very large datasets, there may be a computational advantage to aggregating to a grid and using a CAR or SAR model. That aggregation trades speed for loss of tow-level information.
