---
title: "Grid areal models in sdmTMB: CAR, SAR, and SPDE comparison"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grid areal models in sdmTMB: CAR, SAR, and SPDE comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**If the code in this vignette has not been evaluated, a rendered version is available on the [documentation site](https://sdmTMB.github.io/sdmTMB/index.html) under 'Articles'.**

```{r setup, include = FALSE, cache=FALSE}
dplyr_installed <- require("dplyr", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
sf_installed <- require("sf", quietly = TRUE)
pkgs <- dplyr_installed && ggplot_installed && sf_installed
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = EVAL,
  purl = EVAL
)
```

```{r packages, message=FALSE, warning=TRUE}
library(dplyr)
library(ggplot2)
library(sdmTMB)
library(sf)
```

This vignette illustrates converting point-based observations into an areal grid that can be fit with CAR or SAR models. It compares:

- CAR model on the areal grid
- SAR model on the areal grid
- SPDE model on point data, summarized back to the same grid for comparison

## Build an areal grid and assign observations

The original `dogfish` data are survey tow-level points. To fit a CAR or SAR model, we first define polygon areal units (grid cells) and assign each tow to a cell. We keep the tow-level observations for model fitting so that the model sees the original catch, effort, and tow-specific covariates.

```{r grid-build}
# draw a boundary around our points
dogfish_boundary <- st_as_sf(dogfish, coords = c("X", "Y"), crs = NA) |> 
  st_union() |>
  st_convex_hull()

# pass our data and our spatial boundary to make an areal grid:
dogfish_grid_obj <- make_areal_grid(
  dogfish,
  xy_cols = c("X", "Y"),
  spatial_domain = dogfish_boundary,
  n = c(25L, 20L), # make a grid with 25 cells by 20 cells
  space_column = "cell_id" # optionally rename the output column
)

ggplot(dogfish_grid_obj$grid) + geom_sf() + 
  geom_point(data = dogfish_grid_obj$data, aes(X, Y), alpha = 0.2)
```

For very large datasets, one could aggregate observations to one row per `cell_id` and `year`, for example by summing catch and effort and averaging or summing covariates. That can reduce computation substantial, but it changes the observation support and loses within-cell tow-level variation in covariates, catches, zeros, and effort.
Instead, here we will fit the raw observation data:

```{r areal-data}
areal_data <- dogfish_grid_obj$data |>
  mutate(log_depth = log(depth))

cell_centres <- dogfish_grid_obj$grid |>
  st_drop_geometry() |>
  select(cell_id, X_cell = X, Y_cell = Y)

areal_data_xy <- left_join(areal_data, cell_centres, by = "cell_id")
```

## Fit CAR and SAR areal models

We fit the same linear predictor and distribution in both models so the only intended difference is the spatial structure.
We'll wrap our model fits in `system.time()` just so we can see if these CAR/SAR models are faster than the SPDE (they are).

```{r fit-areal}
system.time(
  fit_car <- sdmTMB(
    catch_weight ~ log_depth,
    data = areal_data,
    mesh = dogfish_grid_obj$domain,
    spatial_model = "car",
    time = "year",
    family = tweedie(link = "log"),
    spatial = "on",
    spatiotemporal = "iid",
    offset = log(areal_data$area_swept)
  )
)
system.time(
  fit_sar <- sdmTMB(
    catch_weight ~ log_depth,
    data = areal_data,
    mesh = dogfish_grid_obj$domain,
    spatial_model = "sar",
    time = "year",
    family = tweedie(link = "log"),
    spatial = "on",
    spatiotemporal = "iid",
    offset = log(areal_data$area_swept)
  )
)
```

```{r checks-areal}
fit_car
fit_sar
AIC(fit_car, fit_sar)
```

At this stage, check that both models converge and that fixed-effect magnitudes are broadly comparable. Large differences can indicate sensitivity to the areal dependence specification.

## Predict CAR/SAR on observed tows

Here we predict on cells that have observed tow rows (`areal_data`) rather than expanding to all cell-year combinations, which we might choose to do in other contexts. Alternatively, we could have created a data frame of one cell per year first and predicted to that.

```{r pred-areal}
pred_car <- predict(
  fit_car,
  newdata = areal_data,
  type = "response",
  offset = rep(0, nrow(areal_data))
)

pred_sar <- predict(
  fit_sar,
  newdata = areal_data,
  type = "response",
  offset = rep(0, nrow(areal_data))
)

pred_areal <- areal_data |>
  mutate(
    est_car = pred_car$est,
    est_sar = pred_sar$est,
    omega_s_car = pred_car$omega_s,
    omega_s_sar = pred_sar$omega_s,
    epsilon_st_car = pred_car$epsilon_st,
    epsilon_st_sar = pred_sar$epsilon_st
  )

# condense to one prediction per cell for plotting
pred_cell_year <- pred_areal |>
  group_by(cell_id, year) |>
  summarise(
    est_car = mean(est_car),
    est_sar = mean(est_sar),
    omega_s_car = first(omega_s_car),
    omega_s_sar = first(omega_s_sar),
    epsilon_st_car = first(epsilon_st_car),
    epsilon_st_sar = first(epsilon_st_sar),
    .groups = "drop"
  )

pred_grid <- left_join(
  dogfish_grid_obj$grid,
  pred_cell_year,
  by = "cell_id"
) |> 
  filter(!is.na(year))
```

```{r map-car}
ggplot(pred_grid) +
  geom_sf(aes(fill = est_car), colour = "grey40") +
  facet_wrap(~year) +
  scale_fill_viridis_c(trans = "log10") +
  labs(fill = "Predicted catch", title = "Mean CAR fitted catch (per unit effort)")
```

```{r map-sar}
ggplot(pred_grid) +
  geom_sf(aes(fill = est_sar), colour = "grey40") +
  facet_wrap(~year) +
  scale_fill_viridis_c(trans = "log10") +
  labs(fill = "Predicted catch", title = "Mean SAR fitted catch (per unit effort)")
```

```{r map-car-omega}
ggplot(pred_grid) +
  geom_sf(aes(fill = omega_s_car), colour = "grey40") +
  scale_fill_gradient2() +
  labs(fill = "Omega", title = "CAR spatial random effect (omega_s)")
```

```{r map-car-epsilon}
ggplot(pred_grid) +
  geom_sf(aes(fill = epsilon_st_car), colour = "grey40") +
  facet_wrap(~year) +
  scale_fill_gradient2() +
  labs(fill = "Epsilon", title = "CAR spatiotemporal random effect (epsilon_st)")
```

## Fit an SPDE model on the original point data

Next we fit a continuous-space geostatistical SPDE model to the original tow locations. 

```{r spde-fit}
spde_mesh <- make_mesh(dogfish, c("X", "Y"), cutoff = 8.5)
spde_mesh$mesh$n
dogfish_grid_obj$domain$n_s
plot(spde_mesh)

system.time(
fit_spde <- sdmTMB(
  catch_weight ~ log(depth),
  data = dogfish,
  mesh = spde_mesh,
  spatial_model = "spde",
  time = "year",
  family = tweedie(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  offset = log(dogfish$area_swept)
)
)
fit_spde

AIC(fit_spde, fit_sar)
```

That took approximately twice as long as the CAR/SAR models.

## Project SPDE predictions onto observed grid-cell locations

To align our predictions, we predict the SPDE model at grid-cell center locations for the same observed tow rows used by the CAR/SAR fits. We then summarize fitted values to `cell_id`-`year` rows for maps and model-comparison plots.

```{r spde-to-grid}
spde_newdata <- areal_data_xy |>
  mutate(X = X_cell, Y = Y_cell)

pred_spde_grid <- predict(
  fit_spde,
  newdata = spde_newdata,
  type = "response",
  offset = rep(0, nrow(spde_newdata))
)

pred_spde_cell_year <- spde_newdata |>
  mutate(
    est_spde = pred_spde_grid$est
  ) |>
  group_by(cell_id, year) |>
  summarise(
    est_spde = mean(est_spde),
    .groups = "drop"
  )

pred_grid_compare <- left_join(
  pred_grid,
  pred_spde_cell_year,
  by = c("cell_id", "year")
)
```

```{r map-spde-grid}
ggplot(pred_grid_compare) +
  geom_sf(aes(fill = est_spde), colour = "grey40") +
  facet_wrap(~year) +
  scale_fill_viridis_c(trans = "log10", na.value = "grey90") +
  labs(fill = "Predicted catch", title = "SPDE fitted catch from point-data fit, predicted at grid centers")
```

## Compare fitted values

These scatter plots are a quick check of how well the predictions agree:

```{r compare-models}
ggplot(st_drop_geometry(pred_grid_compare), aes(est_car, est_sar)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1) +
  coord_fixed() +
  labs(title = "CAR vs SAR fitted values")
```

```{r compare-car-spde}
ggplot(st_drop_geometry(pred_grid_compare), aes(est_car, est_spde)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1) +
  coord_fixed() +
  labs(title = "CAR vs SPDE fitted values")
```

For continuous data, a geostatistical SPDE model is a more natural approach and it better fits the data here. However, sometimes with very large datasets, there may be a computational advantage to using the CAR/SAR models and an even bigger computational advantage to aggregating to a grid and using a CAR or SAR model. That aggregation trades speed for loss of tow-level information. Even without aggregation, the CAR model followed by the SAR model were faster than the SPDE model (approximately twice as fast in this case).
