---
title: "Estimating covariate diffusion with sdmTMB"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: '../refs.bib'
csl: '../mee.csl'
vignette: >
  %\VignetteIndexEntry{Estimating covariate diffusion with sdmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**If the code in this vignette has not been evaluated, a rendered version is available on the [documentation site](https://sdmTMB.github.io/sdmTMB/index.html) under 'Articles'.**

```{r setup, include = FALSE, cache=FALSE}
ggplot_installed <- require("ggplot2", quietly = TRUE)
pkgs <- ggplot_installed
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = EVAL,
  purl = EVAL
)
```

```{r packages, message=FALSE, warning=TRUE}
library(ggplot2)
library(sdmTMB)
theme_set(theme_light())
```

This vignette demonstrates covariate-diffusion models using simulated data. These models could alternatively be described as "distributed lag" models.

When might you want to fit these models? Sometimes, the spatial or temporal scale at which covariates influence the response is unclear.

For example, fish move. Therefore, if we measure temperature at the location where a fish was sampled, it may be unclear whether the model should use the temperature at the sampling site or a smoothed version of the temperature surface that accounts for fish movement and integration over a broader spatial range.

One option is to fit models using covariates with different levels of smoothing and compare them. Alternatively, we can estimate the appropriate degree of smoothing simultaneously with the parameters of the main model. This is what the covariate-diffusion functionality does.

Users are encouraged to read and cite @lindmark2025 for spatial models and @thorson2026 for extensions to temporal and spatiotemporal settings, including the introduction of the MSD and RMSD terms.

We begin with a simple spatial-diffusion example and then fit an example that combines `space()` and `time()` terms.

## Simple spatial diffusion example

We will start by simulating data with a fine-scale predictor (`x1`) and a spatially diffused effect:

```{r simulate-space}
set.seed(1)
n_sites <- 220
site <- data.frame(X = runif(n_sites), Y = runif(n_sites))
dat <- transform(
  site,
  x1 = as.numeric(scale(
    sin(6 * pi * X) + cos(6 * pi * Y) + rnorm(n_sites, sd = 1)
  ))
)
mesh <- make_mesh(dat, c("X", "Y"), cutoff = 0.05)

sim_space <- simulate_new(
  formula = ~ 1,
  data = dat,
  mesh = mesh,
  family = gaussian(),
  spatial = "off",
  spatiotemporal = "off",
  range = 0.2,
  sigma_O = 0,
  phi = 0.1,
  B = c(0, 0.8),
  covariate_diffusion = ~ space(x1),
  diffusion_kappaS = 1.5,
  seed = 1
)

dat$observed <- sim_space$observed
dat$x1_truth <- sim_space$diffusion_truth_space_x1
head(dat)
```

Fit the matching covariate-diffusion model:

```{r fit-space}
fit <- sdmTMB(
  observed ~ 1,
  mesh = mesh,
  covariate_diffusion = ~ space(x1), #<
  spatial = "off",
  data = dat
)

fit
tidy(fit)
tidy(fit, effects = "ran_pars")
```

Here `kappaS_cov_diff[x1]` is the estimated spatial diffusion scale parameter for `x1` (larger values imply less smoothing).

The output also reports `MSD[x1]` and `RMSD[x1]`, the implied mean squared displacement (`4 / kappaS_cov_diff[x1]^2`) and root mean squared displacement (`2 / kappaS_cov_diff[x1]`).
`RMSD[x1]` is a typical diffusion distance (a blur radius) in coordinate units.
Users applying this spatial diffusion method should cite @lindmark2025.
Users using these MSD and RMSD metrics should consult and cite @thorson2026.

Compare true and estimated diffused covariates:

```{r compare-space}
pred <- predict(fit)
pred$X <- dat$X
pred$Y <- dat$Y

cor(pred$diffusion_cov_space_x1, pred$x1_truth)

g1 <- ggplot(pred, aes(X, Y, colour = x1)) +
  geom_point(size = 0.7) +
  scale_colour_gradient2() +
  coord_equal() +
  ggtitle("Fine-scale observed predictor (x1)")

g2 <- ggplot(pred, aes(X, Y, colour = x1_truth)) +
  geom_point(size = 0.7) +
  scale_colour_gradient2() +
  coord_equal() +
  ggtitle("True diffused covariate (from simulation)")

g3 <- ggplot(pred, aes(X, Y, colour = diffusion_cov_space_x1)) +
  geom_point(size = 0.7) +
  scale_colour_gradient2() +
  coord_equal() +
  ggtitle("Estimated diffused covariate")

print(g1)
print(g2)
print(g3)
```

We can visualize the diffusion operator and the original and smoothed covariate on the mesh:

```{r}
plot_diffusion_kernel(fit, component = "space")
plot_diffused_covariate(fit, component = "space")
```

## Space-time example with additive terms

Next we will simulate and fit a model containing spatial and temporal components: `~ space(x1) + time(x1)`.

```{r fit-space-time-all}
set.seed(1)
n_t <- 16
n_sites <- 300
site_st <- data.frame(X = runif(n_sites), Y = runif(n_sites))
dat_st <- data.frame(
  X = rep(site_st$X, times = n_t),
  Y = rep(site_st$Y, times = n_t),
  year = rep(seq_len(n_t), each = n_sites)
)
dat_st$x1 <- as.numeric(scale(
  sin(2 * pi * (dat_st$X + dat_st$year / 8)) +
    cos(2 * pi * (dat_st$Y - dat_st$year / 10)) +
    0.6 * sin(4 * pi * dat_st$X) * cos(dat_st$year / 3) +
    rnorm(nrow(dat_st), sd = 0.15)
))
mesh_st <- make_mesh(dat_st, c("X", "Y"), cutoff = 0.07)

sim_st <- simulate_new(
  formula = ~ 1,
  data = dat_st,
  mesh = mesh_st,
  time = "year",
  family = gaussian(),
  spatial = "on",
  spatiotemporal = "off",
  range = 0.3,
  sigma_O = 0.2,
  phi = 0.1,
  B = c(0, 0.7, 0.6),
  covariate_diffusion = ~ space(x1) + time(x1),
  diffusion_kappaS = 4.4,
  diffusion_rhoT = 0.3,
  seed = 123
)

dat_st$observed <- sim_st$observed
```

```{r fit-st}
fit_st <- sdmTMB(
  observed ~ 1,
  mesh = mesh_st,
  time = "year",
  covariate_diffusion = ~ space(x1) + time(x1), #<
  spatial = "on",
  spatiotemporal = "off",
  data = dat_st
)

sanity(fit_st)
fit_st

tidy(fit_st, effects = "ran_pars")
```

In the random-effect parameters, `kappaS_cov_diff[x1]` controls spatial smoothing and `kappaT_cov_diff[x1]` controls temporal diffusion.
Users applying the time diffusion extension should read and cite @thorson2026.

## Supplying a separate covariate grid

By default, the covariate field used for diffusion is built from `data` alone.
If the covariate is available on a finer or more complete grid than the
observation locations, we can supply it via `covariate_diffusion_data`. This
also lets us define the covariate at a forecast (`extra_time`) slice without
padding `data` with extra rows.

```{r fit-space-time-grid}
extra_yr <- n_t + 1
grid_st <- expand.grid(
  X = seq(0, 1, length.out = 25),
  Y = seq(0, 1, length.out = 25),
  year = seq_len(extra_yr)
)
grid_st$x1 <- as.numeric(scale(
  sin(2 * pi * (grid_st$X + grid_st$year / 8)) +
    cos(2 * pi * (grid_st$Y - grid_st$year / 10)) +
    0.6 * sin(4 * pi * grid_st$X) * cos(grid_st$year / 3)
))

fit_st_grid <- sdmTMB(
  observed ~ 1,
  mesh = mesh_st,
  time = "year",
  covariate_diffusion = ~ space(x1) + time(x1), #<
  covariate_diffusion_data = grid_st, #<
  extra_time = extra_yr, #<
  spatial = "on",
  spatiotemporal = "off",
  data = dat_st
)
sanity(fit_st_grid)
```

Because the diffused field now comes entirely from `grid_st`, `predict()` can
reuse it for `newdata` that only contains coordinates, time, any regular covariate data, but no diffused covariate data.

```{r predict-space-time-grid}
nd_st <- replicate_df(site_st, "year", seq_len(extra_yr))
pred_grid <- predict(fit_st_grid, newdata = nd_st)

ggplot(pred_grid, aes(X, Y, colour = diffusion_cov_space_x1)) +
  geom_point() +
  facet_wrap(~year) +
  scale_colour_gradient2()
```

## Plot diffusion kernels

We can separately visualize the spatial diffusion, time diffusion, or their combined effects as either the diffusion kernel or the smoothed covariate field represented on the mesh:

```{r plot-diffusion}
plot_diffusion_kernel(fit_st, component = "space")
plot_diffusion_kernel(fit_st, component = "time", common_scale = TRUE)
plot_diffusion_kernel(fit_st, component = "combined")
plot_diffusion_kernel(fit_st, component = "combined", common_scale = TRUE)
```

```{r}
plot_diffused_covariate(fit_st, component = "space")
plot_diffused_covariate(fit_st, component = "time")
plot_diffused_covariate(fit_st, component = "combined")
```

We could also have plotted the spatially diffused covariate at the original locations from the predictions.

```{r}
pred <- predict(fit_st)

ggplot(pred, aes(X, Y, colour = x1)) + geom_point() +
  scale_colour_gradient2()

ggplot(pred, aes(X, Y, colour = diffusion_cov_space_x1)) + geom_point() +
  scale_colour_gradient2()
```

# References
