---
title: "Estimating covariate diffusion with sdmTMB"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: '../refs.bib'
csl: '../mee.csl'
vignette: >
  %\VignetteIndexEntry{Estimating covariate diffusion with sdmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**If the code in this vignette has not been evaluated, a rendered version is available on the [documentation site](https://sdmTMB.github.io/sdmTMB/index.html) under 'Articles'.**

```{r setup, include = FALSE, cache=FALSE}
ggplot_installed <- require("ggplot2", quietly = TRUE)
pkgs <- ggplot_installed
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = EVAL,
  purl = EVAL
)
```

```{r packages, message=FALSE, warning=TRUE}
library(ggplot2)
library(sdmTMB)
theme_set(theme_light())
```

This vignette demonstrates covariate-diffusion models with simulated data.
We start with a simple spatial diffusion example and then fit a an example that combines `space()` and `time()` terms.

## Simple spatial diffusion example

Simulate data with a fine-scale predictor (`x1`) and a spatially diffused effect:

```{r simulate-space}
set.seed(1)
n_sites <- 220
site <- data.frame(X = runif(n_sites), Y = runif(n_sites))
dat <- transform(
  site,
  x1 = as.numeric(scale(
    sin(6 * pi * X) + cos(6 * pi * Y) + rnorm(n_sites, sd = 1)
  ))
)
mesh <- make_mesh(dat, c("X", "Y"), cutoff = 0.05)

sim_space <- simulate_new(
  formula = ~ 1,
  data = dat,
  mesh = mesh,
  family = gaussian(),
  spatial = "off",
  spatiotemporal = "off",
  range = 0.2,
  sigma_O = 0,
  phi = 0.1,
  B = c(0, 0.8),
  covariate_diffusion = ~ space(x1),
  diffusion_kappaS = 1.5,
  seed = 1
)

dat$observed <- sim_space$observed
dat$x1_truth <- sim_space$diffusion_truth_space_x1
head(dat)
```

Fit the matching covariate-diffusion model:

```{r fit-space}
fit <- sdmTMB(
  observed ~ 1,
  mesh = mesh,
  covariate_diffusion = ~ space(x1),
  spatial = "off",
  data = dat
)

fit
tidy(fit)
tidy(fit, effects = "ran_pars")
```

Here `kappaS_dl[x1]` is the estimated spatial diffusion scale parameter for `x1` (larger values imply less smoothing).

The output also reports `MSD[x1]` and `RMSD[x1]`, the implied mean squared displacement (`4 / kappaS_dl[x1]^2`) and root mean squared displacement (`2 / kappaS_dl[x1]`).
`RMSD[x1]` is a typical diffusion distance (a blur radius) in coordinate units.
Users applying this spatial diffusion method should cite @lindmark2025.
Users using these MSD and RMSD metrics should consult and cite @thorson2026.

Compare true and estimated diffused covariates:

```{r compare-space}
pred <- predict(fit)
pred$X <- dat$X
pred$Y <- dat$Y

cor(pred$diffusion_cov_space_x1, pred$x1_truth)

g1 <- ggplot(pred, aes(X, Y, colour = x1)) +
  geom_point(size = 0.7) +
  scale_colour_gradient2() +
  coord_equal() +
  ggtitle("Fine-scale observed predictor (x1)")

g2 <- ggplot(pred, aes(X, Y, colour = x1_truth)) +
  geom_point(size = 0.7) +
  scale_colour_gradient2() +
  coord_equal() +
  ggtitle("True diffused covariate (from simulation)")

g3 <- ggplot(pred, aes(X, Y, colour = diffusion_cov_space_x1)) +
  geom_point(size = 0.7) +
  scale_colour_gradient2() +
  coord_equal() +
  ggtitle("Estimated lagged covariate")

print(g1)
print(g2)
print(g3)
```

We can visualize the diffusion operator:

```{r}
plot_diffusion_kernel(fit, component = "space")
plot_diffused_covariate(fit, component = "space")
```

Users applying spatial diffusion should read and cite @lindmark2025.

## Space-time example with additive terms

Now simulate and fit a model containing additive spatial and temporal components: `~ space(x1) + time(x1)`.

```{r fit-space-time-all}
set.seed(1)
n_t <- 16
n_sites <- 300
site_st <- data.frame(X = runif(n_sites), Y = runif(n_sites))
dat_st <- data.frame(
  X = rep(site_st$X, times = n_t),
  Y = rep(site_st$Y, times = n_t),
  year = rep(seq_len(n_t), each = n_sites)
)
dat_st$x1 <- as.numeric(scale(
  sin(2 * pi * (dat_st$X + dat_st$year / 8)) +
    cos(2 * pi * (dat_st$Y - dat_st$year / 10)) +
    0.6 * sin(4 * pi * dat_st$X) * cos(dat_st$year / 3) +
    rnorm(nrow(dat_st), sd = 0.15)
))
mesh_st <- make_mesh(dat_st, c("X", "Y"), cutoff = 0.07)

sim_st <- simulate_new(
  formula = ~ 1,
  data = dat_st,
  mesh = mesh_st,
  time = "year",
  family = gaussian(),
  spatial = "on",
  spatiotemporal = "off",
  range = 0.3,
  sigma_O = 0.2,
  phi = 0.1,
  B = c(0, 0.7, 0.6),
  covariate_diffusion = ~ space(x1) + time(x1),
  diffusion_kappaS = 4.4,
  diffusion_rhoT = 0.3,
  seed = 123
)

dat_st$observed <- sim_st$observed
```

```{r fit-st}
fit_st <- sdmTMB(
  observed ~ 1,
  mesh = mesh_st,
  time = "year",
  covariate_diffusion = ~ space(x1) + time(x1),
  spatial = "on",
  spatiotemporal = "off",
  data = dat_st
)

tidy(fit_st, effects = "ran_pars")
```

In the random-effect parameters, `kappaS_dl[x1]` controls spatial smoothing and `kappaT_dl[x1]` controls temporal carryover.
Users applying the time diffusion extension should read and cite @thorson2026.

## Plot diffusion kernels

We can separately visualize the spatial diffusion, time diffusion, or their combined effects:

```{r plot-diffusion}
plot_diffusion_kernel(fit_st, component = "space")
plot_diffusion_kernel(fit_st, component = "time")
plot_diffusion_kernel(fit_st, component = "combined")
```

```{r}
plot_diffused_covariate(fit_st, component = "space")
plot_diffused_covariate(fit_st, component = "time")
plot_diffused_covariate(fit_st, component = "combined")
```


# References
