---
title: "Custom priors and penalties"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom priors and penalties}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**If the code in this vignette has not been evaluated, a rendered version is available on the [documentation site](https://sdmTMB.github.io/sdmTMB/index.html) under 'Articles'.**

```{r setup, include = FALSE, cache=FALSE}
ggplot_installed <- require("ggplot2", quietly = TRUE)
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && ggplot_installed
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = EVAL,
  purl = EVAL
)
```

```{r packages, message=FALSE, warning=FALSE}
library(sdmTMB)
library(ggplot2)
```

`sdmTMBpriors()` has built-in priors for common cases, such as normal priors on the main effects or PC priors on Matérn random fields.
With the RTMB backend (the default), you can also write your own prior or penalty as an R function.
The function can refer to any parameter, transformed parameters, or even combine several of them.

The function is passed to `sdmTMBpriors(custom = ...)`.
It's called as `custom(par, theta)`, where:

- `par` is the list of raw (internal) parameters, and
- `theta` is a list of natural-scale versions of some of them, such as `phi = exp(ln_phi)` or the Matérn `range`.

It returns one or more log densities, which are added to the log likelihood.
Because the function is evaluated inside the model's objective, the priors are included in the gradients, the Laplace approximation, and the standard errors.

A few rules apply because the function is automatically differentiated:

- Use densities from RTMB, such as `RTMB::dnorm(x, mean, sd, log = TRUE)`. `stats::dnorm()` won't work.
- Return *log* densities. Remember `log = TRUE`; sdmTMB can't check this for you.
- Don't use `if` statements on parameter values.

You can find common density functions available in RTMB at <https://kaskr.r-universe.dev/articles/RTMB/RTMB-tips.html>, and the full list in the RTMB help page for `dnorm`:

```{r rtmb-help, eval=FALSE}
help("dnorm,ad,ad.,ad.,logical.-method", package = "RTMB")
```

## A first example: a prior on a coefficient

We'll fit a model to the Pacific cod survey data with a quadratic effect of depth and spatial and spatiotemporal random fields.

```{r fit}
mesh <- make_mesh(pcod, c("X", "Y"), cutoff = 15)
fit <- sdmTMB(
  density ~ depth_scaled + depth_scaled2,
  data = pcod,
  mesh = mesh,
  family = tweedie(),
  time = "year",
  spatiotemporal = "iid"
)
```

We can find what parameters and derived parameters are available for us to place penalties/priors on them:

```{r find}
get_prior_parameters(fit) |> head()
```

To save time, that first model could have used `do_fit = FALSE`.

The main-effect coefficients are in `par$b_j`, in the same order as in the model output: the intercept, then `depth_scaled`, then `depth_scaled2`.
Here's a normal prior with mean 0 and standard deviation 0.5 on the `depth_scaled` coefficient, the second element:

```{r fit-b}
fit_b <- update(fit, priors = sdmTMBpriors(
  custom = function(par, theta) {
    RTMB::dnorm(par$b_j[2], 0, 0.5, log = TRUE)
  }
))
tidy(fit)
tidy(fit_b)
```

The prior shrinks the depth coefficient towards zero.

This particular prior is also available as a built-in prior, which gives the same answer:

```{r fit-b-builtin}
fit_b2 <- update(fit, priors = sdmTMBpriors(
  b = normal(c(NA, 0, NA), c(NA, 0.5, NA))
))
tidy(fit_b2)
```

Custom priors are especially useful for what the built-in priors can't do --- put any prior on any parameter or transformed parameter.
The rest of this article shows some examples.

## Finding the parameters

As shown above, `get_prior_parameters()` lists every element of `par` and `theta`, how to refer to it, any coefficient label, and its value.
It leaves out raw parameters that are fixed in this model, since a prior on a constant has no effect.
`theta` lists every transformation, including ones of parameters this model doesn't use (such as `rho_sar` in this case), so check that the `par` elements behind a `theta` element are listed before putting a prior on it.
By default, it also leaves out the random effects (here, the many random field values); use `random = TRUE` to include them.

The `expression` column shows how to refer to each element inside your prior function.
For example, the `depth_scaled` coefficient is `par$b_j[2]` and the Matérn range is `theta$range[1, 1]`.
Matrix columns are model components (only one here; two for a delta model), and the rows of `ln_kappa` and `range` are the spatial and spatiotemporal fields.
The `"shared"` status shows that the two rows of `ln_kappa` are one estimated parameter here, since `share_range = TRUE` by default.

You can also call `get_prior_parameters()` on a model built with `do_fit = FALSE`, which is useful when writing priors without having to wait for an initial model to fit.

## Priors on transformed parameters

Priors can also go on natural-scale parameters in `theta`.
Suppose we think the spatial range is about 50 km and the spatiotemporal standard deviation is probably less than 1.
We'll put a normal penalty on the log of the range and a half-normal prior on `sigma_E`.
A normal density on `log(range)` isn't quite a lognormal prior on `range`, which would also include a `-log(range)` term; the next section explains when that difference matters.
Several priors can go in one function. Just return them together and optionally give each a name:

```{r range}
fit_range <- update(fit, priors = sdmTMBpriors(
  custom = function(par, theta) {
    c(
      range = RTMB::dnorm(log(theta$range[1, 1]), log(50), 0.3, log = TRUE),
      sigma_E = RTMB::dnorm(theta$sigma_E[1, 1], 0, 0.5, log = TRUE)
    )
  }
))
tidy(fit, "ran_pars")
tidy(fit_range, "ran_pars")
```

The priors pull the estimates towards these values, but the data still have most of the influence.

`get_prior_densities()` shows the log density of each named term at the estimates:

```{r range-terms}
get_prior_densities(fit_range)
```

This function only shows the custom priors.

Custom priors are added to any built-in priors also passed to `sdmTMBpriors()`. I.e., they can and will be stacked on each other, so you probably don't want to specify the same parameter in both.

## A more advanced example: smoothing year effects

Built-in priors treat each coefficient on its own.
A custom prior can even penalize differences between coefficients.
Here we fit a separate mean for each year and shrink each year's mean towards the previous survey's, like a random walk.
Since the surveys aren't evenly spaced, the standard deviation grows with the square root of the gap between surveys.

```{r rw}
years <- sort(unique(pcod$year))
n_yr <- length(years)
rw_sd <- 0.2 * sqrt(diff(years))

fit_yr <- update(fit, formula. = density ~ 0 + as.factor(year) + depth_scaled + depth_scaled2)
fit_rw <- update(fit_yr, priors = sdmTMBpriors(
  custom = function(par, theta) {
    yr <- par$b_j[seq_len(n_yr)] # the year effects are the first coefficients
    sum(RTMB::dnorm(diff(yr), 0, rw_sd, log = TRUE))
  }
))
```

The function uses `n_yr` and `rw_sd` from the environment where it was defined.
That's the way to pass fixed values (hyperparameters) to a prior.

Comparing the year effects shows that the penalty pulls them together:

```{r rw-plot}
yr_est <- function(x, label) {
  out <- tidy(x, conf.int = TRUE)[seq_len(n_yr), ]
  out$year <- years
  out$model <- label
  out
}
d <- rbind(yr_est(fit_yr, "No prior"), yr_est(fit_rw, "Random walk prior"))
ggplot(d, aes(year, estimate, ymin = conf.low, ymax = conf.high, colour = model)) +
  geom_pointrange(position = position_dodge(width = 0.6)) +
  labs(x = "Year", y = "Year effect (log link)", colour = NULL)
```

This example shows how far custom priors can go; most uses will be simpler.

## A gamma prior on a time-varying SD

Instead of fixed year effects, we can let the intercept follow a random walk with `time_varying = ~ 1`.
The SD of the random walk steps is `theta$sigma_V` (its raw parameter is `par$ln_tau_V = log(sigma_V)`).
The rows are the time-varying coefficients (only the intercept here) and the columns are model components.

```{r tv}
fit_tv <- sdmTMB(
  density ~ 0 + depth_scaled + depth_scaled2,
  data = pcod,
  mesh = mesh,
  family = tweedie(),
  time = "year",
  time_varying = ~ 1,
  time_varying_type = "rw0",
  spatiotemporal = "off",
  extra_time = min(pcod$year):max(pcod$year)
)
get_prior_parameters(fit_tv) |> subset(name %in% c("ln_tau_V", "sigma_V"))
```

A gamma prior is often easier to specify through its mean and coefficient of variation (CV) than its shape and scale.
A gamma with mean `mu` and CV `cv` has `shape = 1 / cv^2` and `scale = cv^2 * mu`.
Here's a prior with a mean of 0.2 and a CV of 0.3 on the SD of the year-to-year changes in the intercept:

```{r tv-gamma}
mu <- 0.2
cv <- 0.3
fit_tv_gamma <- update(fit_tv, priors = sdmTMBpriors(
  custom = function(par, theta) {
    c(sigma_V = RTMB::dgamma(theta$sigma_V[1, 1],
      shape = 1 / cv^2, scale = cv^2 * mu, log = TRUE))
  }
))
get_prior_parameters(fit_tv) |> subset(name == "sigma_V")
get_prior_parameters(fit_tv_gamma) |> subset(name == "sigma_V")
```

This prior is also available built in as `sdmTMBpriors(sigma_V = gamma_cv(mu, cv))`, which gives the same fit.
The custom version is a starting point for other distributions or for linking priors across coefficients.
For a Bayesian fit, the log Jacobian from `ln_tau_V` to `sigma_V` is `par$ln_tau_V[1, 1]`. We'll talk more about that next.

## Penalties versus Bayesian priors and Jacobians

When fitting by maximum likelihood (the default), a custom prior is a penalty on the likelihood, and the optimum is the posterior mode on the scale the prior is written on.
No Jacobian is needed.
For a Bayesian fit (see the [Bayesian vignette](https://sdmTMB.github.io/sdmTMB/articles/bayesian.html)), a prior on a transformed parameter needs a Jacobian adjustment, because the sampler works with the raw parameters.

sdmTMB doesn't automatically work out the Jacobian for a custom prior for you.
Instead, supply its log in `custom_log_jacobian`, which is only used with `bayesian = TRUE`.
For the normal prior on the log range, the raw parameter is `ln_kappa` and `log(range) = log(sqrt(8)) - ln_kappa`.
The absolute derivative of `log(range)` with respect to `ln_kappa` is 1, so the log Jacobian is 0 and no adjustment is needed.
For the half-normal prior on `phi = exp(ln_phi)`, the log Jacobian is `ln_phi`:

```{r jacobian, eval=FALSE}
fit_bayes <- update(fit,
  bayesian = TRUE,
  priors = sdmTMBpriors(
    custom = function(par, theta) {
      RTMB::dnorm(theta$phi[1], 0, 20, log = TRUE)
    },
    custom_log_jacobian = function(par, theta) {
      par$ln_phi[1]
    }
  )
)
```

A prior directly on a raw parameter as it enters the model, such as `par$ln_phi`, never needs a Jacobian.

### The Matérn range and field standard deviations

The Matérn `range` and the field standard deviations (`sigma_O`, `sigma_E`) are more complex than most parameters because they depend on each other.
The range depends only on `ln_kappa`, but each standard deviation depends on both `ln_kappa` and its `ln_tau_*`.
So the Jacobian for a prior on `sigma_E` alone isn't well defined: it has to come from the joint transformation from (`ln_kappa`, `ln_tau_E`) to (`range`, `sigma_E`).
The log Jacobian of that transformation is `log(range) + log(sigma_E)`.
With a shared range (the default, `share_range = TRUE`), one `ln_kappa` maps to the range for both fields, so add `log(range)` once plus `log(sigma)` for each field.

For example, for a Bayesian fit with the range and `sigma_E` priors from earlier, the prior on `log(range)` and `sigma_E` together needs a log Jacobian of `log(sigma_E)`.
The `log(range)` term from the joint transformation cancels with the `-log(range)` from putting the prior on `log(range)` rather than `range`:

```{r jacobian-range, eval=FALSE}
fit_bayes_range <- update(fit,
  bayesian = TRUE,
  priors = sdmTMBpriors(
    custom = function(par, theta) {
      c(
        range = RTMB::dnorm(log(theta$range[1, 1]), log(50), 0.3, log = TRUE),
        sigma_E = RTMB::dnorm(theta$sigma_E[1, 1], 0, 0.5, log = TRUE)
      )
    },
    custom_log_jacobian = function(par, theta) {
      theta$log_sigma_E[1, 1]
    }
  )
)
```

This model also has a spatial field, which shares the range.
Without a custom prior on `sigma_O`, its `ln_tau_O` keeps a flat prior on its raw scale.

For priors on these parameters, we suggest the built-in penalized complexity (PC) priors instead: `pc_matern()` through the `matern_s` and `matern_st` arguments to `sdmTMBpriors()`.
They put a joint prior on the range and standard deviation of each field and take care of the Jacobian for you when `bayesian = TRUE`.

## Limitations

- Custom priors need the RTMB backend.
- Like the built-in priors, custom priors are part of the model's objective, so they're included in `logLik()` and `AIC()`.
- The names and shapes of `par` and `theta` are internal and could change in future versions. Check them with `get_prior_parameters()`.
