---
title: "Fitting delta (hurdle) models with sdmTMB"
author: "Julia Indivero, Philina English, Sean Anderson, Eric Ward, Lewis Barnett, James Thorson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting delta (hurdle) models with sdmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**If the code in this vignette has not been evaluated, a rendered version is available on the [documentation site](https://sdmTMB.github.io/sdmTMB/index.html) under 'Articles'.**

```{r setup, include = FALSE, cache=FALSE}
dplyr_installed <- require("dplyr", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
pkgs <- dplyr_installed && ggplot_installed
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = EVAL,
  purl = EVAL
)
```

```{r packages, message=FALSE, warning=TRUE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
```

sdmTMB has built-in hurdle models (also called delta models).
These models include one component for zero versus non-zero data and another component for the positive observations.
Hurdle models could also be implemented by fitting the two components separately and combining the predictions.

Hurdle models are more appropriate than something like a Tweedie when there are differences in the processes controlling presence vs. abundance, or when greater flexibility to account for dispersion is required.

Built-in hurdle models can be specified with the `family` argument within the `sdmTMB()` function. Current options include:

1. Delta-Gamma: `family = delta_gamma(link1 = "logit", link2 = "log")`.
This fits a binomial presence-absence model (i.e., `binomial(link = "logit")`) and then a model for the positive catches only with a Gamma observation distribution and a log link (i.e., `Gamma(link = "log")`).
Here and with other delta models, the `link1` and `link2` can be omitted and left at their default values.

2. Delta-lognormal: `family = delta_lognormal()`.
This fits a binomial presence-absence model (i.e., `binomial(link = "logit")`) and then a model for the positive catches only with a lognormal observation distribution and a log link (i.e., `lognormal(link = "log")`).

3. Delta-generalized-gamma: `family = delta_gengamma()`.
This fits a binomial presence-absence model and then a model for the positive catches with a generalized gamma observation distribution and a log link (i.e., `gengamma(link = "log")`).

4. Poisson-link delta-Gamma, delta-lognormal, or delta-generalized-gamma. These are specified with `delta_gamma(type = "poisson-link")`, `delta_lognormal(type = "poisson-link")`, or `delta_gengamma(type = "poisson-link")`. See the [Poisson-link delta model vignette](https://sdmTMB.github.io/sdmTMB/articles/poisson-link.html).

5. Delta-mixture families: `family = delta_gamma_mix()` or `family = delta_lognormal_mix()`.
These fit a binomial presence-absence model and then a positive-catch model using either a gamma-mixture or lognormal-mixture distribution.

6. Delta-truncated-negative-binomial: `family = delta_truncated_nbinom1()` or `family = delta_truncated_nbinom2()`.
This fits a binomial presence-absence model (`binomial(link = "logit")`) and a `truncated_nbinom1(link = "log")` or `truncated_nbinom2(link = "log")` distribution for positive catches.

7. Delta-beta: `family = delta_beta()`.
This is intended for continuous proportions with exact zeros, combining a binomial presence-absence model with a beta model for the positive component.

To summarize the built-in delta models and the separate components:

Model Type               |Built-in delta function         | Presence-absence model            | Positive-component model               |
-------------------------|--------------------------------|-----------------------------------|----------------------------------------|
Delta-gamma              |`delta_gamma()`                 |`binomial(link = "logit")`         |`Gamma(link = "log")`                   |
Delta-lognormal          |`delta_lognormal()`             |`binomial(link = "logit")`         |`lognormal(link = "log")`               |
Delta-generalized-gamma  |`delta_gengamma()`              |`binomial(link = "logit")`         |`gengamma(link = "log")`                |
Delta-gamma-mixture      |`delta_gamma_mix()`             |`binomial(link = "logit")`         |`gamma_mix(link = "log")`               |
Delta-lognormal-mixture  |`delta_lognormal_mix()`         |`binomial(link = "logit")`         |`lognormal_mix(link = "log")`           |
Delta-NB1                |`delta_truncated_nbinom1()`     |`binomial(link = "logit")`         |`truncated_nbinom1(link = "log")`       |
Delta-NB2                |`delta_truncated_nbinom2()`     |`binomial(link = "logit")`         |`truncated_nbinom2(link = "log")`       |
Delta-beta               |`delta_beta()`                  |`binomial(link = "logit")`         |`Beta(link = "logit")`                  |


# Example with built-in delta model

Here, we will show an example of fitting using the built-in delta functionality, as well as how to build each model component separately and then combine them.
The built-in approach is convenient, allows for parameters to be shared across components, and allows for calculation of derived quantities such as standardized indexes (`get_index()`) with internally calculated standard errors.

We will use a dataset built into the sdmTMB package: trawl survey data for Pacific Cod in Queen Charlotte Sound, British Columbia, Canada. The density units are kg/km^2^. Here, X and Y are coordinates in UTM zone 9.

We will first create a mesh that we will use for all the models.

```{r}
pcod_mesh <- make_mesh(pcod, c("X", "Y"), cutoff = 15)
```

Then we can fit a model of Pacific cod density using a delta-gamma model, including a smoothed effect of depth.

```{r, warning=FALSE, message=FALSE}
fit_dg <- sdmTMB(density ~ 1 + s(depth),
  data = pcod,
  mesh = pcod_mesh,
  time = "year",
  family = delta_gamma()
)
```

The default in built-in delta models is for the formula, spatial and spatiotemporal structure, and anisotropy to be shared between the two model components.
However, some elements (`formula`, `spatial`, `spatiotemporal`, and `share_range`) can also be specified independently for each model using a list format within the function argument (see examples below).
The first element of the list is for the binomial component and the second element is for the positive component (e.g., Gamma).
Some elements must be shared for now (e.g., smoothers, spatially varying coefficients, and time-varying coefficients).

To specify the settings for spatial and spatiotemporal effects in each model component, create a list of settings within the `spatial` and `spatiotemporal` arguments. For example, `spatial = list("on", "off"), spatiotemporal = list("off", "rw")`.

```{r, warning=FALSE, message=FALSE, echo=FALSE, eval=FALSE}
fit_dg2 <- sdmTMB(density ~ 1 + s(depth),
  data = pcod,
  mesh = pcod_mesh,
  time = "year",
  spatial = list("on", "off"), #<
  spatiotemporal = list("off", "rw"), #<
  family = delta_gamma()
)
```

We could similarly specify a different formula for each component of the model, using `list(y ~ x1, y ~ x2)`
For instance, we could include the effect of depth for only the positive model, and remove it for the presence-absence model.

However, there are currently limitations if specifying separate formulas for each model component. The two formulas *cannot* have:

* smoothers
* threshold effects
* random intercepts

For now, these must be specified through a single formula that is shared across the two models.

```{r, warning=FALSE, message=FALSE, echo=FALSE, results='hide', eval=FALSE}
fit_dg3 <- sdmTMB(
  list(
    density ~ 1, #<
    density ~ poly(depth, 2) #<
  ),
  data = pcod,
  mesh = pcod_mesh,
  time = "year",
  spatial = list("off", "on"),
  spatiotemporal = list("off", "rw"),
  family = delta_gamma()
)
```

Each model component can similarly have separate settings for `share_range`, which determines whether there is a shared spatial and spatiotemporal range parameter (`TRUE`) or independent range parameters (`FALSE`), by using a list.

```{r, warning=FALSE, message=FALSE, echo=FALSE, results='hide', eval=FALSE}
fit_dg4 <- sdmTMB(density ~ 1 + s(depth),
  data = pcod,
  mesh = pcod_mesh,
  time = "year",
  share_range = list(TRUE, FALSE), #<
  family = delta_gamma()
)
```

Lastly, whether anisotropy is included in the model is determined with the logical argument `anisotropy` (i.e., `TRUE` or `FALSE`) and cannot be separately specified for each model. If anisotropy is included, it is by default shared across the two model components. However, it can be made unique in each model component by using `sdmTMBcontrol(map = ...)` and adding the argument `control` when fitting the model. This "maps" the anisotropy parameters to be unique across model components.

```{r, warning=FALSE, message=FALSE, echo=FALSE, results='hide', eval=FALSE}
fit_dg5 <- sdmTMB(density ~ 1 + s(depth),
  data = pcod,
  mesh = pcod_mesh,
  time = "year",
  control = sdmTMBcontrol(map = list(ln_H_input = factor(c(1, 2, 3, 4)))), #<
  family = delta_gamma()
)
```

Once we fit the delta model, we can evaluate and plot the output, diagnostics, and predictions similar to other sdmTMB models.

The printed model output will show estimates and standard errors of parameters for each model separately.

```{r}
print(fit_dg)
```

Using the `tidy()` function will turn the sdmTMB model output into a data frame, with the argument `model = 1` or `model = 2` specifying which model component to extract. See `tidy.sdmTMB()` for additional arguments and options.

```{r}
tidy(fit_dg) # model = 1 is default
tidy(fit_dg, model = 1)
tidy(fit_dg, model = 1, "ran_pars", conf.int = TRUE)
tidy(fit_dg, model = 2)
tidy(fit_dg, model = 2, "ran_pars", conf.int = TRUE)
```

<!-- We can use `predict()` to predict density in either log space (the default) or transformed into the response scale (`type = "response"`). -->

For built-in delta models, the default `predict()` call returns estimated values for each grid cell for each model component separately, denoted with a `1` (for the presence-absence model) or `2` (for the positive-catch model) in the column name. See `predict.sdmTMB()` for a description of the values in the returned data frame.

```{r, error=FALSE, message=FALSE}
grid_yrs <- replicate_df(qcs_grid, "year", unique(pcod$year))
p <- predict(fit_dg, newdata = grid_yrs)
str(p)
```

```{r, eval=FALSE, echo=FALSE}
# p_response <- predict(fit_dg, newdata = grid_yrs, type = "response")
```

<!-- We can use the `model` argument to return only predictions from model 1 or model 2. -->

```{r, error=FALSE, message=FALSE, echo=FALSE, eval=FALSE}
# p1 <- predict(fit_dg, newdata = grid_yrs, model = 1)
# p2 <- predict(fit_dg, newdata = grid_yrs, model = 2)
```

<!-- To return combined predictions from both components on the response scale, we can use `model=NA` -->

```{r, error=FALSE, message=FALSE, echo=FALSE, eval=FALSE}
# p_combined <- predict(fit_dg, newdata = grid_yrs, model = NA, type = "response")
```

We can pass the built-in delta model and prediction grid directly to `get_index()` to calculate index values.
The `get_index()` function will automatically combine predictions from the first and second model in calculating the index values.
For more on modelling for the purposes of creating an index see the vignette on [Index standardization with sdmTMB](https://sdmTMB.github.io/sdmTMB/articles/index-standardization.html).

```{r, message=FALSE, warning=FALSE, eval=FALSE}
ind_dg <- get_index(fit_dg, newdata = grid_yrs, bias_correct = FALSE)
```

We can plot conditional effects of covariates (such as depth in the example model) using the package `visreg` by specifying each model component with `model = 1` for the presence-absence model or `model = 2` for the positive-catch model. Currently, plotting effects of built-in delta models with `ggeffects` is not supported. See the vignette on using [visreg](https://sdmTMB.github.io/sdmTMB/articles/visreg.html) with sdmTMB for more information.

```{r}
visreg_delta(fit_dg, xvar = "depth", model = 1)
```

```{r eval=FALSE}
visreg_delta(fit_dg, xvar = "depth", model = 2)
```

The built-in delta models can also be evaluated with the `residuals()` function in sdmTMB.
Similarly to generating predictions, we can specify which model component we want to return residuals for using the `model` argument and specifying `1` or `2`.
See `residuals.sdmTMB()` for additional options for evaluating residuals in sdmTMB models.

```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE, eval=FALSE}
residuals_1 <- residuals(fit_dg, type = "response", model = 1)
residuals_2 <- residuals(fit_dg, type = "response", model = 2)
```

We can also simulate new observations from a fitted delta model. As in other functions, we can specify which model to simulate from using the argument `model = 1` for presence-absence only, `model = 2` for positive catches only, or `model = NA` for combined predictions. See `simulate.sdmTMB()` for more details on simulation options.

```{r, message=FALSE, warning=FALSE, eval=FALSE}
simulations <- simulate(fit_dg, nsim = 5, seed = 5090, model = NA)
```

# Delta models by fitting two components separately and combining predictions

Next, we will show an example of how to implement a delta-gamma model in sdmTMB with each component fit separately and then combined.
This approach gives maximum flexibility for each model and lets you develop them one at a time.
It has limitations if you are calculating an index of abundance or if you want to share some parameters.

```{r glimpse-pcod}
glimpse(pcod)
```

```{r spde, fig.asp=0.7}
mesh1 <- make_mesh(pcod, c("X", "Y"), cutoff = 20) # coarse for vignette speed
```

It is not necessary to use the same mesh for both models, but one can do so by updating the first mesh to match the reduced data frame as shown here:

```{r, fig.asp=0.9}
dat2 <- subset(pcod, density > 0)
mesh2 <- make_mesh(dat2,
  xy_cols = c("X", "Y"),
  mesh = mesh1$mesh
)
```

This delta-gamma model is similar to the Tweedie model in the [Intro to modelling with sdmTMB](https://sdmTMB.github.io/sdmTMB/articles/basic-intro.html) vignette, except that we will use `s()` for the depth effect.

```{r model1, warning=FALSE}
m1 <- sdmTMB(
  formula = present ~ 0 + as.factor(year) + s(depth, k = 3),
  data = pcod,
  mesh = mesh1,
  time = "year", family = binomial(link = "logit"),
  spatiotemporal = "iid",
  spatial = "on"
)
m1
```

One can use different covariates in each model, but in this case we will just let the depth effect be more wiggly by not specifying `k = 3`.

```{r model2, warning=FALSE}
m2 <- sdmTMB(
  formula = density ~ 0 + as.factor(year) + s(depth),
  data = dat2,
  mesh = mesh2,
  time = "year",
  family = Gamma(link = "log"),
  spatiotemporal = "iid",
  spatial = "on"
)
m2
```

Next, we need some way of combining the predictions across the two models.
If all we need are point predictions, we can just multiply the predictions from the two models after applying the inverse link:

```{r delta-pred-simple, echo=TRUE}
pred <- grid_yrs # use the grid as template for saving our predictions
p_bin <- predict(m1, newdata = grid_yrs)
p_pos <- predict(m2, newdata = grid_yrs)
p_bin_prob <- m1$family$linkinv(p_bin$est)
p_pos_exp <- m2$family$linkinv(p_pos$est)
pred$est_exp <- p_bin_prob * p_pos_exp
```

But if a measure of uncertainty is required, we can simulate from the joint parameter precision matrix using the `predict()` function with any number of simulations selected (e.g., `sims = 500`).
Because the predictions come from simulated draws from the parameter covariance matrix, the predictions will become more consistent with a larger number of draws.
However, a greater number of draws takes longer to calculate and will use more memory (larger matrix), so fewer draws (~100) may be fine for experimentation.
A larger number (say ~1000) may be appropriate for final model runs.

```{r delta-pred-sim, echo=TRUE}
set.seed(28239)
p_bin_sim <- predict(m1, newdata = grid_yrs, nsim = 100)
p_pos_sim <- predict(m2, newdata = grid_yrs, nsim = 100)
p_bin_prob_sim <- m1$family$linkinv(p_bin_sim)
p_pos_exp_sim <- m2$family$linkinv(p_pos_sim)
p_combined_sim <- p_bin_prob_sim * p_pos_exp_sim
```

`p_combined_sim` is a matrix with a row for each row of data that was predicted on and width `nsim`.
You can process this matrix however you would like.
We can save median predictions and upper and lower 95% confidence intervals:

```{r, fig.width=5}
pred$median <- apply(p_combined_sim, 1, median)
plot(pred$est_exp, pred$median)
```

```{r}
ggplot(subset(pred, year == 2017), aes(X, Y, fill = median)) +
  geom_raster() +
  coord_fixed() +
  scale_fill_viridis_c(trans = "sqrt")
```

And we can calculate spatial uncertainty:

```{r cv}
pred$cv <- apply(p_combined_sim, 1, function(x) sd(x) / mean(x))
ggplot(subset(pred, year == 2017), aes(X, Y, fill = cv)) + # 2017 as an example
  geom_raster() +
  coord_fixed() +
  scale_fill_viridis_c(trans = "log10")
```

<!-- sdmTMB also has a function for calculating an index from those draws `get_index_sims()`. -->
<!-- This function is just summing biomass or abundance across grid cells for each simulation draw and for each year and then calculating quantiles on the distribution of samples. -->
<!-- The default for this function expects the simulations to still be in log space. -->
<!-- This index from simulated draws is an approximation.  -->
<!-- We recommend using the built-in delta model and `get_index(..., bias_correct = TRUE)` for final inference in important applications. -->

```{r get_index, eval=FALSE, echo=FALSE}
grid_yrs$area <- 4 # all 2 x 2km
p_log_combined <- log(p_combined_sim / 1000) # log space; convert from kg to tonnes
attr(p_log_combined, "link") <- "log" # suppresses a warning below

ind <- get_index_sims(
  p_log_combined,
  area = grid_yrs$area
)
ggplot(ind, aes(year, est)) +
  geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4) +
  ylab("Biomass (t)")
```
