---
title: "Accounting for hook competition with a censored Poisson"
bibliography: '../refs.bib'
csl: '../mee.csl'
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Accounting for hook competition with a censored Poisson}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.asp = 0.618
)
```

```{r packages}
library(sdmTMB)
library(ggplot2)
```

# Introduction

A longline set has a fixed number of baited hooks.
Once most baits have been taken, by the target species, other species, or lost to mechanical failure, fish of the target species have fewer chances to find a baited hook.
Catch counts from sets with high bait removal therefore underestimate what would have been caught without hook competition.
This bias can differ by year, depending on how many sets were saturated.

@watson2023 proposed treating catch counts from these sets as right-censored.
For a set where the proportion of baits removed reaches a species-specific breakdown point $p^*$, the observed count $c$ is treated as a lower bound on the count that would have been observed without hook competition.
For all other sets, the count is used as usual:

$$
\Pr(c \mid \mu) =
\begin{cases}
  f(c; \mu), & p < p^* \ \text{(uncensored)},\\
  1 - F(c - 1; \mu), & p \ge p^* \ \text{(right-censored)},
\end{cases}
$$

where $f$ and $F$ are the Poisson probability mass function and cumulative distribution function and $p$ is the proportion of baits removed.
A censored zero has a likelihood of exactly 1, so it contributes no information.

@watson2023 used simulations to show that this approach recovers relative abundance trends much better than raw CPUE or instantaneous catch rate corrections.
It is available in sdmTMB through the `censored_poisson()` family.

# The data

We'll use the `yelloweye` dataset that comes with sdmTMB.
It contains Yelloweye Rockfish catch counts from the Hard Bottom Longline Survey (South) off the west coast of Vancouver Island, British Columbia.
For each set, we know the number of hooks deployed and the number of hooks that came back still baited.

```{r data}
d <- yelloweye
d$prop_removed <- 1 - d$bait_returned / d$hook_count
d$fyear <- as.factor(d$year)
d$log_depth <- log(d$depth)
d$obs_id <- as.factor(seq_len(nrow(d)))
head(d)
```

In many sets, nearly all baits were removed:

```{r hist}
ggplot(d, aes(prop_removed)) +
  geom_histogram(bins = 30) +
  labs(x = "Proportion of baits removed", y = "Number of sets")
```

# Choosing the breakdown point

@watson2023 suggest choosing $p^*$ by looking at how mean catch changes as bait removal increases.
They fit a smoother of catch against the proportion of baits removed, accounting for year and with log hook count as an offset, and look for where the curve starts to decline.
We can do this with mgcv:

```{r gam}
library(mgcv)
fit_gam <- gam(
  catch_count ~ s(prop_removed) + fyear + offset(log(hook_count)),
  family = nb(), data = d
)
nd <- data.frame(
  prop_removed = seq(min(d$prop_removed), 1, length.out = 200),
  fyear = d$fyear[1],
  hook_count = 1
)
p <- predict(fit_gam, newdata = nd, se.fit = TRUE)
nd$est <- p$fit
nd$se <- p$se.fit
ggplot(nd, aes(prop_removed, exp(est - max(est)))) +
  geom_ribbon(aes(ymin = exp(est - 2 * se - max(est)),
    ymax = exp(est + 2 * se - max(est))), alpha = 0.3) +
  geom_line() +
  geom_vline(xintercept = 0.85, lty = 2) +
  labs(x = "Proportion of baits removed", y = "Relative mean catch")
```

Mean catch rises with bait removal, because sets with more fish remove more baits, until about 0.85.
After that it drops, consistent with hook competition.
@watson2023 also chose $p^* = 0.85$ for Yelloweye Rockfish, so we'll use that value.

```{r pstar}
pstar <- 0.85
mean(d$prop_removed >= pstar)
```

About `r round(100 * mean(d$prop_removed >= pstar))`% of sets will be treated as censored.
The proportion varies by year, and that variation is what biases an uncorrected index:

```{r prop-censored}
round(tapply(d$prop_removed >= pstar, d$year, mean), 2)
```

# Fitting the models

We specify censoring with the `censored_upper` argument in `sdmTMBcontrol()`.
It is a vector with one element per row of data:

- a value equal to the observed count means the observation is **not** censored;
- `NA` means the observation is right-censored with no upper bound;
- a value greater than the observed count means the observation is interval-censored between the observed count and that value.

The response is the observed catch count.

```{r upr}
upr <- ifelse(d$prop_removed >= pstar, NA, d$catch_count)
```

We'll fit a model with year effects, a quadratic effect of log depth, a spatial random field, and an observation-level random intercept, `(1 | obs_id)`.
The observation-level intercept makes this a Poisson-lognormal model and accounts for overdispersion, as in @watson2023.
The offset is log hook count, so the index is relative catch per hook.

First, a standard Poisson model that ignores hook competition:

```{r fit-pois}
mesh <- make_mesh(d, c("X", "Y"), cutoff = 10)

fit_pois <- sdmTMB(
  catch_count ~ 0 + fyear + poly(log_depth, 2) + (1 | obs_id),
  offset = log(d$hook_count),
  data = d,
  mesh = mesh,
  family = poisson(),
  time = "year",
  spatial = "on",
  spatiotemporal = "off"
)
```

Now the censored version.
We only need to change the family and pass `censored_upper`:

```{r fit-cens}
fit_cens <- update(
  fit_pois,
  family = censored_poisson(),
  control = sdmTMBcontrol(censored_upper = upr)
)
sanity(fit_cens)
fit_cens
```

## Adding an upper bound

When many observations are censored and the data are limited, the right-censored model may not converge.
@watson2023 describe an optional upper bound on the censored counts (their Supplementary Material S.3), turning right-censored observations into interval-censored ones.
The bound adds the expected number of target-species catches "missed" because of hook competition, based on the number of hooks emptied past the breakdown point.
The function `get_censored_upper()` calculates this bound:

```{r upr-bound}
upr_bounded <- get_censored_upper(
  prop_removed = d$prop_removed,
  n_catch = d$catch_count,
  n_hooks = d$hook_count,
  pstar = pstar
)
fit_cens_bounded <- update(
  fit_pois,
  family = censored_poisson(),
  control = sdmTMBcontrol(censored_upper = upr_bounded)
)
```

@watson2023 also suggest treating the largest catch counts each year as uncensored, regardless of bait removal, to help convergence.
To do that, set `censored_upper` equal to `catch_count` for those rows.

# Comparing indices

We'll predict over the survey grid `hbll_s_grid` and calculate an index for each model.
We set the offset to 0 for prediction (i.e., one hook) and leave out the observation-level random intercepts with `re_form_iid = NA`.
The grid contains depths outside the sampled range, so we clamp depth to the range of the data so that the quadratic is not extrapolated.

```{r grid}
grid <- replicate_df(hbll_s_grid, "year", unique(d$year))
grid$fyear <- as.factor(grid$year)
grid$log_depth <- log(pmin(pmax(grid$depth, min(d$depth)), max(d$depth)))
```

```{r index}
get_ind <- function(fit, label, ...) {
  ind <- get_index(
    fit, newdata = grid, offset = rep(0, nrow(grid)),
    predict_args = list(re_form_iid = NA),
    bias_correct = TRUE, ...
  )
  ind$model <- label
  ind
}
ind <- rbind(
  get_ind(fit_pois, "Poisson"),
  get_ind(fit_cens, "Censored Poisson"),
  get_ind(fit_cens_bounded, "Censored Poisson (upper bound)")
)
```

Since these are relative indices, we'll scale each by its geometric mean before plotting:

```{r plot-index}
scale_index <- function(ind) {
  do.call(rbind, lapply(split(ind, ind$model), function(x) {
    gm <- exp(mean(log(x$est)))
    x[, c("est", "lwr", "upr")] <- x[, c("est", "lwr", "upr")] / gm
    x
  }))
}
plot_index <- function(ind) {
  ggplot(ind, aes(year, est, colour = model, fill = model)) +
    geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.15, colour = NA) +
    geom_line() +
    geom_point() +
    labs(x = "Year", y = "Relative index", colour = NULL, fill = NULL) +
    theme(legend.position = "bottom")
}
ind <- scale_index(ind)
plot_index(ind)
```

The censored models differ most from the uncensored Poisson model in years with an unusually high or low proportion of censored sets.
For example, 2011 had the highest proportion of saturated sets, and its index is higher once we account for hook competition.
2014 had very few saturated sets, and its index is lower relative to the other years.
Here, the upper bound mostly has a modest effect, pulling some years (e.g., 2022) partway back towards the Poisson estimates by limiting how large the censored counts can be.

# A censored beta-binomial

A Poisson model has no upper limit on the count, but a set can't catch more fish than it has hooks.
An alternative is a beta-binomial with the number of hooks as the number of trials.
With a complementary log-log link, the mean probability that a hook catches the target species is $p = 1 - \exp(-\exp(\eta))$, where $\eta$ is the linear predictor.
If each hook caught fish as a Poisson process with rate $\lambda$, then $p = 1 - \exp(-\lambda)$ and $\exp(\eta) = \lambda$ would be a per-hook catch rate.
The beta-binomial dispersion parameter accounts for overdispersion instead of an observation-level random intercept, so the two models describe overdispersion differently.

The `censored_betabinomial()` family works like `censored_poisson()`.
The latent count without hook competition still can't exceed the number of hooks, so a right-censored set has a count between the observed count and the number of hooks:

$$
\Pr(c \mid p, \phi) = \Pr(c \le Y \le n), \quad Y \sim \operatorname{BetaBinomial}(n, p, \phi),
$$

where $n$ is the number of hooks.
This gives every censored set an upper bound, so we can use the same `censored_upper` vector with `NA` for censored sets and don't need to calculate one.
The number of hooks is supplied through `weights`.
Counts must be whole numbers, and a non-integer bound means the largest count it allows (e.g., 2.5 means 2).

As with the censored Poisson, this is a true interval-censoring likelihood only if which sets are censored, and their bounds, carry no extra information about the parameters.
Hook competition doesn't censor counts at random: it reduces catches through a process that also depends on the abundance of the target and other species.
The censored likelihood is therefore best thought of as a working likelihood, and how well it recovers trends depends on how competition actually reduces catches [see the simulations in @watson2023].

```{r fit-bb}
fit_bb <- sdmTMB(
  catch_count ~ 0 + fyear + poly(log_depth, 2),
  weights = d$hook_count,
  data = d,
  mesh = mesh,
  family = betabinomial(link = "cloglog"),
  time = "year",
  spatial = "on",
  spatiotemporal = "off"
)
fit_bb_cens <- update(
  fit_bb,
  family = censored_betabinomial(link = "cloglog"),
  control = sdmTMBcontrol(censored_upper = upr)
)
sanity(fit_bb_cens)
```

The beta-binomial distribution function has no closed form, so the likelihood of each censored set is a sum over every possible count in its interval.
Fitting is therefore slower than for the censored Poisson when sets have many hooks.

For the index, `derived_link = "log"` sums $\exp(\eta) = -\log(1 - p)$ over the grid instead of the probability $p$, which can't exceed 1.
This is a transformation of the mean catch probability.
Because the beta-binomial mixes over hook-level probabilities, it isn't in general the mean of an underlying Poisson catch rate, so treating it as proportional to abundance is an additional assumption:

```{r index-bb}
ind_bb <- rbind(
  get_ind(fit_bb, "Beta-binomial", derived_link = "log"),
  get_ind(fit_bb_cens, "Censored beta-binomial", derived_link = "log")
)
ind_bb <- scale_index(ind_bb)
plot_index(rbind(ind_bb, ind[ind$model == "Censored Poisson", ]))
```

Accounting for hook competition moves the beta-binomial index in the same direction as the Poisson index, e.g., up in 2011 and down in 2014, but by less here.
The two models account for overdispersion differently (a beta-binomial dispersion parameter versus a lognormal observation-level random intercept) and their indices are on different scales, so differences in their indices and uncertainty aren't only due to the upper limit on the count.

# Summary

- Calculate the proportion of baits removed for each set.
- Choose a breakdown point $p^*$, e.g., by plotting a smoother of catch against bait removal.
- Build a `censored_upper` vector: the observed count for uncensored sets and `NA` (or an upper bound) for sets at or above $p^*$.
- Fit with `family = censored_poisson()` and `control = sdmTMBcontrol(censored_upper = ...)`, ideally with an observation-level random intercept to account for overdispersion.
- Alternatively, fit with `family = censored_betabinomial(link = "cloglog")` and the number of hooks as `weights`, which limits each count to the number of hooks.

See @watson2023 for the full method, simulation tests, and an application to 11 species.

# References
