---
title: "Accounting for hook competition with censored count models"
bibliography: '../refs.bib'
csl: '../mee.csl'
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Accounting for hook competition with censored count models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  error = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.asp = 0.618
)
```

```{r packages}
library(sdmTMB)
library(ggplot2)
```

# Introduction

In fisheries-indepedent surveys, longline set has a fixed number of baited hooks.
Once most baits have been taken by fish or lost to mechanical failure, fish have fewer chances to find a baited hook.
Catch counts from sets with high bait removal therefore underestimate what would have been caught without hook competition.
This bias can differ by year, depending on how many sets were saturated.

@watson2023 proposed treating catch counts from these sets as right-censored.
For a set where the proportion of baits removed reaches a species-specific breakdown point $p^*$, the observed count $c$ is treated as a lower bound on the count that would have been observed without hook competition.
For all other sets, the count is used as usual:

$$
\Pr(c \mid \mu) =
\begin{cases}
  f(c; \mu), & p < p^* \ \text{(uncensored)},\\
  1 - F(c - 1; \mu), & p \ge p^* \ \text{(right-censored)},
\end{cases}
$$

where $f$ and $F$ are the Poisson probability mass function and cumulative distribution function and $p$ is the proportion of baits removed.
A censored zero has a likelihood of exactly 1, so it contributes no information.

@watson2023 used simulations to show that this approach recovers relative abundance trends better than raw CPUE or instantaneous catch rate corrections.
This censored Poisson approach is available in sdmTMB through the `censored_poisson()` family.
We also show an instantaneous catch rate correction for comparison, censored negative binomial models, and censored binomial and beta-binomial models, which limit each count to the number of hooks.

# The data

We'll use the `yelloweye` dataset that comes with sdmTMB.
It contains Yelloweye Rockfish catch counts from the Hard Bottom Longline Survey (South) off the west coast of Vancouver Island, British Columbia.
For each set, we know the number of hooks deployed and the number of hooks that came back still baited.

```{r data}
d <- yelloweye
d$prop_removed <- 1 - d$bait_returned / d$hook_count
d$fyear <- as.factor(d$year)
d$log_depth <- log(d$depth)
d$obs_id <- as.factor(seq_len(nrow(d)))
head(d)
```

In some sets, nearly all baits were removed:

```{r hist}
ggplot(d, aes(prop_removed)) +
  geom_histogram(bins = 30) +
  labs(x = "Proportion of baits removed", y = "Number of sets")
```

# Choosing the breakdown point

@watson2023 suggest choosing $p^*$ (point at which one deems catch rate to be affected strongly by hook comeptition) by looking at how mean catch changes as bait removal increases.
They fit a smoother of catch against the proportion of baits removed, accounting for year and with log hook count as an offset, and look for where the curve starts to decline.
We can do this with mgcv:

```{r gam}
library(mgcv)
fit_gam <- gam(
  catch_count ~ s(prop_removed) + fyear + offset(log(hook_count)),
  family = nb(), data = d
)
nd <- data.frame(
  prop_removed = seq(min(d$prop_removed), 1, length.out = 200),
  fyear = d$fyear[1],
  hook_count = 1
)
p <- predict(fit_gam, newdata = nd, se.fit = TRUE)
nd$est <- p$fit
nd$se <- p$se.fit
ggplot(nd, aes(prop_removed, exp(est - max(est)))) +
  geom_ribbon(aes(ymin = exp(est - 2 * se - max(est)),
    ymax = exp(est + 2 * se - max(est))), alpha = 0.3) +
  geom_line() +
  geom_vline(xintercept = 0.85, lty = 2) +
  labs(x = "Proportion of baits removed", y = "Relative mean catch")
```

Mean catch rises with bait removal, because sets with more fish remove more baits, until about 0.85.
After that it drops, consistent with hook competition.
@watson2023 also chose $p^* = 0.85$ for Yelloweye Rockfish, so we'll use that value.

```{r pstar}
pstar <- 0.85
mean(d$prop_removed >= pstar)
```

About `r round(100 * mean(d$prop_removed >= pstar))`% of sets will be treated as censored.
The proportion varies by year, and that variation is what biases an uncorrected index:

```{r prop-censored}
round(tapply(d$prop_removed >= pstar, d$year, mean), 2)
```

# Fitting the models

We specify censoring with the `censored_upper` argument in `sdmTMB()`.
It is a vector with one element per row of data (or the name of a column in the data):

- a value equal to the observed count means the observation is **not** censored;
- `Inf` means the observation is right-censored with no upper bound;
- a value greater than the observed count means the observation is interval-censored between the observed count and that value.

The response is the observed catch count.

```{r upr}
upr <- ifelse(d$prop_removed >= pstar, Inf, d$catch_count)
```

We'll fit a model with year effects, a quadratic effect of log depth, a spatial random field, and an observation-level random intercept, `(1 | obs_id)`.
The observation-level intercept makes this a Poisson-lognormal model and accounts for overdispersion, as in @watson2023.
The offset is log hook count, so the index is relative catch per hook.

First, a standard Poisson model that ignores hook competition:

```{r fit-pois}
mesh <- make_mesh(d, c("X", "Y"), cutoff = 10)

fit_pois <- sdmTMB(
  catch_count ~ 0 + fyear + poly(log_depth, 2) + (1 | obs_id),
  offset = log(d$hook_count),
  data = d,
  mesh = mesh,
  family = poisson(),
  time = "year",
  spatial = "on",
  spatiotemporal = "off"
)
```

Now the censored version.
We only need to change the family and pass `censored_upper`:

```{r fit-pois-cens}
fit_pois_cens <- update(
  fit_pois,
  family = censored_poisson(),
  censored_upper = upr
)
sanity(fit_pois_cens)
fit_pois_cens
```

## Adding an upper bound

When many observations are censored and the data are limited, the right-censored model may not converge.
@watson2023 describe an optional upper bound on the censored counts (their Supplementary Material S.3), turning right-censored observations into interval-censored ones.
The bound adds the expected number of target-species catches "missed" because of hook competition, based on the number of hooks emptied past the breakdown point.
The function `get_censored_upper()` calculates this bound:

```{r upr-bound}
upr_bounded <- get_censored_upper(
  prop_removed = d$prop_removed,
  n_catch = d$catch_count,
  n_hooks = d$hook_count,
  pstar = pstar
)
fit_pois_cens_bounded <- update(
  fit_pois,
  family = censored_poisson(),
  censored_upper = upr_bounded
)
```

@watson2023 also suggest treating the largest catch counts each year as uncensored, regardless of bait removal, to help convergence.
To do that, set `censored_upper` equal to `catch_count` for those rows.

## An instantaneous catch rate offset

A more traditional correction is the instantaneous catch rate (ICR) [@rothschild1967; @somerton1995], which @watson2023 also compared against.
It assumes each hook loses its bait at a constant rate $\lambda$, to any species or other causes, of which the target species makes up a rate $\lambda_T$.
At the end of the soak, the proportion of hooks still baited is $P_b = \exp(-\lambda)$, and a hook holds the target species with probability $(\lambda_T / \lambda)(1 - P_b)$.
The expected catch on $n$ hooks is therefore

$$
\operatorname{E}(c) = n \lambda_T \frac{1 - P_b}{-\log P_b},
$$

so with $P_b$ from the baits returned, a Poisson model with an offset of $\log(n (1 - P_b) / (-\log P_b))$ estimates the per-hook catch rate $\lambda_T$.
The fraction is 1 when no baits are removed and falls towards 0 as more are removed.
It is undefined when no baits come back, so we treat those `r sum(d$bait_returned == 0)` sets as if one bait came back.
This is an arbitrary choice, and it affects the sets most affected by competition.

```{r fit-icr}
baited <- pmax(d$bait_returned, 1) / d$hook_count
icr <- ifelse(baited < 1, (1 - baited) / -log(baited), 1)
fit_icr <- update(fit_pois, offset = log(d$hook_count * icr))
```

Unlike censoring, the ICR applies this model of competition to every set, including sets with little bait removal.

## Censored negative binomial models

Instead of an observation-level random intercept, a negative binomial accounts for overdispersion with a dispersion parameter $\phi$.
A negative binomial count is a Poisson count whose rate is gamma distributed, so each set's per-hook catch rate is now gamma instead of lognormal.
The `censored_nbinom2()` and `censored_nbinom1()` families differ in how the variance grows with the mean $\mu$:

- NB2: $\operatorname{Var}(c) = \mu + \mu^2 / \phi$. The gamma catch rate has a constant coefficient of variation, like the lognormal catch rate of the Poisson-lognormal model.
- NB1: $\operatorname{Var}(c) = \mu (1 + \phi)$. The coefficient of variation of the gamma catch rate decreases as the mean increases.

Like the censored Poisson, these have no upper limit on the count, and they use the same `censored_upper` vector:

```{r fit-nb}
fit_nb2_cens <- sdmTMB(
  catch_count ~ 0 + fyear + poly(log_depth, 2),
  offset = log(d$hook_count),
  data = d,
  mesh = mesh,
  family = censored_nbinom2(),
  censored_upper = upr,
  time = "year",
  spatial = "on",
  spatiotemporal = "off"
)
fit_nb1_cens <- update(fit_nb2_cens, family = censored_nbinom1())
sanity(fit_nb2_cens)
sanity(fit_nb1_cens)
```

Each set's per-hook catch rate has mean $\exp(\eta)$, so with an offset of 0 (one hook), the index is the sum of the mean per-hook catch rate over the grid.
No observation-level random intercepts need to be left out, and no other adjustment is needed.

# Comparing indices

We'll predict over the survey grid `hbll_s_grid` and calculate an index for each model.
We set the offset to 0 for prediction (i.e., one hook) and leave out the observation-level random intercepts with `re_form_iid = NA`.

```{r grid}
grid <- replicate_df(hbll_s_grid, "year", unique(d$year))
grid$fyear <- as.factor(grid$year)
grid$log_depth <- log(grid$depth)
```

```{r index}
get_ind <- function(fit, label, ...) {
  ind <- get_index(
    fit, newdata = grid, offset = rep(0, nrow(grid)),
    predict_args = list(re_form_iid = NA),
    bias_correct = TRUE, ...
  )
  ind$model <- label
  ind
}
ind <- rbind(
  get_ind(fit_pois, "Poisson"),
  get_ind(fit_pois_cens, "Censored Poisson"),
  get_ind(fit_pois_cens_bounded, "Censored Poisson (upper bound)"),
  get_ind(fit_icr, "ICR Poisson")
)
```

Since these are relative indices, we'll scale each by its geometric mean before plotting:

```{r plot-index}
scale_index <- function(ind) {
  do.call(rbind, lapply(split(ind, ind$model), function(x) {
    gm <- exp(mean(log(x$est)))
    x[, c("est", "lwr", "upr")] <- x[, c("est", "lwr", "upr")] / gm
    x
  }))
}
plot_index <- function(ind) {
  ggplot(ind, aes(year, est, colour = model, fill = model)) +
    geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.15, colour = NA) +
    geom_line() +
    geom_point() +
    labs(x = "Year", y = "Relative index", colour = NULL, fill = NULL) +
    theme(legend.position = "bottom")
}
ind <- scale_index(ind)
plot_index(ind)
```

The censored models differ most from the uncensored Poisson model in years with an unusually high or low proportion of censored sets.
For example, 2011 had the highest proportion of saturated sets, and its index is higher once we account for hook competition.
2014 had very few saturated sets, and its index is lower relative to the other years.
Here, the upper bound mostly has a modest effect, pulling some years (e.g., 2022) partway back towards the Poisson estimates by limiting how large the censored counts can be.
The ICR index moves in the same direction as the censored indices, but generally by less.

The censored negative binomial indices also move in the same direction:

```{r index-nb}
ind_nb <- rbind(
  get_ind(fit_nb2_cens, "Censored NB2"),
  get_ind(fit_nb1_cens, "Censored NB1")
)
ind_nb <- scale_index(ind_nb)
plot_index(rbind(ind_nb, ind[ind$model == "Censored Poisson", ]))
```

The NB2 index is close to the censored Poisson-lognormal index, since both assume the same relative variability in catch rate at any mean.
The NB1 index changes less between years.
We'll see below that it's very similar to the censored beta-binomial index, and the two models fit about equally well.

# Limiting counts to the number of hooks

A Poisson model has no upper limit on the count, but a set can't catch more fish than it has hooks.
Binomial and beta-binomial models with the number of hooks as the number of trials respect that limit.

## A censored binomial

With a complementary log-log link, the probability that a hook catches the target species is $p = 1 - \exp(-\exp(\eta))$, where $\eta$ is the linear predictor.
This is what we would expect if each hook caught fish as a Poisson process with rate $\lambda = \exp(\eta)$ and could hold at most one fish, since then $p = 1 - \exp(-\lambda)$.
As in the Poisson model, an observation-level random intercept, `(1 | obs_id)`, accounts for overdispersion, so each set's $\lambda$ is lognormal.
The model is therefore a Poisson-lognormal model of the per-hook catch rate with at most one fish per hook.

The `censored_binomial()` family works like `censored_poisson()`, with the number of hooks supplied through `weights`.
The latent count without hook competition still can't exceed the number of hooks, so a right-censored set has a count between the observed count and the number of hooks:

$$
\Pr(c \mid p) = \Pr(c \le Y \le n), \quad Y \sim \operatorname{Binomial}(n, p),
$$

where $n$ is the number of hooks.
This gives every censored set an upper bound, so we can use the same `censored_upper` vector with `Inf` for censored sets and don't need to calculate one.
Counts must be whole numbers, and a non-integer bound means the largest count it allows (e.g., 2.5 means 2).

```{r fit-binom}
fit_binom_cens <- sdmTMB(
  catch_count ~ 0 + fyear + poly(log_depth, 2) + (1 | obs_id),
  weights = d$hook_count,
  data = d,
  mesh = mesh,
  family = censored_binomial(link = "cloglog"),
  censored_upper = upr,
  time = "year",
  spatial = "on",
  spatiotemporal = "off"
)
sanity(fit_binom_cens)
```

The likelihood of each censored set is a sum over every possible count in its interval, so fitting can be slower than for the censored Poisson when sets have many hooks.

As with the censored Poisson, this is a true interval-censoring likelihood only if which sets are censored, and their bounds, carry no extra information about the parameters.
Hook competition doesn't censor counts at random: it reduces catches through a process that also depends on the abundance of the target and other species.
The censored likelihood is therefore best thought of as a working likelihood, and how well it recovers trends depends on how competition actually reduces catches [see the simulations in @watson2023].

For the index, we want the per-hook catch rate $\lambda = \exp(\eta)$, not the catch probability $p$, which can't exceed 1 and so levels off as density increases.
`derived_link = "log"` sums $\exp(\eta)$ over the grid instead of $p$.
Leaving out the observation-level random intercepts with `re_form_iid = NA` gives the median of each lognormal catch rate, which is a constant multiple of its mean, so the relative index is the same as one based on the mean catch rate, as for the Poisson-lognormal model.

## A censored beta-binomial

A beta-binomial accounts for overdispersion differently: each set's catch probability $p$ follows a beta distribution with mean $\bar{p} = 1 - \exp(-\exp(\eta))$ and a dispersion parameter $\phi$, instead of an observation-level random intercept.
The `censored_betabinomial()` family works like `censored_binomial()`:

$$
\Pr(c \mid \bar{p}, \phi) = \Pr(c \le Y \le n), \quad Y \sim \operatorname{BetaBinomial}(n, \bar{p}, \phi).
$$

```{r fit-bb}
fit_bb <- sdmTMB(
  catch_count ~ 0 + fyear + poly(log_depth, 2),
  weights = d$hook_count,
  data = d,
  mesh = mesh,
  family = betabinomial(link = "cloglog"),
  time = "year",
  spatial = "on",
  spatiotemporal = "off"
)
fit_bb_cens <- update(
  fit_bb,
  family = censored_betabinomial(link = "cloglog"),
  censored_upper = upr
)
sanity(fit_bb_cens)
```

To speed up fitting, the censored beta-binomial sums over the counts inside each interval or outside it, whichever is shorter, and checks the precision of each set after fitting.

With `derived_link = "log"`, the index sums $\exp(\eta) = -\log(1 - \bar{p})$.
Because the beta-binomial mixes over the catch probability rather than the catch rate, this isn't proportional to the mean per-hook catch rate in general.
The mean rate is $\operatorname{E}(-\log(1 - p)) = \psi(\phi) - \psi((1 - \bar{p})\phi)$, where $\psi$ is the digamma function, and how much it exceeds $-\log(1 - \bar{p})$ depends on $\bar{p}$ and $\phi$.
We can check how much this matters here by calculating both over the grid:

```{r bb-mean-rate}
p_bar <- 1 - exp(-exp(predict(fit_bb_cens, newdata = grid)$est))
phi <- exp(fit_bb_cens$model$par[["ln_phi"]])
rate_mean <- digamma(phi) - digamma((1 - p_bar) * phi)
rate_eta <- -log(1 - p_bar)
quantile(p_bar, c(0.5, 0.99))
ratio <- tapply(rate_mean, grid$year, sum) / tapply(rate_eta, grid$year, sum)
round(ratio / mean(ratio), 4)
```

Even though $\bar{p}$ is large in some grid cells, the mean rate is close to a constant multiple of $\exp(\eta)$ here, so the relative index is nearly the same either way:

```{r index-binom}
ind_binom <- rbind(
  get_ind(fit_binom_cens, "Censored binomial", derived_link = "log"),
  get_ind(fit_bb_cens, "Censored beta-binomial", derived_link = "log")
)
ind_binom <- scale_index(ind_binom)
plot_index(rbind(ind_binom, ind_nb[ind_nb$model == "Censored NB1", ],
  ind[ind$model == "Censored Poisson", ]))
```

The censored binomial index is very close to the censored Poisson index.
With catch probabilities this low, a limit of one fish per hook rarely matters, and the two models account for overdispersion in the same way.
The censored beta-binomial moves the index in the same direction as the other censored models, e.g., up in 2011 and down in 2014, but by less here, and is very close to the censored NB1 index.
It accounts for overdispersion differently, so differences in its index and uncertainty aren't only due to the upper limit on the count.

The censored likelihoods are probabilities of the same intervals of counts, so their marginal AIC values are comparable (comparisons with the uncensored models, the ICR model, or the bounded censored Poisson are not, because they treat the observations differently):

```{r aic}
AIC(fit_pois_cens, fit_nb2_cens, fit_nb1_cens, fit_binom_cens, fit_bb_cens)
```

We can compare how well simulated data from each model reflect the properties of the observed data:

```{r}
set.seed(1)
s_bb <- simulate(fit_bb_cens, nsim = 300, type = "mle-mvn")
s_binom <- simulate(fit_binom_cens, nsim = 300, type = "mle-mvn")
s_pois_bounded <- simulate(fit_pois_cens_bounded, nsim = 300, type = "mle-mvn")
s_pois <- simulate(fit_pois_cens, nsim = 300, type = "mle-mvn")
s_nb1 <- simulate(fit_nb1_cens, nsim = 300, type = "mle-mvn")
s_nb2 <- simulate(fit_nb2_cens, nsim = 300, type = "mle-mvn")

dharma_residuals(s_bb, object = fit_bb_cens)
dharma_residuals(s_binom, object = fit_binom_cens)
dharma_residuals(s_pois_bounded, object = fit_pois_cens_bounded)
dharma_residuals(s_pois, object = fit_pois_cens)
dharma_residuals(s_nb1, object = fit_nb1_cens)
dharma_residuals(s_nb2, object = fit_nb2_cens)
```

Note how these families have different implications for the maximum number of fish:

```{r}
max(s_bb) # can't exceed the number of hooks
max(s_binom) # can't exceed the number of hooks
max(s_pois_bounded) # not limited: `censored_upper` only affects the likelihood
max(s_pois) # not limited
max(s_nb1) # not limited
max(s_nb2) # not limited; very high
```

# Summary

- Calculate the proportion of baits removed for each set.
- Choose a breakdown point $p^*$, e.g., by plotting a smoother of catch against bait removal.
- Build a `censored_upper` vector: the observed count for uncensored sets and `Inf` (or an upper bound) for sets at or above $p^*$.
- Fit with `family = censored_poisson()` and `censored_upper = ...`, ideally with an observation-level random intercept to account for overdispersion, or with `family = censored_nbinom2()` or `censored_nbinom1()`.
- To limit each count to the number of hooks, fit with `family = censored_binomial(link = "cloglog")`, the number of hooks as `weights`, and an observation-level random intercept, or with `family = censored_betabinomial(link = "cloglog")`.
- For binomial and beta-binomial models, calculate the index with `derived_link = "log"` so it sums the per-hook catch rate rather than the catch probability.

See @watson2023 for the full method, simulation tests, and an application to 11 species.

# References
