---
title: "Index standardization with sdmTMB"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Index standardization with sdmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**If the code in this vignette has not been evaluated, a rendered version is available on the [documentation site](https://sdmTMB.github.io/sdmTMB/index.html) under 'Articles'.**

```{r setup, include = FALSE, cache=FALSE}
dplyr_installed <- require("dplyr", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
pkgs <- dplyr_installed && ggplot_installed
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = EVAL,
  purl = EVAL
)
```

```{r packages, message=FALSE, warning=TRUE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
```

Let's perform index standardization with the built-in data for Pacific Cod.

- The density units should be kg/km^2^.
- Here, X and Y are coordinates in UTM zone 9.

```{r glimpse-pcod}
glimpse(pcod)
```

First we will create our SPDE mesh. We will use a relatively coarse mesh as a balance between speed and accuracy in this vignette (`cutoff = 10`, where cutoff is in the units of X and Y (km here) and represents the minimum distance between points before a new mesh vertex is added). In applied scenarios, you will likely want to use a finer mesh and check that increasing the number of knots does not change the conclusions.

```{r spde, fig.asp=0.8}
pcod_spde <- make_mesh(pcod, c("X", "Y"), cutoff = 10)
plot(pcod_spde)
```

Let's fit a GLMM. Note that if you want to use this model for index standardization, then you will likely want to include `0 + as.factor(year)` or `-1 + as.factor(year)` so that there is a factor predictor representing the mean estimate for each time slice.

```{r model}
m <- sdmTMB(
  data = pcod, 
  formula = density ~ 0 + as.factor(year),
  time = "year", mesh = pcod_spde, family = tweedie(link = "log"))
```

We can inspect randomized quantile residuals:

```{r residuals, fig.width = 8}
pcod$resids <- residuals(m) # randomized quantile residuals
# Also see residuals(..., type = "mle-mcmc") which are better but slower
hist(pcod$resids)
qqnorm(pcod$resids)
abline(a = 0, b = 1)
ggplot(pcod, aes(X, Y, col = resids)) + scale_colour_gradient2() +
  geom_point() + facet_wrap(~year) + coord_fixed()
```

Now we want to predict on a fine-scale grid over the entire survey domain. There is a grid built into the package for Queen Charlotte Sound named `qcs_grid`. Our prediction grid also needs to have all the covariates that we used in the model above.

```{r glimpse-grid}
glimpse(qcs_grid)
```

We can replicate our grid across all years:

```{r replicate-grid}
grid_yrs <- replicate_df(qcs_grid, "year", unique(pcod$year))
```


Now make the predictions on new data.

```{r predictions}
predictions <- predict(m, newdata = grid_yrs, return_tmb_object = TRUE)
```

Let's make a small function to help make maps.

```{r plot-map}
plot_map <- function(dat, column) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    facet_wrap(~year) +
    coord_fixed()
}
```

There are four kinds of predictions that we get out of the model. First, we will show the predictions that incorporate all fixed and random effects:

```{r plot-all-effects, fig.width = 8}
plot_map(predictions$data, exp(est)) +
  scale_fill_viridis_c(trans = "sqrt") +
  ggtitle("Prediction (fixed effects + all random effects)")
```

We can also look at just the fixed effects, here year:

```{r plot-fix-defects, fig.width = 8}
plot_map(predictions$data, exp(est_non_rf)) +
  ggtitle("Prediction (fixed effects only)") +
  scale_fill_viridis_c(trans = "sqrt")
```

We can look at the spatial random effects that represent persistent deviations in space through time that are not accounted for by our fixed effects. In other words, these deviations represent persistent biotic and abiotic factors affecting biomass density that are not accounted for in the model.

```{r plot-spatial-effects, fig.width = 8}
plot_map(predictions$data, omega_s) +
  ggtitle("Spatial random effects only") +
  scale_fill_gradient2()
```

And finally we can look at the spatiotemporal random effects that represent deviations from the fixed-effect predictions and the spatial random-effect deviations. These represent biotic and abiotic factors that change through time and are not accounted for in the model.

```{r plot-spatiotemporal-effects, fig.width = 8}
plot_map(predictions$data, epsilon_st) +
  ggtitle("Spatiotemporal random effects only") +
  scale_fill_gradient2()
```

When we ran our `predict.sdmTMB()` function, it also returned a report from TMB in the output because we included `return_tmb_object = TRUE`. We can then run our `get_index()` function to extract the total biomass estimates and standard errors.

We will need to set the `area` argument to 4 km^2^ since our grid cells are 2 km x 2 km. If some grid cells were not fully in the survey domain (or were on land), we could instead pass a vector of grid-cell areas to the `area` argument matching the number of grid cells.

```{r get-index}
index <- get_index(predictions, area = 4, bias_correct = TRUE)
```

```{r plot-index}
ggplot(index, aes(year, est)) + geom_line() +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.4) +
  xlab('Year') + ylab('Biomass estimate (kg)')
```

These are our biomass estimates:

```{r index-table, results='asis'}
mutate(index, cv = sqrt(exp(se^2) - 1)) |> 
  select(-log_est, -se) |>
  knitr::kable(format = "pandoc", digits = c(0, 0, 0, 0, 2))
```

We can also calculate an index for part of the survey domain. We'll make an index for everything south of UTM 5700 by subsetting our prediction grid. For more complicated spatial polygons, you could intersect the polygon with the prediction grid using something like `sf::st_intersects()`.

```{r index-subset}
qcs_grid_south <- grid_yrs[grid_yrs$Y < 5700, ]
predictions_south <- predict(m, newdata = qcs_grid_south, 
  return_tmb_object = TRUE)
index_south <- get_index(predictions_south, area = 4, bias_correct = TRUE)
head(index_south)
```

We can visually compare the two indices:

```{r south-index-plot}
mutate(index, region = "all") |> 
  bind_rows(mutate(index_south, region = "south")) |> 
  ggplot(aes(year, est)) + 
  geom_line(aes(colour = region)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr, fill = region), alpha = 0.4)
```
