---
title: "Length-at-age from length-stratified age samples"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Length-at-age from length-stratified age samples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_ggplot <- requireNamespace("ggplot2", quietly = TRUE)
has_dplyr <- requireNamespace("dplyr", quietly = TRUE)
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && has_dplyr && has_ggplot
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = EVAL,
  purl = EVAL
)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4)
```

```{r setup, message=FALSE, warning=FALSE}
library(sdmTMB)
library(ggplot2)
library(dplyr)
theme_set(theme_light())
```

Fish surveys usually measure the length of every fish caught but read ages for only a subsample.
A common design is length-stratified age sampling: ages are read for up to a fixed number of fish in each length bin (e.g., 5 fish per 2 cm bin).
Fish in rare length bins, such as unusually small young fish or unusually large old fish, are then much more likely to be aged than fish in common bins.
Fitting length-at-age to the aged fish as if they were a random sample therefore biases length-at-age: too short for young ages and too long for old ages (Perreault et al. 2020).
If abundance changes over time, the sampling fractions change, and so does the bias, which can create spurious trends in length-at-age.

The `length_stratified` argument of `sdmTMB()` corrects for this by conditioning each fish's length on it having been selected for ageing (Candy et al. 2007).
The likelihood of fish $i$'s length $l_i$, given its age $a_i$ and that it was aged, is

$$
f(l_i \mid a_i, \textrm{aged}) =
\frac{\pi_{k(l_i)} f(l_i \mid a_i)}{\sum_k \pi_k \Pr(l \in \textrm{bin } k \mid a_i)},
$$

where $\pi_k$ is the fraction of measured fish that were aged in length bin $k$ of the fish's sampling unit (e.g., a survey set or survey-year).
The denominator is the probability that a fish of this age (and location and time, in a spatiotemporal model) would have been aged.
Because the model is conditional on age, no model of the age distribution is needed.

This implements Candy's conditional-on-age method: sampling fractions are treated as fixed, and bins with no measured fish are excluded from the denominator (sampling fraction zero).
It is not the joint age-length empirical proportion (EP) likelihood.
Excluding empty bins matters when they carry appreciable model probability; even ageing every measured fish does not remove this truncation if empty bins remain.

This requires the RTMB backend (the default) and the `lognormal()` or `gaussian()` family.
The log likelihood is conditional on selection for ageing, so it can't be compared (e.g., with AIC) to that of a model fitted without `length_stratified`.
Residuals, response simulations, and DHARMa diagnostics are currently blocked because they do not account for selection for ageing.
Population predictions remain available with `predict()`.

## Simulating a length-stratified age sample

We simulate a population of fish caught in 5 survey years with lognormal lengths around a von Bertalanffy curve and declining abundance with age.
Every fish is measured, but ages are read for at most 5 fish per year and 2 cm length bin.

```{r simulate}
set.seed(1)
ages <- 1:12
true <- data.frame(age = ages, length = 60 * (1 - exp(-0.25 * (ages + 0.5))))

measured <- lapply(1:5, function(year) {
  age <- sample(ages, 4000, replace = TRUE, prob = exp(-0.35 * ages))
  mu <- true$length[age]
  data.frame(year = year, age = age,
    length = rlnorm(length(age), log(mu) - 0.1^2 / 2, sdlog = 0.1))
}) |> bind_rows()

breaks <- seq(0, 80, by = 2)
measured <- measured |>
  mutate(bin = findInterval(length, breaks)) |>
  group_by(year, bin) |>
  mutate(aged = row_number() %in% sample(n(), min(n(), 5))) |>
  ungroup()

aged <- filter(measured, aged)
nrow(aged)
```

The aged fish over-represent the tails of each age's length distribution:

```{r plot-sample}
ggplot(measured, aes(age, length)) +
  geom_jitter(width = 0.2, alpha = 0.05, size = 0.6) +
  geom_jitter(data = aged, width = 0.2, colour = "red", size = 0.6, alpha = 0.5) +
  labs(x = "Age", y = "Length (cm)",
    title = "All measured fish (grey) and aged fish (red)")
```

## The sampling design

`length_strata()` takes all measured fish, with their lengths and whether each was aged, and tabulates the fraction aged in each sampling unit and length bin.
Here, ages were sampled by year, so the year is the sampling unit.
Every sampling unit in the fitted data must occur in `measured`, and every fitted fish must fall in a bin with aged fish.

```{r design}
design <- length_strata(measured, unit = "year", breaks = breaks,
  length = "length", aged = "aged")
```

## Fitting length-at-age

We fit lognormal length-at-age models to the aged fish, ignoring and accounting for the design.
For comparison, we also fit the model to all fish with their (normally unknown) ages.
Spatial and spatiotemporal fields, smoothers, and other terms can be added as usual; here we keep the model non-spatial.

```{r fit}
fit_ignore <- sdmTMB(length ~ 0 + factor(age), data = aged,
  family = lognormal(), spatial = "off")

fit_lsas <- sdmTMB(length ~ 0 + factor(age), data = aged,
  family = lognormal(), spatial = "off",
  length_stratified = design)

fit_all <- sdmTMB(length ~ 0 + factor(age), data = measured,
  family = lognormal(), spatial = "off")

fit_lsas
```

Predictions from the corrected model are of the population, not of the aged sample:

```{r compare}
nd <- data.frame(age = ages)
fits <- list(
  "Ignoring design" = fit_ignore,
  "Length-stratified" = fit_lsas,
  "All fish aged" = fit_all
)
pred <- lapply(names(fits), function(model) {
  p <- predict(fits[[model]], newdata = nd, re_form = NA, se_fit = TRUE)
  mutate(p, model = model)
}) |> bind_rows()

ggplot(pred, aes(age, exp(est), colour = model)) +
  geom_line(data = true, aes(age, length), colour = "black", lty = 2,
    inherit.aes = FALSE) +
  geom_pointrange(aes(ymin = exp(est - 2 * est_se), ymax = exp(est + 2 * est_se)),
    position = position_dodge(width = 0.5), size = 0.3) +
  labs(x = "Age", y = "Mean length (cm)", colour = NULL,
    caption = "Dashed line: true mean length-at-age")
```

Ignoring the design overestimates length at older ages by several cm and overestimates the variability in length-at-age:

```{r sigma}
sapply(fits, function(fit) tidy(fit, "ran_pars")$estimate)
```

Accounting for the design recovers length-at-age close to the estimates from ageing every fish.

## References

Candy, S.G., Constable, A.J., Lamb, T., and Williams, R. 2007. A von Bertalanffy growth model for toothfish at Heard Island fitted to length-at-age data and compared to observed growth from mark-recapture studies. CCAMLR Science, 14: 43–66.

Perreault, A.M.J., Zheng, N., and Cadigan, N.G. 2020. Estimation of growth parameters based on length-stratified age samples. Canadian Journal of Fisheries and Aquatic Sciences, 77(3): 439–450. <https://doi.org/10.1139/cjfas-2019-0129>
