---
title: "Estimating density-dependent effects on spatial distribution with sdmTMB"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating density-dependent effects on spatial distribution with sdmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**If the code in this vignette has not been evaluated, a rendered version is available on the [documentation site](https://sdmTMB.github.io/sdmTMB/index.html) under 'Articles'.**

```{r setup, include = FALSE, cache=FALSE}
dplyr_installed <- require("dplyr", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
pkgs <- dplyr_installed && ggplot_installed
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = EVAL,
  purl = EVAL
)
```

```{r packages, message=FALSE, warning=TRUE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
```

## Overview

In some applications, the spatial distribution of a species may change as its overall expected abundance changes.
For example, fish may redistribute as density changes.
See Thorson et al. (2022).

`model_index()` lets a spatially varying coefficient (SVC) use a time index that is derived from the fitted model.
The index is estimated jointly with the other model parameters and is centered across time.
This is an association between a model-derived index and spatial redistribution.

There are two useful ways to fit this type of model:

1. Fit an overall time index first, then use the estimated index as an
   ordinary SVC covariate in a second, separate model (a two-step
   workflow).
2. Use `model_index()` to estimate the index and the SVC jointly

The first approach is easier to break into separate diagnostics.
The second propagates uncertainty between the index and the spatially varying coefficient.

## Data and mesh

We will use the built-in dogfish trawl-survey data.
The response is catch weight, and `area_swept` is included as an offset so that the model describes catch rate.

```{r data}
d <- dogfish
d$year_factor <- factor(d$year)

glimpse(d)
sort(unique(d$year))
```

We use the same mesh for all fits.
The relatively coarse mesh keeps this example manageable; in an analysis, check that the conclusions are stable to a finer mesh.

```{r mesh, fig.asp=0.8}
mesh <- make_mesh(d, c("X", "Y"), cutoff = 20)
plot(mesh)
```

## Joint model-derived index

The formula below includes a fixed effect for each year and asks for one SVC
whose covariate is the model-derived time index:

```{r joint-model, warning=FALSE, message=FALSE, results='hide'}
fit_joint <- sdmTMB(
  catch_weight ~ 0 + year_factor, #<
  offset = log(d$area_swept),
  spatial_varying = ~ 0 + model_index(year_factor), #<
  data = d,
  mesh = mesh,
  time = "year",
  family = tweedie(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  silent = FALSE
)
```

```{r}
sanity(fit_joint)
```

The `year_factor` term supplies the overall temporal pattern. It remains in
the model in the usual way; `model_index()` adds an additional interaction
between a centered version of the fitted temporal index and a spatial random
field.

We can inspect the fitted SVC field in the predictions. The long name records
the term used to construct the model-derived index.

```{r joint-predictions}
pred_joint <- predict(fit_joint)
names(pred_joint)
```

The estimated spatially varying coefficient is on the log scale for this
Tweedie log-link model. Positive values indicate that the spatial pattern
becomes relatively stronger when the centered model-derived index is positive;
negative values indicate the opposite association.

```{r joint-svc, fig.asp=0.8}
ggplot(pred_joint, aes(X, Y, colour = `zeta_s_model_index(year_factor)`)) +
  geom_point() +
  coord_fixed() +
  scale_colour_gradient2() +
  labs(colour = "Model-index SVC")
```

## Two-step index and SVC model

To make the two-step version, first fit the same temporal model without an
SVC. The fitted fixed effects and the mean spatiotemporal field at each year
form the uncentered time index.

```{r first-step, warning=FALSE, message=FALSE, results='hide'}
fit_overall <- sdmTMB(
  catch_weight ~ 0 + year_factor,
  offset = log(d$area_swept),
  data = d,
  mesh = mesh,
  time = "year",
  family = tweedie(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  silent = FALSE
)

pars <- get_pars(fit_overall)
index <- data.frame(
  year = sort(unique(d$year)),
  fixed_effect = unname(coef(fit_overall)),
  epsilon_mean = apply(pars$epsilon_st, 2, mean)
)
index <- dplyr::mutate(index, index = fixed_effect + epsilon_mean)
index <- dplyr::mutate(index, index = index - mean(index))
index
```

Now add this estimated index to the data and use it as an ordinary SVC
covariate. This is the explicit two-step analogue of the joint model
above.

```{r second-step, warning=FALSE, message=FALSE, results='hide'}
d_index <- dplyr::left_join(d, index[c("year", "index")], by = "year")

fit_two_step <- sdmTMB(
  catch_weight ~ 0 + year_factor,
  spatial_varying = ~ 0 + index,
  offset = log(d_index$area_swept),
  data = d_index,
  mesh = mesh,
  time = "year",
  family = tweedie(link = "log"),
  spatial = "on",
  spatiotemporal = "iid",
  silent = FALSE
)
```

The two fitted SVC fields should be closely related, although they are not
expected to be identical: the joint model estimates the index and SVC at the
same time, whereas the two-step model treats the first-step index as fixed.

```{r compare-svc, fig.asp=0.8}
pred_two_step <- predict(fit_two_step)
svc_comparison <- data.frame(
  joint = pred_joint$`zeta_s_model_index(year_factor)`,
  two_step = pred_two_step$zeta_s_index
)

ggplot(svc_comparison, aes(two_step, joint)) +
  geom_point(alpha = 0.35) +
  coord_fixed() +
  geom_abline(intercept = 0, slope = 1, colour = "grey40") +
  labs(
    x = "Two-step SVC",
    y = "Joint model SVC",
    subtitle = paste0(
      "Correlation = ",
      round(cor(svc_comparison$two_step, svc_comparison$joint), 3)
    )
  )
```

```{r}
tidy(fit_joint, "ran_pars")
tidy(fit_two_step, "ran_pars")
```

## A joint model with two components

The same syntax can be used with a delta model. Here we use a Poisson-link delta-lognormal model.

We won't evaluate it to save time in this vignette.

```{r joint-delta, warning=FALSE, message=FALSE, eval=FALSE}
fit_joint_delta <- sdmTMB(
  catch_weight ~ 0 + year_factor,
  offset = log(d$area_swept),
  spatial_varying = ~ 0 + model_index(year_factor),
  data = d,
  mesh = mesh,
  time = "year",
  family = delta_lognormal(type = "poisson-link"),
  spatial = "off",
  spatiotemporal = "iid",
  silent = FALSE
)
```

The two linear predictors share the same time index, but each has its own SVC field.

## Interpretation and current limits

The model-derived index is an equal-vertex mean of the spatiotemporal field,
so its numerical value can depend on the mesh. The SVC is a model-based association: it describes how spatial deviations vary with the
model-derived overall index.

Only one direct `model_index()` term is supported. The wrapped term must be
a complete fixed-effect term in every component and must be constant within
each fitted time. An explicit `time` argument and at least two fitted times are
required. Prediction and simulation are supported at fitted times, but
`extra_time` and forward `project()` are not currently supported for these
models.

For the underlying statistical definition and the restrictions on the formula,
see [the `model_index()` reference](https://sdmTMB.github.io/sdmTMB/reference/model_index.html).

## References

Thorson, J. T. (2022). Development and simulation testing for a new approach
to density dependence in species distribution models. *ICES Journal of Marine
Science*, 79, 117--128. <https://doi.org/10.1093/icesjms/fsab247>
