---
title: "Estimating non-local covariate effects with sdmTMB: spatial diffusion and time lags"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: '../refs.bib'
csl: '../mee.csl'
vignette: >
  %\VignetteIndexEntry{Estimating non-local covariate effects with sdmTMB: spatial diffusion and time lags}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**If the code in this vignette has not been evaluated, a rendered version is available on the [documentation site](https://sdmTMB.github.io/sdmTMB/index.html) under 'Articles'.**

```{r setup, include = FALSE, cache=FALSE}
ggplot_installed <- require("ggplot2", quietly = TRUE)
pkgs <- ggplot_installed
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = EVAL,
  purl = EVAL
)
```

```{r packages, message=FALSE, warning=TRUE}
library(ggplot2)
library(sdmTMB)
theme_set(theme_light())
```

This vignette demonstrates non-local-covariate models using simulated data. These models could alternatively be described as "distributed lag" models or, in the case of spatial diffusion, "covariate diffusion" models.

When might you want to fit these models? Sometimes, the spatial or temporal scale at which covariates influence the response is unclear.

For example, fish move.
Therefore, if we measure temperature at the location where a fish was sampled, it may be unclear whether the model should use the temperature at the sampling site or a smoothed version of the temperature surface that accounts for fish movement and integration over a broader spatial range.

One option is to fit models using covariates with different levels of smoothing and compare them.
Alternatively, we can estimate the appropriate degree of smoothing simultaneously while we fit the model.
This is what the covariate-diffusion functionality does.

Users are encouraged to read and cite @lindmark2025 for spatial models and @thorson2026 for extensions to temporal and spatiotemporal settings, including the introduction of the MSD and RMSD terms.

We begin with a simple spatial-diffusion example and then fit an example that combines `diffusion()` and `time_lag()` terms.

## Simple spatial diffusion example

We will start by simulating data with a fine-scale predictor (`x1`) and a spatially diffused effect:

```{r simulate-space}
set.seed(1)
n_sites <- 220
site <- data.frame(X = runif(n_sites), Y = runif(n_sites))
dat <- site
mesh <- make_mesh(dat, c("X", "Y"), cutoff = 0.05)
nonlocal_space <- data.frame(
  X = mesh$mesh$loc[, 1],
  Y = mesh$mesh$loc[, 2]
)
nonlocal_space$x1 <- as.numeric(scale(
  sin(6 * pi * nonlocal_space$X) +
    cos(6 * pi * nonlocal_space$Y) +
    rnorm(nrow(nonlocal_space), sd = 1)
))
dat$x1 <- as.numeric(mesh$A_st %*% nonlocal_space$x1)

sim_space <- simulate_new(
  formula = ~ 1,
  data = dat,
  mesh = mesh,
  family = gaussian(),
  spatial = "off",
  spatiotemporal = "off",
  range = 0.2,
  sigma_O = 0,
  phi = 0.1,
  B = c(0, 0.8),
  nonlocal_formula = ~ diffusion(x1),
  nonlocal_data = nonlocal_space,
  lags_kappaS = 1.5,
  seed = 1
)

dat$observed <- sim_space$observed
dat$x1_truth <- sim_space$nl_truth_diffusion_x1
head(dat)
```

Fit the matching covariate-diffusion model:

```{r fit-space}
fit <- sdmTMB(
  observed ~ 1,
  mesh = mesh,
  nonlocal_formula = ~ diffusion(x1), #<
  nonlocal_data = nonlocal_space,
  spatial = "off",
  data = dat
)

fit
tidy(fit)
tidy(fit, effects = "ran_pars")
```

Here `kappaS_nl[x1]` is the estimated spatial diffusion scale parameter for `x1` (larger values imply less smoothing).

The output also reports `MSD[x1]` and `RMSD[x1]`, the implied mean squared displacement (`4 / kappaS_nl[x1]^2`) and root mean squared displacement (`2 / kappaS_nl[x1]`).
`RMSD[x1]` is a typical diffusion distance (a blur radius) in coordinate units.
Users applying this spatial diffusion method should cite @lindmark2025.
Users using these MSD and RMSD metrics should consult and cite @thorson2026.

Compare true and estimated diffused covariates:

```{r compare-space}
pred <- predict(fit)
pred$X <- dat$X
pred$Y <- dat$Y

cor(pred$nl_diffusion_x1, pred$x1_truth)

g1 <- ggplot(pred, aes(X, Y, colour = x1)) +
  geom_point(size = 0.7) +
  scale_colour_gradient2() +
  coord_equal() +
  ggtitle("Fine-scale observed predictor (x1)")

g2 <- ggplot(pred, aes(X, Y, colour = x1_truth)) +
  geom_point(size = 0.7) +
  scale_colour_gradient2() +
  coord_equal() +
  ggtitle("True diffused covariate (from simulation)")

g3 <- ggplot(pred, aes(X, Y, colour = nl_diffusion_x1)) +
  geom_point(size = 0.7) +
  scale_colour_gradient2() +
  coord_equal() +
  ggtitle("Estimated diffused covariate")

print(g1)
print(g2)
print(g3)
```

We can visualize the diffusion operator and the original and smoothed covariate on the mesh:

```{r}
plot_nonlocal_kernel(fit, component = "diffusion")
plot_nonlocal_covariate(fit, component = "diffusion")
```

## Space-time example with additive terms

Next we will simulate and fit a model containing spatial and temporal components: `~ diffusion(x1) + time_lag(x1)`.

```{r fit-space-time-all}
set.seed(1)
n_t <- 16
n_sites <- 300
site_st <- data.frame(X = runif(n_sites), Y = runif(n_sites))
dat_st <- data.frame(
  X = rep(site_st$X, times = n_t),
  Y = rep(site_st$Y, times = n_t),
  year = rep(seq_len(n_t), each = n_sites)
)
mesh_st <- make_mesh(dat_st, c("X", "Y"), cutoff = 0.07)
nonlocal_st <- expand.grid(
  vertex = seq_len(nrow(mesh_st$mesh$loc)),
  year = seq_len(n_t)
)
nonlocal_st$X <- mesh_st$mesh$loc[nonlocal_st$vertex, 1]
nonlocal_st$Y <- mesh_st$mesh$loc[nonlocal_st$vertex, 2]
nonlocal_st$x1 <- as.numeric(scale(
  sin(2 * pi * (nonlocal_st$X + nonlocal_st$year / 8)) +
    cos(2 * pi * (nonlocal_st$Y - nonlocal_st$year / 10)) +
    0.6 * sin(4 * pi * nonlocal_st$X) * cos(nonlocal_st$year / 3) +
    rnorm(nrow(nonlocal_st), sd = 0.15)
))
nonlocal_st$vertex <- NULL
dat_st$x1 <- NA_real_
for (tt in seq_len(n_t)) {
  rows <- dat_st$year == tt
  vertex_values <- nonlocal_st$x1[nonlocal_st$year == tt]
  dat_st$x1[rows] <- as.numeric(mesh_st$A_st[rows, ] %*% vertex_values)
}

sim_st <- simulate_new(
  formula = ~ 1,
  data = dat_st,
  mesh = mesh_st,
  time = "year",
  family = gaussian(),
  spatial = "on",
  spatiotemporal = "off",
  range = 0.3,
  sigma_O = 0.2,
  phi = 0.1,
  B = c(0, 0.7, 0.6),
  nonlocal_formula = ~ diffusion(x1) + time_lag(x1),
  nonlocal_data = nonlocal_st,
  lags_kappaS = 4.4,
  lags_rhoT = 0.3,
  seed = 123
)

dat_st$observed <- sim_st$observed
```

```{r fit-st}
fit_st <- sdmTMB(
  observed ~ 1,
  mesh = mesh_st,
  time = "year",
  nonlocal_formula = ~ diffusion(x1) + time_lag(x1), #<
  nonlocal_data = nonlocal_st,
  spatial = "on",
  spatiotemporal = "off",
  data = dat_st
)

sanity(fit_st)
fit_st

tidy(fit_st, effects = "ran_pars")
```

In the random-effect parameters, `kappaS_nl[x1]` controls spatial diffusion and `kappaT_nl[x1]` controls temporal lags.
Users applying the time-lag functionality should read and cite @thorson2026.

## Supplying a separate covariate grid

By default, the covariate field used for diffusion is built from `data` alone.
If the covariate is available on a finer or more complete grid than the observation locations, we can supply it via `nonlocal_data`.
This also lets us define the covariate at a forecast (`extra_time`) slice without
padding `data` with extra rows.

```{r fit-space-time-grid}
extra_yr <- n_t + 1
grid_st <- expand.grid(
  vertex = seq_len(nrow(mesh_st$mesh$loc)),
  year = seq_len(extra_yr)
)
grid_st$X <- mesh_st$mesh$loc[grid_st$vertex, 1]
grid_st$Y <- mesh_st$mesh$loc[grid_st$vertex, 2]
grid_st$x1 <- as.numeric(scale(
  sin(2 * pi * (grid_st$X + grid_st$year / 8)) +
    cos(2 * pi * (grid_st$Y - grid_st$year / 10)) +
    0.6 * sin(4 * pi * grid_st$X) * cos(grid_st$year / 3)
))
grid_st$vertex <- NULL

fit_st_grid <- sdmTMB(
  observed ~ 1,
  mesh = mesh_st,
  time = "year",
  nonlocal_formula = ~ diffusion(x1) + time_lag(x1), #<
  nonlocal_data = grid_st, #<
  extra_time = extra_yr, #<
  spatial = "on",
  spatiotemporal = "off",
  data = dat_st
)
sanity(fit_st_grid)
```

We will plot the spatially diffused covariate on the non-local data grid locations:

```{r predict-space-time-grid}
pred_grid <- predict(fit_st_grid, newdata = grid_st)

ggplot(pred_grid, aes(X, Y, fill = nl_diffusion_x1)) +
  geom_point(shape = 21) +
  facet_wrap(~year) +
  scale_fill_gradient2()
```

## Plot diffusion kernels

We can separately visualize the spatial diffusion, time diffusion, or their combined effects as either the diffusion kernel or the smoothed covariate field.
By default, the diffusion kernel is shown at the mesh vertices:

```{r plot-diffusion}
plot_nonlocal_kernel(fit_st, component = "diffusion")
plot_nonlocal_kernel(fit_st, component = "time_lag", common_scale = TRUE)
plot_nonlocal_kernel(fit_st, component = "combined")
plot_nonlocal_kernel(fit_st, component = "combined", common_scale = TRUE)
```

In the last case, the colour scale makes it hard to see the variation.
We could add our own ggplot colour scale transformation:

```{r plot-diffusion-scale}
plot_nonlocal_kernel(fit_st, component = "combined", common_scale = TRUE) +
  scale_colour_distiller(trans = "log10", palette = "Blues", direction = 1)
```

All the knots except one in the "original" panel were zero, so they become -Inf when log transformed, but now we can see the space and time nonlocal effects.

```{r}
plot_nonlocal_covariate(fit_st, component = "diffusion")
plot_nonlocal_covariate(fit_st, component = "time_lag", n_steps = 2)
plot_nonlocal_covariate(fit_st, component = "combined", n_steps = 2)
```

We could also have plotted the spatially diffused covariate at the original locations from the predictions.

```{r}
pred <- predict(fit_st)

ggplot(pred, aes(X, Y, colour = x1)) + geom_point() +
  scale_colour_gradient2()

ggplot(pred, aes(X, Y, colour = nl_diffusion_x1)) + geom_point() +
  scale_colour_gradient2()
```

## `predict()` vs. the non-local plotting functions

These two paths answer different questions, so pick based on what you need:

- Use `predict()` and the `nl_*` columns for the modeled effect.
  Columns such as `nl_diffusion_x1` and `nl_time_lag_x1` are the transformed covariate values that enter the linear predictor at your data or `newdata` locations.
  For `time_lag()` these accumulate the response across all prior time slices, and for `diffusion()` they give the smoothed field integrated over space.
  This is what you want for reporting, mapping the estimated effect, correlating against a known truth, or any downstream calculation.
  One column is returned per fitted `nonlocal_formula` term.

- Use `plot_nonlocal_kernel()` and `plot_nonlocal_covariate()` for diagnostics.
  These visualize how the estimated operator behaves.
  `plot_nonlocal_kernel()` shows the impulse response (how a point input spreads in space and/or decays through time) and cannot be reconstructed from the `nl_*` columns.
  `plot_nonlocal_covariate()` isolates a single input time slice and shows how it propagates forward over `n_steps`; it also offers `component = "combined"` (the joint space-and-time response), which is not returned as a `predict()` column.
  Both plot on the mesh vertices by default, or at supplied `newdata` coordinates as points or a raster.

For the pure spatial `diffusion()` case at a single time slice, `plot_nonlocal_covariate()` and mapping `nl_diffusion_x1` from `predict()` show the same field, so either is fine.
The plotting functions become most useful for temporal (`time_lag()`) and `combined` terms, and whenever you want to inspect the operator on the mesh itself.

# References
