---
title: "Fitting a joint preferential-sampling model with sdmTMB"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting a joint preferential-sampling model with sdmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`sdmTMB` is adding a preferential sampling model motivated by

Ward, E.J., K.L. Oken, Y.A. Chen, K.E. Richerson, C. Barnes, and D. Holland 2026. Correcting trend bias from preferential sampling in fishery-dependent indices. In review

```{r setup, include = FALSE, cache=FALSE}
dplyr_installed <- require("dplyr", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
spatstat.data_installed <- require("spatstat.data", quietly = TRUE)
pkgs <- dplyr_installed && ggplot_installed && spatstat.data_installed
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = identical(Sys.getenv("NOT_CRAN"), "true") && pkgs,
  purl = identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
)
```

**This feature is under development.** The previous experimental interface
(the `preferential_*` arguments to `sdmTMBcontrol()`) has been removed. Its
replacement is being implemented for the RTMB backend only. The joint
likelihood can now be fitted for single log-link families and conventional
logit/log delta families. `print()`, `tidy(fit, model = "sampling")`, and
`predict_sampling()` report the sampling model; see `?preferential_sampling`
for which other post-fit methods are supported. This article will be
rewritten with a worked example.

The model couples the usual catch model to a Bernoulli model for
whether each eligible cell of a sampling frame was sampled in each time step:

```
logit(Pr(sampled)) = Z * gamma + b * log(standardized expected catch)
                     [+ optional sampling-only spatial field]
```

`Z * gamma` is a separate sampling regression, for example free annual
intercepts plus distance to port. The standardized expected catch is the
main model evaluated on the sampling frame, like prediction `newdata`. The
covariate values in the frame define the standardization, so catchability
covariates such as gear should be set to a reference value there. The
catch model's spatial and spatiotemporal fields are shared, and `b` is a
constant preference coefficient.

```{r spec, eval = FALSE}
# One row per eligible cell and year, including known unsampled cells (0).
# Unknown sampling history is NA. Main-model covariates in `grid` define the
# standardized catch surface.
grid$gear <- factor("reference", levels = levels(catches$gear))

fit <- sdmTMB(
  catch ~ depth_scaled + gear + (1 | vessel),
  data = catches,
  mesh = mesh,
  time = "year",
  family = tweedie(link = "log"),
  offset = log(catches$effort),
  preferential = preferential_sampling(
    sampled ~ 0 + factor(year) + distance_to_port,
    data = grid,
    re_form_iid = NA, # exclude vessel effects from the shared surface
    offset = 0, # unit effort in the shared surface
    spatial = "off"
  ),
  control = sdmTMBcontrol(backend = "rtmb")
)
```
