---
title: "Fitting a joint preferential-sampling model with sdmTMB"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting a joint preferential-sampling model with sdmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

`sdmTMB` implements a preferential sampling model described in 

Ward, E.J., K.L. Oken, Y.A. Chen, K.E. Richerson, C. Barnes, and D. Holland 2026. Correcting trend bias from preferential sampling in fishery-dependent indices. In review

```{r setup, include = FALSE, cache=FALSE}
dplyr_installed <- require("dplyr", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
spatstat.data_installed <- require("spatstat.data", quietly = TRUE)
pkgs <- dplyr_installed && ggplot_installed && spatstat.data_installed
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = identical(Sys.getenv("NOT_CRAN"), "true") && pkgs,
  purl = identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
)
```

```{r packages, message=FALSE, warning=FALSE}
library(sdmTMB)
library(ggplot2)
theme_set(theme_light())
```

Preferential sampling can occur when the locations at which data are
collected are not independent. For example, fishing
vessels that preferentially return to sites with historically high catch
rates. Ignoring this can bias estimates of the underlying spatial process.

sdmTMB can fit a joint model that couples the usual catch/observation model
to a second, Bernoulli sub-model for whether each cell of a spatial grid was
sampled in a given time slice. The sampling log-odds partly track the main
model's own fitted log-density surface (a user-selected subset of the
fixed effects, plus the spatial and spatiotemporal fields), reprojected onto
the grid, plus an independent nuisance spatial field capturing
sampling-location structure the density surface doesn't explain:

```
logit(Pr(sampled)) = gamma_0 + b_pref * grid_eta + xi_grid
```

`b_pref` governs how strongly sampling tracks the modeled density and can be
constant across time, a random walk, or IID by time slice via
`preferential_b_type`.

Four arguments turn this on, all supplied through `sdmTMBcontrol()`:

* `preferential_response`: the name of a 0/1 (or logical) column indicating
  whether a grid cell/row was sampled. Supplying this is what turns the
  feature on.
* `preferential_grid`: a data frame of grid cells, pre-expanded across every
  fitted time slice (one row per grid cell per time slice), containing the
  mesh coordinates, the `preferential_response` column, and whatever
  predictor column(s) `preferential_formula` needs. Defaults to `data`
  itself if left as `NULL` (only sensible if `data` already has that shape).
* `preferential_formula`: an optional one-sided formula (e.g. `~1` or
  `~ 0 + as.factor(year)`) selecting which term(s) of the main model's
  *already-fitted* fixed effects are reprojected onto the grid. Every term
  must already appear in `formula` -- the sampling sub-model reuses the main
  model's own fitted coefficients rather than estimating new ones. Defaults
  to `NULL`, treated as `~1` (the fitted intercept only).
* `preferential_b_type`: `"constant"` (one shared value), `"rw"` (a random
  walk), or `"iid"` (independent by time slice).

## Example 1: a constant sampling-preference coefficient

We simulate catch data and a spatial grid of sampling indicators. To give
the sampling sub-model's own nuisance field (`xi_s`) a genuine spatial
signal to find, we simulate it with `simulate_new()` rather than as pure
noise:

```{r sim-constant}
set.seed(42)
years <- 1:40
n <- 1500
dat <- data.frame(
  x = runif(n, 0, 10),
  y = runif(n, 0, 10),
  year = rep(years, length.out = n)
)
dat$catch <- rpois(n, exp(1 + 0.15 * dat$y))
mesh <- make_mesh(dat, xy_cols = c("x", "y"), cutoff = 2.5)

grid_xy <- expand.grid(
  x = seq(0.5, 9.5, length.out = 7),
  y = seq(0.5, 9.5, length.out = 7)
)
mesh_grid <- make_mesh(grid_xy, xy_cols = c("x", "y"), cutoff = 1.4)
sim_xi <- simulate_new(
  formula = ~1, data = grid_xy, mesh = mesh_grid, family = gaussian(),
  range = 3, sigma_O = 0.8, phi = 0.01, B = 0, seed = 99
)
xi_true <- sim_xi$omega_s

pref_grid <- do.call(rbind, lapply(years, function(yr) {
  g <- grid_xy
  g$year <- yr
  g$sampled <- rbinom(nrow(g), 1, plogis(-1 + 0.3 * g$y + xi_true))
  g
}))
```

A single year's sampling indicators illustrate the preference for higher `y`
plus the added spatial structure from `xi_true`:

```{r plot-grid}
ggplot(pref_grid[pref_grid$year == 1, ], aes(x, y, colour = factor(sampled))) +
  geom_point(size = 4) +
  labs(colour = "Sampled")
```

Now fit the joint model with `preferential_b_type = "constant"`:

```{r fit-constant}
fit <- sdmTMB(
  catch ~ 1,
  data = dat, mesh = mesh, time = "year",
  family = poisson(),
  control = sdmTMBcontrol(
    preferential_grid = pref_grid,
    preferential_response = "sampled",
    preferential_b_type = "constant"
  )
)
fit
```

The preferential-sampling parameters (`gamma_0`, `b_pref`, and the nuisance
field's `range_xi`/`sigma_xi`) get reported alongside the usual random-field
parameters under `tidy(fit, "ran_pars")`:

```{r tidy-constant}
tidy(fit, "ran_pars")
```

## Example 2: a time-varying sampling-preference coefficient

`preferential_b_type = "rw"` (or `"iid"`) lets `b_pref` vary by time slice,
useful if sampling preference itself has drifted over time (e.g., a fishery
gradually shifting toward historically productive grounds). Everything else
is unchanged:

```{r fit-rw}
fit_rw <- sdmTMB(
  catch ~ 1,
  data = dat, mesh = mesh, time = "year",
  family = poisson(),
  control = sdmTMBcontrol(
    preferential_grid = pref_grid,
    preferential_response = "sampled",
    preferential_b_type = "rw"
  )
)
tidy(fit_rw, "ran_pars")
```

`b_pref` now has one row per time slice. Note that a `"rw"`/`"iid"` `b_pref`
adds its own temporal-variance parameter on top of the nuisance field's
spatial variance; if the data don't have much real year-to-year signal in
the sampling-preference relationship (as in this simulated example), that
variance parameter can sit near a zero-variance boundary, the same way any
random effect does when fit to a near-flat process. This shows up as `NaN`
standard errors from the delta method rather than as a fitting failure.

As with any sdmTMB model, `simulate()` works on the fitted object:

```{r simulate-rw}
sims <- simulate(fit_rw, nsim = 3, seed = 1)
dim(sims)
```
