---
title: "Fitting a joint preferential-sampling model with sdmTMB"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting a joint preferential-sampling model with sdmTMB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, cache=FALSE}
dplyr_installed <- require("dplyr", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
rtmb_installed <- requireNamespace("RTMB", quietly = TRUE)
pkgs <- dplyr_installed && ggplot_installed && rtmb_installed
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = identical(Sys.getenv("NOT_CRAN"), "true") && pkgs,
  purl = identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
)
```

**This feature is experimental.** It requires the RTMB backend, and its
interface and supported models may change.

Fishery-dependent data are rarely collected at random locations. Fishers
return to places where they expect good catches, so where samples are taken
depends on the quantity being measured. An ordinary species distribution
model treats the sample locations as uninformative. When locations are
chosen preferentially, the model's fields are poorly informed in the places
fishers avoid, and an index of abundance can be biased. This is especially
true if the amount or spread of fishing changes over time.

sdmTMB can fit a joint model of the catches and of *where sampling
happened*, motivated by

Ward, E.J., K.L. Oken, Y.A. Chen, K.E. Richerson, C. Barnes, and D. Holland
2026. Correcting trend bias from preferential sampling in fishery-dependent
indices. In review.

The equations are in the preferential-sampling section of the
[model description](https://sdmTMB.github.io/sdmTMB/articles/model-description.html).

## The model

The catch model is an ordinary sdmTMB model with spatial and/or
spatiotemporal fields. The sampling model is a Bernoulli model for whether
each eligible grid cell $g$ was sampled at least once in time step $t$:

$$
\mathrm{logit}\,\Pr(R_{g,t} = 1) = \mathbf{Z}_{g,t} \boldsymbol{\gamma}
  + b\, h_{g,t} + \xi_g.
$$

* $\mathbf{Z}_{g,t} \boldsymbol{\gamma}$ is a separate sampling regression,
  for example free annual intercepts (changing effort) and distance to port.
* $h_{g,t}$ is the log of **standardized expected catch**: the catch model
  evaluated on the cell, including its spatial and spatiotemporal fields.
  You choose the standardization (e.g., a reference gear) through the
  covariate values you supply for each cell.
* $b$ is the preference coefficient. Multiplying the standardized expected
  catch by $c$ multiplies the odds of sampling by $c^b$, holding the other
  sampling terms fixed. $b > 0$ means fishers favor high catches. There is
  no sign restriction.
* $\xi_g$ is an optional sampling-only spatial field. It absorbs spatial
  structure in sampling that isn't explained by the covariates or by $h$.

The two models share the catch fields, and the joint likelihood integrates
over them. The catch fields therefore learn from where fishing did and did
not happen, and $b$ is estimated together with the catch model.

## Simulating a fishery

We simulate one coupled data set that follows this model. Then we can
compare the estimates with the truth. Fish density depends on depth and on
spatial and spatiotemporal fields, and it declines over eight years. The
fleet samples more cells as the years go on, favors cells close to port,
and favors cells with high expected catch ($b = 1$). An area in the
southeast is closed to fishing.

```{r packages, message=FALSE}
library(sdmTMB)
library(dplyr)
library(ggplot2)
```

```{r simulate-biology}
set.seed(42)
years <- 2011:2018
cells <- expand.grid(X = seq(0.2, 9.8, 0.4), Y = seq(0.2, 9.8, 0.4))
cells$depth <- 20 + 15 * cells$X + 20 * sin(cells$Y / 2)
cells$dist_port <- sqrt(cells$X^2 + (cells$Y - 5)^2) / 10
cells$eligible <- !(cells$X > 7 & cells$Y < 3) # a closed area
grid <- merge(cells, data.frame(year = years))

mesh_grid <- make_mesh(grid, c("X", "Y"), cutoff = 0.8)
fields <- sdmTMB_simulate(~1,
  data = grid, mesh = mesh_grid, time = "year",
  family = poisson(), range = 4, sigma_O = 0.8, sigma_E = 0.3, B = 0,
  spatiotemporal = "iid", seed = 1
)
# True log standardized expected catch: a declining trend, a depth
# optimum, and the fields.
grid$h <- log(4) - 0.08 * (grid$year - 2011) -
  0.0006 * (grid$depth - 80)^2 + fields$eta
```

Next we simulate where sampling happened, in the eligible cells only.

```{r simulate-sampling}
frame <- filter(grid, eligible)
a <- seq(-2.6, -1.0, length.out = length(years)) # rising effort
frame$sampled <- rbinom(
  nrow(frame), 1,
  plogis(a[frame$year - 2010] - 1.2 * frame$dist_port +
    1 * (frame$h - mean(frame$h)))
)
group_by(frame, year) |> summarise(prop_sampled = mean(sampled))
```

Catches exist only at sampled cells. Each catch comes from one of two gears
and one of ten vessels, which affect catchability but not fish density.
Gear `"trawl_b"` catches more, and it becomes more common over time. Catch
is a count with negative binomial variation, and effort varies.

```{r simulate-catches}
catches <- frame[frame$sampled == 1, c("X", "Y", "year", "depth", "h")]
n <- nrow(catches)
share_b <- seq(0.2, 0.7, length.out = length(years))[catches$year - 2010]
catches$gear <- factor(ifelse(runif(n) < share_b, "trawl_b", "trawl_a"))
vessel_effect <- rnorm(10, 0, 0.3)
catches$vessel <- factor(sample(paste0("v", 1:10), n, TRUE))
catches$effort <- runif(n, 0.5, 2)
mu <- catches$effort * exp(catches$h + 0.5 * (catches$gear == "trawl_b") +
  vessel_effect[as.integer(catches$vessel)])
catches$catch <- rnbinom(n, size = 2, mu = mu)
```

## Building the sampling frame

The sampling `data` table (the *frame*) is the most important input. It has
one row per eligible cell and time step, and it needs:

* **The sampling response**, coded as 1 (sampled), 0 (eligible, not
  sampled), or `NA` (unknown).
    * A **known unsampled** cell is a 0. It is evidence that fishers chose
      not to go there.
    * An **unknown** cell is `NA`: for example, a year or region without
      reliable logbook coverage. `NA` rows add nothing to the sampling
      likelihood, and the biological time steps are kept.
    * An **ineligible** cell (land, a closed area, outside the fishery's
      range) must be left out of the frame. A 0 there would wrongly suggest
      that fishers avoided it because of low catches.
* **The mesh coordinates** (cell centres) and the model's **`time`
  column**. In real data, assign each catch record to its cell before
  building the indicators. There must be one row per cell and time step.
  The frame is not checked for completeness, so every eligible
  cell-time must be included.
* **The sampling covariates**, here distance to port.
* **Every covariate in the main model's fixed effects and smoothers.**

Suppose logbook records for the north half of the domain were lost in 2014.
Those cells are unknown, not unsampled:

```{r frame-unknown}
frame$sampled[frame$year == 2014 & frame$Y > 5] <- NA
```

### Defining the standardization

The main model is evaluated on the frame like `newdata` in
`predict.sdmTMB()`, so the frame's covariate values define what
"standardized expected catch" means. Covariates that describe biology
(depth here) keep their cell values. Covariates that describe catchability
are set to a reference value:

* **Gear**: a factor column set to the reference gear, with the fitted
  factor levels.
* **Vessel intercepts** (`(1 | vessel)`): `re_form_iid = NA` (the default)
  leaves them out, i.e., sets them to 0, so no `vessel` column is needed.
  `re_form_iid = NULL` would include the fitted intercepts at the vessel
  levels in the frame. Grouping columns must then be factors with levels
  from the fitted data.
* **Effort**: the observation offset is not copied. The `offset` argument
  of `preferential_sampling()` sets the surface's offset on the link scale.
  The default is 0, i.e., unit effort.

```{r frame-standardize}
frame$gear <- factor("trawl_a", levels = levels(catches$gear))
```

Standardization is a scientific choice that the package can't make for
you: a smoother or a random effect could describe biology or catchability.
Holding a covariate at a reference value is not the same as removing its
term, and an interaction is evaluated at the reference value too.

## Fitting the models

First, the ordinary catch-only model:

```{r fit-catch}
mesh <- make_mesh(catches, c("X", "Y"), mesh = mesh_grid$mesh)
fit_catch <- sdmTMB(
  catch ~ 0 + factor(year) + s(depth) + gear + (1 | vessel),
  data = catches, mesh = mesh, time = "year", family = nbinom2(),
  offset = log(catches$effort),
  control = sdmTMBcontrol(backend = "rtmb")
)
```

The joint model adds `preferential`. We use free annual sampling intercepts,
which let the proportion of cells sampled change from year to year without
forcing $b$ or abundance to explain it.

```{r fit-joint}
fit <- update(fit_catch,
  preferential = preferential_sampling(
    sampled ~ 0 + factor(year) + dist_port,
    data = frame
  )
)
```

## Inspecting the fit

`print()` adds a section for the sampling model. Note that the objective,
and so `logLik()` and `AIC()`, include the sampling likelihood.

```{r print}
fit
sanity(fit)
```

`tidy(fit, model = "sampling")` gives the sampling coefficients and $b$
(`b_pref`). The default `tidy()` output describes the catch model.

```{r tidy}
tidy(fit, model = "sampling")
```

The true preference coefficient, 1, is inside the 95% confidence interval
for `b_pref`.
`predict_sampling()` returns the fitted sampling probability for each frame
row, and its pieces on the logit scale: the sampling formula (`est_fixed`),
the preference term (`est_preference`, $b h$), and the shared target
`est_target` ($h$).

```{r predict-sampling, fig.asp = 0.5}
ps <- predict_sampling(fit)
ggplot(filter(ps, year %in% c(2011, 2018)), aes(X, Y, fill = est)) +
  geom_raster() +
  geom_point(data = filter(ps, year %in% c(2011, 2018), sampled == 1),
    inherit.aes = FALSE, aes(X, Y), size = 0.4) +
  facet_wrap(~year) +
  scale_fill_viridis_c() +
  coord_fixed() +
  labs(fill = "Pr(sampled)")
```

With `nsim > 0`, `predict_sampling()` returns draws from the joint
precision matrix. Use them for uncertainty:

```{r predict-sampling-draws}
draws <- predict_sampling(fit, nsim = 200)
ps$lwr <- apply(draws, 1, quantile, probs = 0.025)
ps$upr <- apply(draws, 1, quantile, probs = 0.975)
head(ps[, c("X", "Y", "year", "sampled", "est", "lwr", "upr")])
```

## The standardized index

Catch predictions and indices work as usual. The prediction grid can differ
from the sampling frame: here it includes the closed area. Prediction rows
are not sampling observations. The index uncertainty comes from the joint
model. Keep the default `bias_correct = TRUE`: in our simulations, the
plug-in index (`bias_correct = FALSE`) was biased low for joint delta
models.

```{r index, message=FALSE}
nd <- grid
nd$gear <- factor("trawl_a", levels = levels(catches$gear))
nd$vessel <- factor("v1", levels = levels(catches$vessel))
index_args <- list(re_form_iid = NA)
ind <- bind_rows(
  get_index(fit_catch, newdata = nd, predict_args = index_args) |>
    mutate(model = "Catch only"),
  get_index(fit, newdata = nd, predict_args = index_args) |>
    mutate(model = "Joint")
)
truth <- group_by(grid, year) |> summarise(est = sum(exp(h)))

ggplot(ind, aes(year, est, ymin = lwr, ymax = upr, colour = model)) +
  geom_pointrange(position = position_dodge(width = 0.4)) +
  geom_line(data = truth, aes(year, est), inherit.aes = FALSE) +
  labs(y = "Index", colour = NULL,
    caption = "Line: true standardized index")
```

In this realization, the catch-only model overestimates abundance, because
it fills unsampled, lower-density cells with values that are too high. The
joint model is closer to the truth. This is one simulated data set. The
replicated simulations in the package's
[development files](https://github.com/sdmTMB/sdmTMB/tree/main/tests/optional/preferential-recovery)
include cases where the joint model helps little (e.g., $b = 0$) and cases
where a misspecified sampling model biases it.

## Is a sampling-only field warranted?

Unexplained spatial structure in sampling, such as access or tradition, can
be mistaken for preference if it happens to line up with the catch fields.
Measured access covariates, like distance to port, are the first defense.
A sampling-only field is the second: `spatial = "on"`. The joint models
share the same catch data and sampling frame, so their AIC values are
comparable with each other (but not with the catch-only model).

```{r fit-xi, warning=FALSE}
fit_xi <- update(fit,
  preferential = preferential_sampling(
    sampled ~ 0 + factor(year) + dist_port,
    data = frame, spatial = "on"
  )
)
sanity(fit_xi)
tidy(fit_xi, "ran_pars", model = "sampling")
tidy(fit_xi, model = "sampling") |> filter(term == "b_pref")
AIC(fit, fit_xi)
```

Our simulation has no sampling-only field, and the fit shows it: the field
SD collapses toward zero, the Hessian isn't positive definite, and AIC
favors the simpler model. The estimate of `b_pref` barely changes. When
sampling has a real residual spatial pattern, the field SD is clearly
positive, and you should report how much `b_pref` and the index change
with and without it.

## Delta models

For a delta model the shared target combines both components at the same
parameter values:
$h = \log(\mathrm{logit}^{-1}(\eta_1)) + \eta_2$, i.e., the log of encounter
probability times positive mean, not the sum of the two linear predictors.
Conventional logit/log `delta_gamma()` and `delta_lognormal()` are
supported. As in prediction, the surface's `offset` enters the positive
component only.

Poisson-link delta models (`delta_gamma(type = "poisson-link")` or
`delta_lognormal(type = "poisson-link")`) are also supported. Their
expected catch is $\exp(o + \eta_1 + \eta_2)$ for offset $o$, so here the
target is the sum of the linear predictors plus the surface's `offset`.
Unlike the target, `predict()` ignores offsets for these models, so the two
agree only at the default `offset = 0`.

## Interpretation and limitations

* **A model for cell visits.** The sampling model says whether a cell was
  visited at least once per time step. It is not a point-process model of
  fishing locations. Coefficients, including $b$, depend on the cell size
  and the time step. Use cells fine enough that biology is roughly constant
  within a cell, and check sensitivity to the resolution.
* **Repeated visits.** One indicator per cell-time is modeled. Multiple
  catches within a cell enter the catch likelihood, but the number of
  visits and where fishers fish within a cell aren't modeled.
* **What "standardized" means.** $h$ is expected catch at your reference
  catchability, conditional on the fields. It isn't absolute biomass, and
  the method can't tell biological covariates from catchability covariates.
  The sampling target and the index grid can use different
  standardizations; state both.
* **Simulation and residuals are conditional.** `simulate()` draws catches
  at the observed locations given the fitted fields; it doesn't draw new
  sampling locations. Residuals check the catch model only.
  `simulate(re_form = NA)`, `project()`, and `sdmTMB_cv()` aren't supported.
  `predict_sampling()` covers the fitted frame only.
* **Currently supported.** The RTMB backend; single log-link Poisson, NB2,
  Gamma, Tweedie, or lognormal families, or delta-gamma and
  delta-lognormal with default links (standard or Poisson-link); univariate `s()` smoothers; IID random intercepts
  (excluded, or included at known levels); ordinary isotropic spatial and
  spatiotemporal fields; ordinary fixed effects in the sampling formula; a
  constant preference coefficient or one that varies by time step
  (`coefficient = "iid"` or `"rw"`); and an optional IID or random-walk
  sampling baseline (`baseline`). Other features give an error.

## Troubleshooting

* **"Observed sampling indicators are all 0/1 wherever column ... is 1".**
  A free intercept for a time step with no sampled (or no unsampled)
  eligible cells has no finite estimate. Drop that time step from the frame,
  set its indicators to `NA`, or use a pooled baseline (`sampled ~ 1`).
* **"The sampling design is not full rank".** Often a time step whose
  indicators are all `NA` under `factor(year)`. Drop those rows from the
  frame. The biological time step is kept either way.
* **The sampling-field SD goes to zero,** with a non-positive-definite
  Hessian. The data don't support the field (see above). Turn it off, or
  report it as a boundary estimate.
* **`sigma_b_pref` (or `sigma_alpha_pref`) goes to zero.** With
  `coefficient = "iid"` or `"rw"`, the data may not show year-to-year
  variation in preference beyond sampling noise. With 8 or so time steps,
  this also happens by chance when the preference does vary. A collapsed SD
  gives the constant model's estimates; report it as a boundary estimate or
  use `coefficient = "constant"`.
* **`b_pref` is uncertain or strongly correlated with other parameters.**
  The preference coefficient is identified by variation in the shared fields
  that the sampling covariates don't explain. A sampling formula that
  duplicates the catch covariates, a flexible sampling field with a range
  like the catch fields, or weak spatial signal can all compete with $b$.
  Prefer measured sampling covariates, and check how sensitive the
  estimate is to the sampling specification.
* **An error about a covariate or new factor level in the sampling
  `data`.** The frame needs every main-model covariate, with no missing
  values, and factors with levels from the fitted data. For a covariate you
  want to vary in the sampling formula but standardize in the catch surface,
  use two columns (e.g., `gear_actual` and `gear`).
