---
title: "Sharing Matérn ranges with range_groups"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sharing Matérn ranges with range_groups}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**If the code in this vignette has not been evaluated, a rendered version is available on the [documentation site](https://sdmTMB.github.io/sdmTMB/index.html) under 'Articles'.**

```{r setup, include = FALSE, cache=FALSE}
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = EVAL,
  purl = EVAL
)
```

```{r packages, message=FALSE, warning=TRUE}
library(sdmTMB)
```

Each Gaussian random field in an sdmTMB model has a Matérn range: the distance at which two points are effectively independent.
By default, the spatial and spatiotemporal fields of a model component share one range (`share_range = TRUE`), and `share_range = FALSE` estimates one range for each.

The `range_groups` argument is a more flexible alternative.
You give each field a label, and fields with the same label share a range.
Labels work across the components of a delta model and, with the RTMB backend, for spatially varying coefficients.

We'll use the Pacific Cod survey data and a coarse mesh to keep things fast.

```{r mesh}
mesh <- make_mesh(pcod, c("X", "Y"), cutoff = 15)
```

## Spatial and spatiotemporal fields

Labels are given as a character vector named by field.
It doesn't matter what the labels you give are, as long as they are unique when they should be separate and the same when they should be shared/mirrored.
These two models are the same:

```{r separate}
fit_sep <- sdmTMB(
  density ~ 0 + as.factor(year),
  data = pcod, mesh = mesh, time = "year", family = tweedie(),
  range_groups = c(spatial = "s", spatiotemporal = "st")
)
fit_sep2 <- update(fit_sep, range_groups = NULL, share_range = FALSE)
logLik(fit_sep)
logLik(fit_sep2)
```

Giving both fields the same label (or leaving `range_groups` out) shares the range:

```{r shared}
fit_shared <- update(fit_sep, range_groups = c(spatial = "s", spatiotemporal = "s"))
tidy(fit_shared, "ran_pars")
```

## Delta models

In a delta model, labels are matched across the two components.
A single vector applies to both components, so this model estimates one spatial range and one spatiotemporal range, each shared by the encounter and positive components:

```{r delta-shared}
fit_dg <- sdmTMB(
  density ~ 0 + as.factor(year),
  data = pcod, mesh = mesh, time = "year", family = delta_gamma(),
  range_groups = c(spatial = "s", spatiotemporal = "st")
)
fit_dg
```

The printed output notes which ranges are shared across components.
A list gives one vector per linear predictor (sometimes called component here).
Here, the spatial range is shared across components but each component gets its own spatiotemporal range:

```{r delta-list}
fit_dg2 <- update(
  fit_dg,
  range_groups = list(
    c(spatial = "s", spatiotemporal = "st1"),
    c(spatial = "s", spatiotemporal = "st2")
  )
)
AIC(fit_dg, fit_dg2)
```

## Spatially varying coefficients

Spatially varying coefficient fields use their component's spatial range by default.
With the RTMB backend, you can name a coefficient (by its `spatial_varying` model matrix column name) to give it its own range or to share another field's range:

```{r svc}
mesh_2011 <- make_mesh(pcod_2011, c("X", "Y"), cutoff = 7)
fit_svc <- sdmTMB(
  density ~ depth_scaled,
  data = pcod_2011, mesh = mesh_2011, family = tweedie(),
  spatial_varying = ~ 0 + depth_scaled,
  range_groups = c(spatial = "s", depth_scaled = "d"),
  control = sdmTMBcontrol(backend = "rtmb")
)
tidy(fit_svc, "ran_pars")
```

The coefficient's range appears as `range_Z`, in the same order as the `sigma_Z` rows.

## Priors

PC Matérn priors (see `?sdmTMBpriors`) put a prior on each range once, however many fields share it.
Spatially varying coefficient fields can get their own PC prior with `sdmTMBpriors(matern_svc = pc_matern(...))`.
