---
title: "Spatially varying coefficients with factor predictors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatially varying coefficients with factor predictors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**If the code in this vignette has not been evaluated, a rendered version is available on the [documentation site](https://sdmTMB.github.io/sdmTMB/index.html) under 'Articles'.**

```{r setup, include = FALSE, cache=FALSE}
dplyr_installed <- require("dplyr", quietly = TRUE)
ggplot_installed <- require("ggplot2", quietly = TRUE)
pkgs <- dplyr_installed && ggplot_installed
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = EVAL,
  purl = EVAL
)
```

```{r packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(sdmTMB)
```

When a `spatial_varying` formula contains a factor variable—such as gear type, survey method, or year—the choice of there are several possible model structures.
Because each SVC (spatially varying coefficient) field is a penalized spatial random field with its own variance parameter (by default), the number and interpretation of those fields depend on how `spatial` and `spatial_varying` are specified together.

This vignette explains the main configurations, shows how to verify what was actually fit, and gives guidance on which structure to choose for a given scientific question.

## Quick reference

| Case | `spatial` | `spatial_varying` | Fields estimated | When to use |
|------|-----------|-------------------|-----------------|-------------|
| 1 | `"on"` | `~ 0 + factor` | `omega_s` + K SVC fields | shared average spatial baseline + factor-level deviations |
| 2 | `"on"` | `~ 1 + factor` | `omega_s` as reference + K−1 SVC deviation fields | one level is a natural reference/baseline |
| 3 | `"off"` | `~ 1 + factor` | K SVC fields | same as Case 2 but intercept in SVC component |
| 4 | `"off"` | `~ 0 + factor` | K independent SVC fields, no `omega_s` | levels treated symmetrically, no shared surface |

## Setup

We use the built-in British Columbia Pacific Cod dataset. Coordinates X and Y are in UTM zone 9 (km). We create a pre-factored year column so that the SVC column names are clean. We use the year column because it is easily available in the `pcod_2011` dataset, but the factor column might more commonly be a gear type, survey type, or maybe a month.

```{r data}
pcod_2011$fyear <- factor(pcod_2011$year)
mesh <- make_mesh(pcod_2011, c("X", "Y"), cutoff = 20)
```

For some structural demonstrations we use `do_fit = FALSE`, which builds the model object without fitting it. This lets us inspect `fit$spatial` and `colnames(fit$tmb_data$z_i)` (the SVC model matrix) cheaply.

---

## Case 1: Global spatial field plus factor-level SVC deviations

```{r caseB}
fit1 <- sdmTMB(
  density ~ 1,
  data = pcod_2011,
  mesh = mesh,
  spatial = "on",
  spatial_varying = ~ 0 + fyear,
  family = tweedie(link = "log"),
  do_fit = FALSE
)
fit1$spatial
colnames(fit1$tmb_data$z_i)
```

This estimates one global `omega_s` plus one SVC field per factor level (K fields total). The linear predictor for level k is:

$$\eta_k(s) = \ldots + \omega_s + \zeta_{k,s}$$

The interpretation is an average spatial pattern—estimated from all data—plus factor-level spatial deviations from that average. Because the global field and the level-specific deviations can trade off against each other, this model may be harder to estimate than the other cases shown here.

Use this when a scientifically meaningful shared spatial baseline exists and you want to allow each factor level to depart from it spatially.

---

## Case 2: Reference-level field plus SVC deviations

```{r caseC}
fit2 <- sdmTMB(
  density ~ 1,
  data = pcod_2011,
  mesh = mesh,
  spatial = "on",
  spatial_varying = ~ 1 + fyear,
  family = tweedie(link = "log"),
  do_fit = TRUE
)
fit2$spatial
colnames(fit2$tmb_data$z_i) # note that fyear2011 is missing
```

The `~ 1 + fyear` formula creates treatment contrasts: one intercept plus K−1 deviation columns. When `spatial = "on"` and `spatial_varying` includes an intercept, sdmTMB aliases the ordinary spatial field `omega_s` as the SVC intercept/reference-level field, drops the duplicate `(Intercept)` column from the SVC design matrix, and prints a one-line informational message (suppressible via `silent = TRUE`). The K−1 SVC fields are spatial deviations of the other levels from that reference.

Use this when one factor level is a natural reference—a trusted gear, a primary survey, a control method—and you want the other levels to be expressed as spatial differences from it.

---

## Case 3: SVC intercept field plus deviations, no ordinary spatial field

```{r caseD}
fit3 <- sdmTMB(
  density ~ 1,
  data = pcod_2011,
  mesh = mesh,
  spatial = "off",
  spatial_varying = ~ 1 + fyear,
  family = tweedie(link = "log"),
  do_fit = TRUE
)
fit3$spatial
colnames(fit3$tmb_data$z_i)
```

With `spatial = "off"` and an intercept in `spatial_varying`, the `(Intercept)` column is *retained* in the SVC design matrix and becomes a genuine SVC intercept field. There is no `omega_s`. The linear predictor is:

Conceptually the same as Case 2, but the intercept field is an SVC field rather than `omega_s`.

We can prove they are the same:

```{r}
logLik(fit2)
logLik(fit3)
```

---

## Case 4: Independent SVC fields per factor level, no ordinary spatial field

```{r caseE}
fit4 <- sdmTMB(
  density ~ 1,
  data = pcod_2011,
  mesh = mesh,
  spatial = "off",
  spatial_varying = ~ 0 + fyear,
  family = tweedie(link = "log"),
  do_fit = FALSE
)
fit4$spatial
colnames(fit4$tmb_data$z_i)
```

This estimates one independent SVC field per factor level and no global `omega_s`. The linear predictor for level k is:

$$\eta_k(s) = \ldots + \zeta_{k,s}$$

Use Case 4 when the factor levels should be treated indepdently with no shared spatial surface. This is the cleanest syntax for "one independent spatial field per factor level."

---

## Fitted example: independent year fields (Case 4)

We now fit Case 4 on the full dataset with independent spatial fields for each year. This is appropriate when years are treated as exchangeable levels with no shared spatial trend.

```{r fit-caseE, warning=FALSE, message=FALSE, cache=TRUE}
fit <- sdmTMB(
  density ~ 1,
  data = pcod_2011,
  mesh = mesh,
  spatial = "off",
  spatial_varying = ~ 0 + fyear,
  family = tweedie(link = "log")
)
```

```{r tidy}
tidy(fit, "ran_pars", conf.int = TRUE)
```

The `sigma_Z` rows give the marginal standard deviation of each year's spatial field. Larger values indicate more spatial variation in that year's density.

Let's predict onto a spatial grid for all years and plot the year-specific spatial fields.

```{r predict-grid}
nd <- replicate_df(qcs_grid, "year", unique(pcod_2011$year))
nd$fyear <- factor(nd$year)
pred <- predict(fit, newdata = nd)
```

```{r plot-fields, fig.asp=1.2}
# select the zeta_s columns for each year
zeta_cols <- grep("^zeta_s_fyear", names(pred), value = TRUE)

pred_long <- pred |>
  select(X, Y, year, all_of(zeta_cols)) |>
  tidyr::pivot_longer(
    cols = all_of(zeta_cols),
    names_to = "field_year",
    values_to = "zeta_s"
  ) |>
  mutate(field_year = sub("zeta_s_fyear", "", field_year)) |>
  filter(as.character(year) == field_year)  # each row's year matches its field

ggplot(pred_long, aes(X, Y, fill = zeta_s)) +
  geom_raster() +
  facet_wrap(~year) +
  scale_fill_gradient2() +
  coord_fixed() +
  labs(fill = "Spatial field\n(log space)", title = "Year-specific spatial fields (zeta_s)")
```

Each panel shows the estimated log-scale spatial deviation for that year relative to the overall mean (fixed intercept). Positive values indicate areas of higher-than-average density in that year; negative values indicate lower-than-average density.

---

## Choosing a parameterization

The choice should be driven by the scientific question:

- **Case 1** — use when an average spatial pattern is scientifically meaningful and you want to allow factor levels to depart from it spatially. Be aware that the shared field and factor deviations can trade off, sometimes making estimation difficult.

- **Case 2/3** — use when one factor level is a natural reference (a primary survey, a standard gear, a control treatment). The global spatial field serves as that reference level's field, and the other levels are estimated as spatial differences from it. This is often the most interpretable parameterization for comparative studies. Case 2 includes the intercept as `omega_s`, i.e. a separate spatial random field. Case 3 includes the intercept as a column in the SVC design matrix.

- **Case 4** — use when the factor levels should be treated symmetrically/independently and no shared spatial surface is scientifically motivated. One independent spatial field per level.

---

## Changes to the SVC factor and intercept behaviour compared to earlier releases

Behaviour for one of these specifications changed compared to the sdmTMB 1.0 release and before:

| Case | Specification | Fit changed? | Notes |
|---|---|---|---|
| 1 | `spatial = "on"`, `~ 0 + factor` | No | Unchanged. The old message recommending `spatial = "off"` has been removed; this is a valid specification. |
| 2 | `spatial = "on"`, `~ 1 + factor` | No | Unchanged fit. A one-line informational message now reports that `omega_s` is being aliased as the SVC intercept (suppressible via `silent = TRUE`). |
| 3 | `spatial = "off"`, `~ 1 + factor` | **Yes** | Previously the intercept column was stripped *and* `omega_s` was zeroed, leaving only K−1 deviation fields with no reference. Now fits a genuine SVC intercept field plus K−1 deviations. A one-cycle warning is emitted. Use Case 2 to recover the historically-intended reference-level model. |
| 4 | `spatial = "off"`, `~ 0 + factor` | No | Same K SVC fields as before; the fitted object's `spatial` slot now reports `"off"` instead of incorrectly reporting `"on"`. |

Only Case 3 has a different fitted-model structure. In all other cases the fitted model is the same; only labelling and messaging have changed.
