---
title: "Visualizing sdmTMB conditional effects using visreg"
author: "Julia Indivero, Sean Anderson, Lewis Barnett, Philina English, Eric Ward"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizing sdmTMB conditional effects using visreg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**If the code in this vignette has not been evaluated, a rendered version is available on the [documentation site](https://sdmTMB.github.io/sdmTMB/index.html) under 'Articles'.**

```{r setup, include = FALSE, cache=FALSE}
ggplot_installed <- require("ggplot2", quietly = TRUE)
visreg_installed <- require("visreg", quietly = TRUE)
pkgs <- ggplot_installed && visreg_installed
EVAL <- identical(Sys.getenv("NOT_CRAN"), "true") && pkgs
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.asp = 0.618,
  eval = EVAL,
  purl = EVAL
)
```

```{r echo=FALSE}
library(sdmTMB)
library(ggplot2)
theme_set(theme_light())
library(visreg)
set.seed(1)
```

We can use the [visreg](https://CRAN.R-project.org/package=visreg) package to visually inspect the conditional effects of explanatory variables in sdmTMB models.

We will use the built-in Pacific cod survey data for this example.
We will fit a model with scaled depth and a factor effect of year using a Tweedie distribution.

```{r}
pcod_2011$fyear <- as.factor(pcod_2011$year)
mesh <- make_mesh(pcod_2011, c("X", "Y"), cutoff = 20)
fit <- sdmTMB(
  density ~ 0 + s(depth_scaled) + fyear,
  data = pcod_2011,
  mesh = mesh,
  spatial = "off", # off for vignette building speed
  family = tweedie(link = "log")
)
```

We can then plot the effect of depth in link space. 
This shows the value of depth on the x-axis and the change in response on the y-axis, holding all other variables constant.

```{r}
visreg(fit, xvar = "depth_scaled")
```

Note that by default, randomized quantile residuals are included. 
These should be normally distributed in link space if the model is consistent with the data.
These can be turned off with `partial = FALSE`

visreg plots use ggplot2, so we can customize them with ggplot2 functions:

```{r}
g <- visreg(fit, xvar = "depth_scaled")
g + xlab("Depth scaled")
```

We can also grab the underlying data it creates and make our own plot:

```{r}
d <- visreg(fit, xvar = "depth_scaled", plot = FALSE)
head(d$fit)
g <- ggplot(d$fit, aes(x = depth_scaled, y = visreg_fit)) +
  geom_line() +
  geom_ribbon(aes(ymin = visreg_lwr, ymax = visreg_upr), alpha = 0.5)

# residuals are in d$res
head(d$res)

g + geom_point(aes(y = visreg_res), data = d$res, size = 1, alpha = 0.4)
```

We can look at the effect of depth in each year separately, either in separate plots or overlaid on the same plot.

```{r}
visreg(fit, xvar = "depth_scaled", by = "fyear")
visreg(fit, xvar = "depth_scaled", by = "fyear", overlay = TRUE)
```

And at the effect of depth on the response scale, rather than link (log) scale.

```{r}
visreg(fit, xvar = "depth_scaled", scale = "response")
```

In this case, visreg adds "rug" lines indicating the where the data were observed along the x-axis variable.

`visreg()` can also be used to visualize just the categorical effect of year.

```{r}
visreg(fit, xvar = "fyear")
```

We can also simultaneously compare both depth and year with a two-dimensional contour plot.

```{r}
visreg2d(fit, xvar = "fyear", yvar = "depth_scaled")
```

By default, `visreg2d()` produces a ggplot2 raster and contour plot.
To produce a static 3D perspective plot, save the calculations and pass them to `persp()`:

```{r}
v2d <- visreg2d(fit, xvar = "fyear", yvar = "depth_scaled", plot = FALSE)
persp(v2d)
```

## Delta models

`visreg()` can also be used to plot the output of delta models from `sdmTMB()` by using similar code for the previous plots, but using the sdmTMB wrapper function `visreg_delta()` and specifying `model = 1` for the encounter (0 vs. non-zero) model or `model = 2` for the positive component model (e.g., Gamma, lognormal). 
For example:

```{r}
fit_dg <- sdmTMB(
  density ~ s(depth_scaled, year, k = 8),
  data = pcod_2011,
  mesh = pcod_mesh_2011,
  spatial = "off", # for vignette speed
  family = delta_gamma()
)

visreg_delta(fit_dg, xvar = "depth_scaled", model = 1)
visreg_delta(fit_dg, xvar = "depth_scaled", model = 2)
```

Note that for plotting with `visreg_delta()`, categorical variables like year need to be designated as a factor in the data frame, as in the example above with `fyear`, rather than in the model formula (e.g., `+ as.factor(year)`).
